### 2011-06-09 Paper/Paper.tex
paper added

\documentclass[a4paper,10pt]{article}

% Fancy maths stuff
\usepackage{amsmath, amsthm, amssymb}

% Algorithm typesetting
\usepackage{algorithmic}

\begin{document}

% TITELPAGINA

\begin{titlepage}
\begin{center}

\vspace{2.5cm}

% [CHANGE] The title of your thesis. If your thesis has a subtitle, then this
% should appear right below the main title, in a smaller font.
\begin{Huge}
Expressive Performance of Polyphonic Piano Music using Hierarchical Structure
\end{Huge}

\vspace{1.5cm}

% [CHANGE] Your full name. In case of multiple names, you can include their
% initials as well, e.g. "Jan G.J. van der Wegge".
Bastiaan J. van der Weij\\
% [CHANGE] Your student ID, as this has been assigned to you by the UvA
% administration.
5922151

\vspace{1.5cm}

% [DO NOT CHANGE]
Bachelor thesis\\
% [CHANGE] Whether your Bachelor thesis is 6 ECTS (regular) or 9 ECTS (Honours
% programme).
Credits: 15 EC

\vspace{0.5cm}

% [DO NOT CHANGE] The name of the educational programme.
Bachelor Opleiding Kunstmatige Intelligentie

\vspace{0.25cm}

% [DO NOT CHANGE] The addess of the educational programme.
University of Amsterdam\\
Faculty of Science\\
Science Park 904\\
1098 XH Amsterdam

\vspace{4cm}

\emph{Supervisor}\\
% [CHANGE] The name of your supervisor. Include the titles of your supervisor,
% as well as the initials for *all* of his/her first names.
Prof. dr. ir. R.J.H. Scha

\vspace{0.25cm}

% [CHANGE] The address of the institute at which your supervisor is working.
% Be sure to include (1) institute (is appropriate), (2) faculty (if
% appropriate), (3) organisation name, (4) organisation address (2 lines).
Institute for Language and Logic\\
Faculty of Science\\
University of Amsterdam\\
Science Park 904\\
1098 XH  Amsterdam

\vspace{1.5cm}

% [CHANGE] The date at which you will finalize and submit your thesis.
June 24th, 2010

\end{center}

\end{titlepage}

\begin{abstract}
This paper introduces a system for expressively performing transcribed music (scores) using statistical learning from a dataset. The dataset that is used contains performances aligned to the corresponding score. Expression is considered to be the way in which the performance deviates from the score. Expressive actions are predicted from links found between musical structure and expression. Previous expressive actions are also incorporated in the predictions. Musical structure will be automatically extracted from the scores using a set of formal rules that assign a hierarchical grouping structure to the music. To system will be able to deal with polyphonic music by splitting the scores into melody and harmony.

{\bf Keywords:} performance rendering, musical structure, constituent structure, polyphonic piano music
\end{abstract}
\section{Introduction}

A music performance involves a performer and an audience. The performer tries his best to perform music to the audience in a way that pleases the audience. In western art music tradition it is conventional that the performer provides his rendition of a musical score: a detailed transcription of the music in musical notation. A performer should be skilled enough to accurately reproduce the music in the score, but what sets a good performer apart from other performers and drives people to pay money to see a musical performance is his ability deviate from the score and interpret it in a way that pleases his audience. In fact, it is generally agreed on that for most western art music, a perfect reproduction of the score sounds very unpleasant. The performer makes the performance sound `natural' and expresses his musical knowledge and perhaps feelings in the performance, the resulting performance is expressive.

It should be clear that performing a musical piece is not a trivial task. Theoretically it would be possible to construct robotic performers with perfect mastery of their instruments as well as perfect ability to reproduce scores. Yet there is still large gap between a perfect score reproduction by robots and a good sounding performance (the perfect reproduction sounds, indeed, robotic). Performance rendering systems attempt to bridge this gap.

Deviations from a score in a performance can be described as the end result of a number of processes, the most important of which is called expression. It is not immediately clear where expression comes for or even what it is. It is not the within the scope of this thesis to elaborate widely on the complex phenomenon of expression. Instead, it will be assumed that expression is strongly linked to musical structure and that a computational model using this structure can to some extend adequately model expression. The quality of 
the system should be an indication of the correctness of this assumption.

Early performance rendering systems used rule-based approaches to recognise small local structures that are used as indications to add certain kind of expression. A prominent ongoing rule based system, \textit{Director Musices} uses 30 rules with weighting parameters that take music features as input and produce expressive actions as output. The rules where created with the help of musical experts. \textit{Pop-E} is a more recent rule based system that uses aspects of Lerdahl and Jackendoff's Generative Theory of Tonal Music(GTTM) [Introduce this]. Gerhard Widmer is doing research using a large dataset of performances aligned to scores.

As datasets become more available, machine learning approaches are gaining popularity. However, very few approaches make extensive use of structure. Some machine learning systems like the \textit{Music Interpretation System} make use of some aspects of GTTM and Widmer uses musical \textit{closure} [Narmour, introduce this properly] as a feature in his succesful \textit{YQX} system. This means that only the level of structure defined by closure is used by this system. It seems that higher level structure that would indicate for example repetition of parts plays an important [This argument is important and needs more elaboration] role in expression.

This thesis introduces machine learning system that makes of local structure as well as higher level structure to generate expressive performances of polyphonic piano music. Section X will introduce Y, section P will elaborate on Q and finally section Z will discuss the results.

\paragraph*{Evaluation.} Performance context, adaptability, creativity. Subjective listening, correlation.

\subsection{Related Work}
Move part of the introduction here?
\section{Approach}

\subsection{Score Model}
\subsubsection{Musical structure.}
When listening to music, the listener assigns a certain hierarchical structure to the music perceived: Notes make up phrases, phrases make up themes, themes make up parts and parts make up the piece. This structure is usually accentuated in a performance. The most obvious example of this is slowing down at the end of a piece. On a lower level, transitions between themes may be marked by slowing down, changing volume or changing speed. Although structure can be accentuated in a performance, listeners, and trained listeners even more so, will also be able to recognise structure in an expressionless reproduction of the score, or when quietly singing the music in their heads. This suggests that it is possible to extract structure from a merely score. It also seems likely that knowing the structure will help creating an expressive performance.

\paragraph*{Detecting repetition}

Similarity function based on rhythic pattern and melodic pattern.

\paragraph*{Using repetition to improve structural analysis}

Consider the musical sequence $S = \{n_0, n_1, n_2, n_3\}$ is ambiguous and has two possible assignments of constituents: $S_0 = \{n_0\}, S_1 = \{n_1, n_2, n_3\}$ or $S'_0 = \{n_0, n_1\}, S'_1 = \{n_2, n_3\}$. The notion off repetition can help us here. Consider the case that $S'_1$ is a repetition of $S'_0$ but $S_1$ is not a repetition of $S_0$, this tells us that the split $S = \{S'_0, S'_1\}$ is more likely than the split $S = \{S_0, S_1\}$.

\paragraph*{Using repetition to create constituent classes}

Repetition can also be used to make the constituent assignment more interesting. What follows is a method that identifies which consituents are repetitions of which other constituents.

\subsection{Performance Model}
\subsubsection{Smoothness of expression}
To avoid sudden changes in expression that sound weird, you would want a model that describes what a sensible sequence of expressive actions is. It may be sensible to train this model on all available data since this model should not describe what expressive actions are appropriate for one specific pianist or composer, it should rather describe what sequences of expressive actions are acceptable in general. To achieve this, a continuous hidden markov model, trained on all performances in the dataset, will be used.

Let $D$ be a collection of compositions. Let a composition, $c$ be a collection of expressive actions, $a$ and let an expressive action be a tuple $(l,t)$ describing the loudness deviation, timing deviation (and/or local tempo?) of the note. The hidden markov model can be described as follows:
\begin{align*}
D &= \{c_0, c_1, \cdots, c_n\}\\
c_i &= \{a_0, a_1, \cdots, a_n\}\\
a_i &= (l,t)\\
P(c) &= P(a_1 | a_0) * P(a_2|a_1, a_0) * \cdots * P(a_n|a_{n-1}, a_{n-2}, \cdots, a_0))
\end{align*}

\subsubsection{Expressive timing} Honing and Desain, Honing and timmers, Honing.
\subsection{Learning}

\section{Method}
\section{Evaluation and Results}
\subsection{Subjective Listening}
\subsection{Correlation}
\section{Conclusion}
\section{Future Work}

 \bibliographystyle{plain} % plain, nature, jbact
 \bibliography{myref}

\end{document}

### 2011-06-13 Code/analysescore.py
non-functioning performance generator

@@ -1,17 +1,22 @@
 import music21 as m21
+import sys
 
 
 def analyse(score):
   recursive_score(score[1][0:4], 0)
   recursive_score(score[2][0:4], 0)
 
+def toText(score):
+  recursive_score(score, 0)
+
 def recursive_score(obj, level):
   string = ''
   for tab in range(0,level):
     string += '\t'
-  print string + str(obj)
+  print '{0}{1} [duration:{2}]'.format(string, str(obj), str(obj.duration)) 
   if hasattr(obj,'__iter__'):
     for x in obj:
       recursive_score(x, level + 1)
 
-
+if __name__ == "__main__":
+  toText(m21.converter.parse(sys.argv[1]))

### 2011-06-13 Code/database.py

@@ -1,3 +1,5 @@
+#!/usr/bin/env python
+
 import music21 as m21
 from representation import *
 from midifile import *
@@ -142,7 +144,12 @@ def getScore(composer, work, pianist, soundfont):
 
 def getDeviation(composer, work, pianist, soundfont):
   path = getDeviationPath(composer, work, pianist, soundfont)
-  
+
+def select():
+  comp = util.menu('Choose a composer', getComposers())
+  work = util.menu('Choose a work and performer', getWorks(getComposers()[comp]), cancel=True)
+  selected = getWorks(getComposers()[comp])[work]
+  return getComposers()[comp], selected[0], selected[1], selected[2]
 
 if __name__ == "__main__":
   import sequencer

### 2011-06-13 Code/deviationmodel.py

@@ -6,30 +6,234 @@
 # Hash an encoding string for note and measure store performance features in dictionary
 
 import music21 as m21
-import re
+import re, database,os
+from representation import *
 from BeautifulSoup import *
 
+class Deviations:
+
+  def __init__(self, tempo, tempo_deviations, note_deviations):
+    self.tempo = tempo
+    self.tempo_deviations = tempo_deviations
+    self.note_deviations = note_deviations
+
+
+
 def note_deviations(deviation):
+  print "Reading deviations file"
   f = open(deviation)
   devfile = f.read()
   soup = BeautifulStoneSoup(devfile)
-  tags = soup.findAll('note-deviation')
-# There are also misses!
-  print len(tags)
+  print "Reading target score"
+  f = open(os.path.dirname(deviation) + '/' + soup.deviation['target'])
+  targetfile = f.read()
+  target = BeautifulStoneSoup(targetfile)
+
+  
+  print "Constructing measures dictionary"
+  measures = {}
+  for measure in target.findAll('measure'):
+    measures[str(measure['number'])] = measure
+
+  print "Parsing deviations"
+  tags = soup.findAll('note-deviation') + soup.findAll('miss-note')
   note_deviations = {}
   pointerexp = re.compile('@id=\'(.*)\'\]/measure\[\@number=\'(.*)\'\]/note\[(.*)\]')
   for tag in tags:
     m = re.search(pointerexp, tag.attrs[0][1])
-    if not m: continue
-    key = "Part-id: {0} measure: {1} number: {2}".format(m.group(1), m.group(2), m.group(3))
-    value = (\
-        float(tag.attack.contents[0]),\
-        float(tag.release.contents[0]),\
-        float(tag.dynamics.contents[0]),\
-        float(tag.contents[3].contents[0]),\
-      )
-    note_deviations[key] = value
-    print("{0}: {1}".format(key, value))
-  return note_deviations
+    if not m: 
+      print "xpointer reference didn't match, bug in code or bug in deviation file"
+      continue
 
-note_deviations('/home/bastiaan/UvA/Expressive-Performance_DATA/CrestMusePEDB/PEDBv2.2/cho-etd003-ashke-b/deviation.xml')
+    # Find the voice and notenumber in that voice
+    measure = measures[m.group(2)]
+    notes = measure.findAll('note')
+    note = notes[int(m.group(3)) - 1]
+    voice = note.voice.contents[0]
+
[…]

### 2011-06-13 Code/representation.py

@@ -38,7 +38,7 @@ class NoteList:
     self.time_signature = None
     self.smtp_offset = None
     self.tempo = 500000
-    self.division = None
+    self.division = 480
     self.sequence_names = []
 
     # If a file is specified, parse it
@@ -154,7 +154,7 @@ class NoteList:
     out.write()
 
   def __str__(self):
-    string = ''
+    string = 'NoteList Object:\n'
     for n in self:
       string += str(n) + "\n" 
     return string
@@ -174,6 +174,9 @@ class NoteList:
       index += 1
 
     self.notes.append(note)
+
+  def microseconds_to_ticks(self, microseconds):
+    return int((microseconds / float(self.tempo)) * self.division)
   
   def ioi(self, note):
     if note == 0: return 0

### 2011-06-21 Paper/Paper.tex
non-functioning performance generator
Worked on repetition detection, fruitless

@@
 
-% Algorithm typesetting
+% Algorithm typesetting environments
+\usepackage{algorithm}
 \usepackage{algorithmic}
+
+% Trees
+\usepackage{qtree}
 
@@
 Move part of the introduction here?
+
 \section{Approach}
 
-\subsection{Score Model}
-\subsubsection{Musical structure.}
+\subsection{Musical structure}
 When listening to music, the listener assigns a certain hierarchical structure to the music perceived: Notes make up phrases, phrases make up themes, themes make up parts and parts make up the piece.
@@
 
-\paragraph*{Detecting repetition}
+\subsubsection*{The delta framework}
 
-Similarity function based on rhythic pattern and melodic pattern.
+The delta framework is based on the idea that small differences indicate low level structural transitions and large differences indicate higher level structural transitions.
+
+\paragraph*{Combining delta trees}
+
+In his MSc dissertation, van den Berg suggests using \textit{yeild rules} to combine trees generated by different delta functions.
+The yeild rule states that nodes in a set of trees can be connected if they are \textit{yeild equal}, that is, if the their set of leaf nodes they recursively contain is equal.
+The result is a multi-tree, in which the top note is always connected and intermediate nodes may or may not be connected.
+
+Unfortunately, a multi tree is hard to interpret since
 
@@
 
+Repetition be formulated in terms of the delta framework to provide greater generality.
+It seems desirable that we consider a transposed repetition of an earlier theme to be a repetition nevertheless.
+The same holds for rhythmic repetition: it may be that a particular rhythm is repeated, we want to treat this as a repetition too.
+The delta framework allows this intuition to be implemented easily: we only look at repetition in the deltas between notes.
+If we take for example the inter
+---------------------------
 Consider the musical sequence $S = \{n_0, n_1, n_2, n_3\}$ is ambiguous and has two possible assignments of constituents: $S_0 = \{n_0\}, S_1 = \{n_1, n_2, n_3\}$ or $S'_0 = \{n_0, n_1\}, S'_1 = \{n_2, n_3\}$.
@@
 Repetition can also be used to make the constituent assignment more interesting. What follows is a method that identifies which consituents are repetitions of which other constituents.
+
+\begin{algorithm}
+\begin{algorithmic}
+\FOR{$\texttt{depth} \in \{\texttt{depth}(\texttt{tree}), \cdots, 0\}$}
+  \FOR{$c \in \{c | \texttt{depth}(c, \texttt{tree}) = \texttt{depth}\}$}
+    \FOR{$\texttt{class} \in C$}
+      \IF{$\texttt{similar}(c, \texttt{class})$}
+        \STATE c.class = class
+      \ENDIF
+    \ENDFOR
+    \IF{c.class = null}
+      \STATE C.append(createClass(c))
+      \STATE c.class = createClass(c)
+    \ENDIF
+  \ENDFOR
+\ENDFOR
+\end{algorithmic}
+\caption{Bottom up tree traversal and similarity relations}
+\end{algorithm}
+
+A resulting tree could look like figure \ref{similtree}
+
+\begin{figure}
+\begin{center}
+\Tree 
+[
+.{$M$}
+	[ .{$d$} 
+		[ .{$a$} 
+			[ .{$x$} ]
+			[ .{$y$} ]
+			[ .{$x$} ]
+		]
+		[ .{$a$} 
+			[ .{$x$} ]
+			[ .{$y$} ]
+			[ .{$x$} ]
+		]
+	]	
+	[ .{$e$} 
+		[.{$c$} 
+			[ .{$b$} 
+				[ .{$p$} ]
+				[ .{$q$} ]
+			]
+			[ .{$r$} ]
+		]		
+		[.{$a$} 
+			[ .{$x$} ]
+			[ .{$y$} ]
+			[ .{$x$} ]
+		]
+	]
+]
+\end{center}
+\caption{Example tree with similarities}
+\label{similtree}
+\end{figure}
 

### 2011-06-21 Code/alignment.py

@@ -64,17 +64,13 @@ class Alignment:
   def expressiveMelody(self):
     return self.performance(self.melody())
 
-  def performance(self):
-    return self.performance(self.score)
-
-  def melodyPerformance(self, melody):
-    return self.performance(self.score)
-
   def melodyPerformance(self):
     return self.performance(self.score, self.melody())
 
-  def performance(self, score, melody=None):
+  def performance(self, score=None, melody=None):
     print "Creating performance"
+    if not score:
+      score = self.score
     # I am not sure about the units in which init silence is expressed
     begintime = self.deviations.init_silence * 1000000.0
     #performancetime = self.deviations.init_silence * self.deviations.getBeatDuration()
@@ -113,7 +109,9 @@ class Alignment:
               if melody:
                 melodynote = None
                 try:
-                  melodynote = melody[1][part.index(measure)].getElementById(1).getElementsByOffset(note.offset)[0]
+                  melodyvoice = melody[1][part.index(measure)].getElementById(1)
+                  if melodyvoice:
+                    melodynote = melodyvoice.getElementsByOffset(note.offset)[0]
                   if measure.number < 4:
                     print "{0} {1} {2}".format(note, melodynote, note.offset)
                 except IndexError:
@@ -132,7 +130,9 @@ class Alignment:
                         continue
                       barduration = melody[1][index].barDuration.quarterLength
                       try:
-                        melodynote = melody[1][index].getElementById(1).getElementsByOffset(0, barduration)[0]
+                        melodyvoice = melody[1][index].getElementById(1)
+                        if melodyvoice:
+                          melodynote = melodyvoice.getElementsByOffset(0, barduration)[0]
                       except IndexError:
                         pass
                     else: 
@@ -171,25 +171,6 @@ class Alignment:
 
   def getDeviations(self,note):
     return self.alignment[note.id, str(note.pitch)]
-
-  def note_ids(self):
-    count = 0
-    max = 20
-    for part in self.score:
-      if not isinstance(part, m21.stream.Part):
-        continue
-      for measure in part:
-        if not isinstance(measure, m21.stream.Measure):
-          continue
-        for voice in measure:
-          if not isinstance(voice, m21.stream.Voice):
-            continue
-          for note in voice:
-            if count > max:
-              break
-            if isinstance(note, m21.note.Note):
-              count += 1
-              print '{0} {1}'.format(str(note), note.id)
 
   def align(self, score, deviations):
     output = NoteList()

### 2011-06-21 Code/database.py

@@ -5,6 +5,7 @@ from representation import *
 from deviations import *
 from midifile import *
 from score import *
+from alignment import *
 import os, re, util, analysescore
 
 DB_PATH = "/home/bastiaan/UvA/Expressive-Performance_DATA/CrestMusePEDB/"
@@ -195,11 +196,12 @@ if __name__ == "__main__":
       else:
         path = getScoreMidiPath(getComposers()[comp], selected[0], selected[1], selected[2])
 
+      selection = (getComposers()[comp], selected[0], selected[1], selected[2])
       print path
 
       choice = util.menu('Choose action', \
           ['Play with internal sequencer', 'Play with audacious', 'View score', 'View midi info', 'View score info',\
-          'Export deviation data to CSV', 'Analyse', 'Extract melody','Cancel'])
+          'Export deviation data to CSV', 'Performance', 'Extract melody','Cancel'])
       if choice == 0:
         seq.play(NoteList(path))
       elif choice == 1:
@@ -230,7 +232,8 @@ if __name__ == "__main__":
         os.system("cmx dev2csv {0} > deviation.csv".format(getDeviationPath(getComposers()[comp], selected[0], selected[1], selected[2])))
         os.system("vim deviation.csv")
       elif choice == 6:
-        analysescore.analyse(getScore(getComposers()[comp], selected[0], selected[1], selected[2]))
+        alignment = Alignment(getScorePath1(selection), getDeviation1(selection))
+        seq.play(alignment.performance())
       elif choice == 7:
         score = Score(getScore(getComposers()[comp], selected[0], selected[1], selected[2]))
         melody = score.melody()

### 2011-06-21 Code/expression_rules.py

@@ -1,91 +1,0 @@
-#!/usr/env/python
-
-from music21 import *
-import sys, math
-
-
-
-def replaceEights(score):
-  for i in range(len(score)):
-    if isinstance(score[i], note.Note):
-      print score[i].fullName
-      if(score[i].quarterLength == 0.25):
-        score[i].quarterLength = 2
-    elif isinstance(score[i], stream.Score) or \
-        isinstance(score[i], stream.Part) or \
-        isinstance(score[i], stream.Measure):
-      replaceEights(score[i])
-
-def averagePitch(score):
-  pitches = []
-  for i in range(len(score)):
-    if isinstance(score[i], stream.Part):
-      for j in range(len(score[i])):
-        if isinstance(score[i][j], stream.Measure):
-          for k in range(len(score[i][j])):
-            if isinstance(score[i][j][k], note.Note):
-              n = score[i][j][k]
-              pitches.append(n.frequency)
-  return sum(pitches)/len(pitches)
-
-def higherIsFaster(score):
-  scale = 10000
-  print "Calculating average pitch"
-  avg = averagePitch(score)
-  print "Average pitch: {0}".format(avg)
-  print "Adding expression"
-  for i in range(len(score)):
-    if isinstance(score[i], stream.Part):
-      for j in range(len(score[i])):
-        if isinstance(score[i][j], stream.Measure):
-          for k in range(len(score[i][j])):
-            if isinstance(score[i][j][k], note.Note):
-              n = score[i][j][k]
-              height = n.frequency
-              diff = height - avg
-              speedup = math.pow(diff/scale, 3.0) + 1.0
-              print speedup
-              score[i][j][k].quarterLength *= speedup
-
-
-
-
-  
-if __name__ == '__main__':
-  print "Importing score"
-  score = converter.parse(sys.argv[1])
-  print "Doing stuff"
-  higherIsFaster(score)
-  print "Storing result"
-  ofile = open('output.xml', 'w')
-  ofile.write(score.musicxml)
-#  score.show()
-
-
-def toNoteStream(score):
-  s = stream.Stream()
-  for i in range(len(score)):
-    if isinstance(score[i], note.Note):
-      s.append(note)
-    elif isinstance(score[i], stream.Score) or \
-        isinstance(score[i], stream.Part) or \
-        isinstance(score[i], stream.Measure):
-      s += toNoteStream(score[i])
[…]

### 2011-06-21 Code/score.py

@@ -20,27 +20,7 @@ class Score:
   def getNoteList():
     return tools.parseScore(self.score)
 
-
-  def note_ids(self):
-    count = 0
-    max = 10
-    for part in self.score:
-      if not isinstance(part, m21.stream.Part):
-        continue
-      for measure in part:
-        if not isinstance(measure, m21.stream.Measure):
-          continue
-        for voice in measure:
-          if not isinstance(voice, m21.stream.Voice):
-            continue
-          for note in voice:
-            if count > max:
-              break
-            if isinstance(note, m21.note.Note):
-              print '{0} {1}'.format(str(note), note.id)
-              count += 1
-
-  #Highest note at any moment
+  #Highest note in part one at any moment
   def melody(self):
     notes = []
     melody = m21.stream.Score()

### 2011-06-21 Code/scoremodel.py

@@ -1,3 +1,25 @@
-import notelist
+import structure, tools
+from alignment import *
 
-class NoteFeatures(NoteList):
+
+# Take a score as input, return a list of features
+def scoremodel(alignment):
+  melody = alignment.melody()
+  groupings = structure.structure(melody)
+  melodynotes = tools.parseScore(melody)
+  features = {}
+  expressive_actions = []
+  for n in melodynotes:
+    features[n.annotation] = features(n)
+    expressive_actions.append(alignment.alignment[n.id, str(n.pitch)])
+    
+
+def features(note):
+  # dummy feature
+  return (n.pitch)
+
+    
+
+  
+
+

### 2011-06-21 Code/test.py

@@ -3,21 +3,107 @@ from alignment import *
 from representation import *
 from sequencer import *
 from score import *
-import tools, structure
+import tools, structure, sys
 
+if len(sys.argv) > 1:
+  deltas = structure.repetition(None, None, sys.argv[1])
+  print deltas
+  s3 = structure.delta_rule(None, 's'+sys.argv[1], deltas)
+  print tools.recursive_print(s3)
+  sys.exit(0)
 selection = db.select()
 score = Score(db.getScore1(selection))
 melody = score.melody()
 notes = tools.parseScore(melody)
-s1 = structure.delta_rule(structure.pitch, notes[:])
-print "PITCH DELTA TREE"
-print tools.recursive_print(s1) 
-s2 = structure.delta_rule(structure.onset, notes[:])
-print "IOI DELTA TREE"
-print tools.recursive_print(s2) 
-struct = structure.combine(s1, s2)
+deltas = []
+
 print "COMBINED DELTA TREES"
-print tools.recursive_print(struct) 
+
+def avg_list(lists, weights=None):
+  sumlist = []
+  if not weights:
+    weights = [1 for i in range(len(lists))]
+  for l in range(len(lists)):
+    index = 0
+    for item in lists[l]:
+      if index + 1 > len(sumlist):
+        sumlist.append(item)
+      else:
+        sumlist[index] += weights[l] * item
+      index += 1
+  result = []
+  for item in sumlist:
+    result.append(item / float(len(lists)))
+  return result
+
+def normalize(l):
+  m = float(max(l))
+  result = []
+  if(m == 0): 
+    m = 1
+  for i in l:
+    result.append(i/m)
+  return result
+
+def square(l):
+  result = []
+  for i in l:
+    result.append(i*i)
+  return result
+
+def preprocess(l):
+  return normalize(square(l))
+
+
+#for feature in [structure.pitch, structure.onset]:
+#  deltas.append(normalize(structure.repetition(feature, notes)))
+
+for feature in [structure.onset, structure.pitch]:
+  deltas.append(preprocess(structure.deltalist(feature, notes)))  
+
+avg = avg_list(deltas, [2, 1])
+s3 = structure.delta_rule(None, notes, avg)
+#avg = avg_list(repdeltas)
+#rep = structure.delta_rule(None, notes, avg)
+print tools.recursive_print(s3)
+#print "REPETITION DELTA:"
+#print tools.recursive_print(rep)
+
+for i in range(15):
[…]

### 2011-06-29 Code/alignment.py
Fiddling with tied notes, draft performance and score features implemented

@@ -5,38 +5,45 @@ from score import *
 import music21 as m21
 import pickle
 
-# This class adds a useful measure_time, indication the onset time in quarternotes from the beginning of the measure
-class AlignNote:
-
-  def __init__(self, ID, pitch, measure_time):
-    self.ID = ID
-    self.pitch = pitch
-    self.measure_time = measure_time
-
-
-
 class Alignment:
 
   def __init__(self, scorepath, deviations):
     print "Creating alignment"
     # Storing this object is not possible, score id's change and Scores are not pickleable :(
     self.deviations = deviations
-    #if deviations.target.split('.')[0] in os.listdir('scores'):
-    #  print "Stored score found, remove scores/{0} to create a new one".format(deviations.target.split('.')[0])
-    #  f = open('scores/{0}'.format(deviations.target.split('.')[0]), 'rb')
-    #  self.score = pickle.load(f)
-    #else:
-    self.score = m21.converter.parse(scorepath)
-    #  self.storeScore()
-    #if deviations.target.split('.')[0] in os.listdir('alignments'):
-    #  print "Stored alignment found, remove alignments/{0} to create a new one".format(deviations.target.split('.')[0])
-    #  f = open('alignments/{0}'.format(deviations.target.split('.')[0]), 'rb')
-    #  self.alignment = pickle.load(f)
-    #else:
-    self.alignment = self.align(self.score, deviations)
+    # Am I really writing ugly code like this? Apparently I am
+    if isinstance(scorepath,str):
+      self.score = m21.converter.parse(scorepath)
+    else:
+      self.score = scorepath
+    # This doesn't actually remove ties, but it does fix the duration of tied notes
+    self.score = self.score.stripTies()
+    # Remove all tied notes except the start notes
+    self.removeTies()
+    self.alignment = None
+    self.align()
+
     #  self.storeAlignment()
-    self.score.flat.stripTies() 
     print "Done"
+
+  def removeTies(self):
+    for part in self.score:
+      if not isinstance(part, m21.stream.Part):
+        continue
+      for measure in part:
+        if not isinstance(measure, m21.stream.Measure):
+          continue
+        for voice in measure:
+          if not isinstance(voice, m21.stream.Voice):
+            continue
+          for note in voice:
+            if isinstance(note, m21.chord.Chord) or isinstance(note, m21.note.Note):
+              if hasattr(note,'tie') and note.tie is not None:
+                if not note.tie.type == 'start':
[…]

### 2011-06-29 Code/database.py

@@ -8,7 +8,7 @@ from score import *
 from alignment import *
 import os, re, util, analysescore
 
-DB_PATH = "/home/bastiaan/UvA/Expressive-Performance_DATA/CrestMusePEDB/"
+DB_PATH = "../../Expressive-Performance_DATA/CrestMusePEDB/"
 VERSIONS = ['PEDBv2.2', 'PEDBv2.3', 'PEDBv2.4.1', 'PEDBv2.5', 'PEDBv3.0']
 
 composers = {'Bach':'bac', 'Bartok':'bar', 'Grieg':'gri', 'Mozart':'moz', \

### 2011-06-29 Code/deviationmodel.py

@@ -1,277 +1,0 @@
-# Take a score and deviation file as input.
-# Parse the score in music21
-# Iterate through every note of the score
-# Find corresponding note in deviationdata
-# Create performance features vector
-# Hash an encoding string for note and measure store performance features in dictionary
-
-import music21 as m21
-import re, os
-import database as db
-from representation import *
-from BeautifulSoup import *
-
-class Deviations:
-
-  def __init__(self, tempo, tempo_deviations, note_deviations):
-    self.tempo = tempo
-    self.tempo_deviations = tempo_deviations
-    self.note_deviations = note_deviations
-
-
-
-def note_deviations(deviation):
-  print "Reading deviations file"
-  f = open(deviation)
-  devfile = f.read()
-  soup = BeautifulStoneSoup(devfile)
-  print "Reading target score"
-  f = open(os.path.dirname(deviation) + '/' + soup.deviation['target'])
-  targetfile = f.read()
-  target = BeautifulStoneSoup(targetfile)
-
-  
-  print "Constructing measures dictionary"
-  measures = {}
-  for measure in target.findAll('measure'):
-    measures[str(measure['number'])] = measure
-
-  print "Parsing deviations"
-  tags = soup.findAll('note-deviation') + soup.findAll('miss-note')
-  note_deviations = {}
-  pointerexp = re.compile('@id=\'(.*)\'\]/measure\[\@number=\'(.*)\'\]/note\[(.*)\]')
-  for tag in tags:
-    m = re.search(pointerexp, tag.attrs[0][1])
-    if not m: 
-      print "xpointer reference didn't match, bug in code or bug in deviation file"
-      continue
-
-    # Find the voice and notenumber in that voice
-    measure = measures[m.group(2)]
-    notes = measure.findAll('note')
-    note = notes[int(m.group(3)) - 1]
-    voice = note.voice.contents[0]
-
-    if note.grace: continue
-    number = 0
-    for n in notes[0:int(m.group(3))-1]:
-      if n.rest: continue
-      if n.grace: continue
-      if n.voice.contents[0] == voice:
-        number += 1
-
-    alteration = ''
-    if note.alter:
-      if note.alter.contents[0] == '1':
-        alteration = '#'
-      elif note.alter.contents[0] == '-1':
-        alteration = '-'
-#    print '========='
-#    print m.group(3)
-#    for n in notes:
-#      print n
-#      print '++++++'
-#    print '---------'
-#    print tag
[…]

### 2011-06-29 Code/deviations.py

@@ -5,7 +5,12 @@ import os, pickle
 
 class Deviations:
 
-  def __init__(self, deviation, target):
+  def __init__(self, deviation=None, target=None):
+    if not deviation or not target:
+      self.bpm = 120
+      self.target = target
+      self.init_silence = 0
+      return
     targetname = os.path.split(target)[1].split('.')[0]
     if targetname in os.listdir('deviations'):
       print "Stored deviations found, remove deviations/{0} to create new ones".format(targetname)
@@ -33,16 +38,21 @@ class Deviations:
     f = open(location, 'wb')
     pickle.dump(self.data, f)
 
+  # Parse a deviation file into a measurelength, tempo_deviations and note_deviations dict
+  # It also reads the base tempo and should read the base dynamics
   def parse(self, deviation, targetpath):
     print "Reading deviations file"
     f = open(deviation)
     devfile = f.read()
+    # We need to look up some information in the score file as this information is lost
+    # in translation when the score is parsed with music21
     soup = BeautifulStoneSoup(devfile)
     print "Reading target score"
     f = open(targetpath)
     targetfile = f.read()
     target = BeautifulStoneSoup(targetfile)
 
+    # For speed: put all measure tags in a dictionary
     print "Constructing measures dictionary"
     measures = {}
     for measure in target.findAll('measure'):
@@ -67,9 +77,12 @@ class Deviations:
       # Instead of continuing, add gracenotes to stream here?
       if note.grace: continue
       number = 0
+      # Count the number of notes in the measure and skip rests, grace notes and tied notes
       for n in notes[0:int(m.group(3))-1]:
         if n.rest: continue
         if n.grace: continue
+        if n.tie and n.tie['type'] != 'start':
+          continue
         if n.voice.contents[0] == voice:
           number += 1
 
@@ -82,6 +95,8 @@ class Deviations:
       pitch = '{0}{1}{2}'.format(note.pitch.step.contents[0], alteration, note.pitch.octave.contents[0])
 
       key = "{0},{1},{2},{3},{4}".format(m.group(1), m.group(2), voice, pitch, number)
+      if m.group(2) == '18':
+        print key
       if tag.name == 'miss-note': 
         value = 'miss'
       else:
@@ -127,7 +142,10 @@ class Deviations:
   def getExpressiveTime(self, measure, beat):
     beat_lengths = []
     for b in range(self.measure_lengths[measure]):
[…]

### 2011-06-29 Code/performancefeatures.py

@@ -1,1 +1,108 @@
+import tools, structure, math
+
+def expression(alignment):
+  performance = alignment.expressiveMelody()
+  melodyscore = alignment.melody()
+  score = tools.parseScore(melodyscore)
+  # Tied notes end up as zero length notes here, clean this mess up
+  print 'SCORE LENGTH: {0}'.format(len(score))
+  remove = []
+  for i in range(len(score)):
+    if structure.duration(score, i) == 0:
+      remove.append(score[i])
+  for n in remove:
+    score.remove(n)
+  print 'SCORE LENGTH: {0}'.format(len(score))
+  # The two above should be guaranteed to be of equal length.
+  # However, better be safe than sorry
+  if len(performance) != len(score):
+    print 'This shouldn\'t happen: melodyscore and performance lengths don\'t match: {0} and {1}'.format(len(performance), len(score))
+  
+  # Mean loudness
+  mean_l = 0.0
+  for note in performance:
+    mean_l += note.onvelocity
+
+  mean_l = mean_l / float(len(performance))
+  print "Average loudness: {0}".format(mean_l)
+
+  expression = []
+  lasttempo = float(alignment.deviations.bpm)
+  lastdynamics = mean_l
+ 
+  # Assume same note order in score and performance (am I being naive?)
+  for i in range(len(performance)):
+    ioi_ratio = math.log(structure.ioi(performance, i) / float(structure.ioi(score, i)))
+    loudness_ratio = math.log(performance[i].onvelocity / mean_l)
+    
+    # This obviously results in zero divisions
+    #articulation = math.log(structure.silence(performance, i) / float(structure.silence(score, i))) 
+    articulation = 0
+
+    duration_ratio = math.log(structure.duration(performance, i) / float(structure.duration(score, i)))
+
+    # To be implemented: second order ioi loudness and articulation
+    ioi_change = 0
+    loudness_change = 0
+    articulation_change = 0
+    
+    # E = (ioi_r, loudness_r, articulation, duration_r, ioi_ch, loundess_ch)
+    e = (ioi_ratio, loudness_ratio, articulation, duration_ratio, ioi_change, loudness_change)
+    expression.append(e)
+
+
+  return expression
+
+
+def expressionOLD(alignment):
+
+  melody = tools.parseScore(melodyscore)
+  lastnote = None
+  lastscorenote = None
+
[…]

### 2011-06-29 Code/representation.py

@@ -194,6 +194,9 @@ class NoteList:
 
     self.notes.append(note)
 
+  def remove(self, item):
+    self.notes.remove(item)
+
   def quarternotes_to_ticks(self, quarternotes):
     return self.microseconds_to_ticks(quarternotes * self.tempo)
 
@@ -206,6 +209,9 @@ class NoteList:
   def seconds_to_ticks(self, microseconds):
     return int((0.000001 * microseconds / float(self.tempo)) * self.division)
   
+  def ticks_to_quarternotes(self, ticks):
+    return self.ticks_to_microseconds(ticks) / self.tempo
+
   def ticks_to_microseconds(self, ticks):
     return (ticks / float(self.division)) * float(self.tempo)
 

### 2011-06-29 Code/scorefeatures.py

@@ -1,3 +1,108 @@
+import tools, structure, math
+
+
 # Take a score, split it in melody and harmony, do structure analysis and output features per note.
 
+def discretize(min, max, value, bins=10):
+  if value < min or value > max:
+    print "Discretize error: value outside range, returning 0"
+    return 0
+  return int((value - min) * (bins / float(max - min)))
 
+def pitch_interval(notes, i):
+  if i == 0: return 0
+  return notes[i].pitch - notes[i-1].pitch
+
+def duration_ratio(notes, i):
+  if i == 0: return 1
+  return math.log((notes[i].off - notes[i].on) / float(notes[i-1].off - notes[i-1].on))
+
+def vanderWeijFeatures(melodyscore):
+  melody = tools.parseScore(melodyscore)
+  onset_tree = structure.second_order_tree(structure.onset, melody, 0.1)
+  pitch_tree = structure.second_order_tree(structure.pitch, melody, 0.1)
+  features = [[0] for i in range(len(melody))]
+  return features
+    
+
+# [onset arch, pitch arch, pitch_interval, duration_ratio]
+def widmerFeatures(melodyscore):
+  melody = tools.parseScore(melodyscore)
+  onset_segments = structure.bestgrouping(melody, structure.second_order_tree(structure.onset, melody, 0.1))
+  pitch_segments = structure.bestgrouping(melody, structure.second_order_tree(structure.pitch, melody, 0.1))
+  features = [[] for i in range(len(melody))]
+
+  index = 0
+  for i in range(len(onset_segments)):
+    start = onset_segments[i][0].on
+    end = onset_segments[i][len(onset_segments[i])-1].off
+    length = float(end - start)
+    if not i+1 >= len(onset_segments):
+      end = onset_segments[i+1][0].on
+    
+    for note in onset_segments[i]:
+      nearest_border = min(end - note.on, note.on - start)
+      features[index] = [nearest_border]
+      index += 1
+
+  index = 0
+  for i in range(len(pitch_segments)):
+    start = pitch_segments[i][0].on
+    end = pitch_segments[i][len(pitch_segments[i])-1].off
+    length = float(end - start)
+    if not i+1 >= len(pitch_segments):
+      end = pitch_segments[i+1][0].on
+    
+    for note in pitch_segments[i]:
+      nearest_border = min(end - note.on, note.on - start)
+      features[index].append(nearest_border)
+      index += 1
+
+  for i in range(len(melody)):
[…]

### 2011-06-29 Code/scoremodel.py

@@ -1,24 +1,52 @@
-import structure, tools
+import structure, tools, matplotlib
 from alignment import *
 
 
+states = {}
+transitions = {}
+emissions = {}
+
+# Features: ()
+
 # Take a score as input, return a list of features
-def scoremodel(alignment):
-  melody = alignment.melody()
-  groupings = structure.structure(melody)
-  melodynotes = tools.parseScore(melody)
-  features = {}
-  expressive_actions = []
-  for n in melodynotes:
-    features[n.annotation] = features(n)
-    expressive_actions.append(alignment.alignment[n.id, str(n.pitch)])
+def scoremodel(features, expression):
+  pitch_positions = [0 for i in range(11)]
+  onset_positions = [0 for i in range(11)]
+  onset_pos_count = [0 for i in range(11)]
+  pitch_pos_count = [0 for i in range(11)]
+  for work in features.keys():
+    f = features[work]
+    e = expression[work]
+    for notefeatures, noteexpression in zip(f,e):
+      onset_pos_count[int(notefeatures[0])] += 1
+      pitch_pos_count[int(notefeatures[1])] += 1
+      onset_positions[int(notefeatures[0])] += noteexpression[3]
+      pitch_positions[int(notefeatures[1])] += noteexpression[3]
+  for i in range(11):
+    if onset_pos_count[i] == 0:
+      print i
+      continue
+    if pitch_pos_count[i] == 0:
+      print i
+      continue
+    onset_positions[i] /= onset_pos_count[i]
+    pitch_positions[i] /= pitch_pos_count[i]
+
+  for i in range(11):
+    print'Onset: {0}: {1} (Based on {2} examples)'.format(i, onset_positions[i], onset_pos_count[i])
+    print'Pitch: {0}: {1} (Based on {2} examples)'.format(i, pitch_positions[i], pitch_pos_count[i])
+
+  return (onset_positions, pitch_positions)
+#  melody = alignment.melody()
+#  groupings = structure.structure(melody)
+#  melodynotes = tools.parseScore(melody)
+#  features = {}
+#  expressive_actions = []
+#  for n in melodynotes:
+#    features[n.annotation] = features(n)
+#    expressive_actions.append(alignment.alignment[n.id, str(n.pitch)])
     
 
-def features(note):
-  # dummy feature
-  return (n.pitch)
-
-    
 
   
 

### 2011-06-29 Code/test.py

@@ -1,147 +1,6 @@
 import database as db
 from alignment import *
-from representation import *
-from sequencer import *
-from score import *
-import tools, structure, sys
 
-def avg_list(lists, weights=None):
-  sumlist = []
-  if not weights:
-    weights = [1 for i in range(len(lists))]
-  for l in range(len(lists)):
-    index = 0
-    for item in lists[l]:
-      if index + 1 > len(sumlist):
-        sumlist.append(item)
-      else:
-        sumlist[index] += weights[l] * item
-      index += 1
-  result = []
-  for item in sumlist:
-    result.append(item / float(len(lists)))
-  return result
-
-def normalize(l):
-  m = float(max(l))
-  result = []
-  if(m == 0): 
-    m = 1
-  for i in l:
-    result.append(i/m)
-  return result
-
-def square(l):
-  result = []
-  for i in l:
-    result.append(i*i)
-  return result
-
-if len(sys.argv) > 1:
-  notes = sys.argv[1]
-  deltas = [int(x) for x in sys.argv[2]]
-  s3 = structure.second_order_deltarule(notes, deltas, 0)
-  print tools.recursive_print(s3)
-  sys.exit(0)
-
-
-selection = db.select()
-score = Score(db.getScore1(selection).stripTies())
-melody = score.melody()
-notes = tools.parseScore(melody)
-deltas = []
-
-
-print "COMBINED DELTA TREES"
-#for feature in [structure.pitch, structure.onset]:
-#  deltas.append(normalize(structure.repetition(feature, notes)))
-
-for feature in [structure.onset, structure.duration, structure.pitch]:
-  print ">>>> {0} absolute:".format(feature)
-  print structure.absolute_deltalist(feature, notes)
-  print ">>>> {0} relative:".format(feature)
-  print structure.relative_deltalist(feature, notes)
-
-
-#deltas.append(normalize(square(structure.absolute_deltalist(structure.onset, notes))))
-#deltas.append(normalize(structure.relative_deltalist(structure.pitch, notes)))
-
-deltas = []
-deltas.append(structure.absolute_deltalist(structure.onset, notes))
-deltas.append(structure.absolute_deltalist(structure.pitch, notes))
-
-sodeltas = []
-sodeltas.append(normalize(square(structure.second_order_deltalist(deltas[0]))))
[…]

### 2011-07-02 Paper/Paper.tex
Fiddling with tied notes, draft performance and score features implemented
Started on HMM featureextraction on mozart may not work

@@
 \begin{abstract}
+In this paper a system will be presented that will learn 
 This paper introduces a system for expressively performing transcribed music (scores) using statistical learning from a dataset.
@@
 \section{Conclusion}
+\cite{markwin}
 \section{Future Work}

### 2011-07-02 Code/alignment.py

@@ -7,7 +7,7 @@ import pickle, copy
 
 class Alignment:
 
-  def __init__(self, scorepath, deviations):
+  def __init__(self, scorepath, deviations, noAlign=False):
     print "Creating alignment"
     # Storing this object is not possible, score id's change and Scores are not pickleable :(
     self.deviations = deviations
@@ -20,14 +20,15 @@ class Alignment:
     m = s.melody()
     self.m = self.removeTies(m)
     self.score = self.removeTies(self.score)
-    self.alignment = None
-    self.align()
+    if not noAlign:
+      self.alignment = None
+      self.align()
 
     print "Done"
 
   def removeTies(self, score):
     # This doesn't actually remove ties, but it does fix the duration of tied notes
-    score = score.stripTies()
+    score.stripTies(inPlace=True)
     # Now remove every note that is not the start of a tie
     for part in score:
       if not isinstance(part, m21.stream.Part):
@@ -93,7 +94,7 @@ class Alignment:
         if not isinstance(measure,m21.stream.Measure): continue
         #if melody: print "Bar {0} done".format(measure.number)
         performancetime += measureduration
-        measureduration = self.deviations.getExpressiveTime(measure.number, self.deviations.measure_lengths[measure.number])
+        measureduration = self.deviations.getExpressiveTime(measure.number, measure.barDuration.quarterLength)
         for voice in measure.voices:
           for note in voice.notes:
             queue = {}
@@ -117,12 +118,13 @@ class Alignment:
               # If this isn't the case, use the tempo curve for expression and some measure of dynamics(nearest melody note?)
               if melody:
                 melodynote = None
+                # 
                 try:
                   melodyvoice = melody[1][part.index(measure)].getElementById(1)
                   if melodyvoice:
                     melodynote = melodyvoice.getElementsByOffset(note.offset)[0]
-                  if measure.number < 4:
-                    print "{0} {1} {2}".format(note, melodynote, note.offset)
+                    if not isinstance(melodynote, m21.note.Note):
+                      melodynote = None
                 except IndexError:
                   pass
                 if not melodynote or not isinstance(melodynote, m21.note.Note):
@@ -156,6 +158,8 @@ class Alignment:
                   # Attack and release should be zero in this case
                   deviation = (0, 0, deviation[2], deviation[3])
                 else:
+                  if not (melodynote.id, str(melodynote.pitch)) in self.alignment:
[…]

### 2011-07-02 Code/bunchofmelodies.py

@@ -1,0 +1,34 @@
+import tools
+import database as db
+import random, sys
+from score import *
+
+selections = []
+
+amount = 2
+
+store_xml = False
+
+if len(sys.argv) > 1:
+  if sys.argv[1] == 'xml':
+    store_xml = True
+
+print "Making selection"
+
+for composer in db.composers:
+  works = db.getWorks(composer)
+  for i in range(amount):
+    work = random.choice(works)
+    works.remove(work)
+    if len(works ) == 0: break
+    selections.append((composer, work[0], work[1], work[2]))
+
+print "Storing melodies"
+
+for selection in selections:
+  score = Score(db.getScore1(selection))
+  if store_xml:
+    tools.saveXML(score.melody(), '_'.join(score.score.metadata.title.split(" ")))
+  else:
+    tools.savePDF(score.melody(), '_'.join(score.score.metadata.title.split(" ")))
+

### 2011-07-02 Code/database.py

@@ -6,7 +6,7 @@ from deviations import *
 from midifile import *
 from score import *
 from alignment import *
-import os, re, util, analysescore
+import os, re, util, analysescore, pickle
 
 DB_PATH = "../../Expressive-Performance_DATA/CrestMusePEDB/"
 VERSIONS = ['PEDBv2.2', 'PEDBv2.3', 'PEDBv2.4.1', 'PEDBv2.5', 'PEDBv3.0']
@@ -151,12 +151,26 @@ def getPerformance(composer, work, pianist, soundfont):
   return NoteList(path)
 
 def getScore(composer, work, pianist, soundfont):
+  #scorepath = 'scoredata/{0}-{1}-{2}-{3}'.format(composer,work,pianist,soundfont)
   path = getScorePath(composer, work, pianist, soundfont)
   os.system('cp {0} ./lastused/score.xml'.format(path))
   # For convenience, also copy the performance
   path2 = getPerformancePath(composer, work, pianist, soundfont)
   os.system('cp {0} ./lastused/performance.mid'.format(path2))
-  return m21.converter.parse(path)
+  #if os.path.exists(scorepath):
+  #  print "Stored score object found, unpickling, remove {0} to load the score from xml".format(scorepath)
+  #  scorefile = open(scorepath, 'rb')
+  #  score = m21.stream.Score()
+  #  freezer = m21.converter.StreamFreezer(score)
+  #  freezer.openPickle(scorepath)
+  #else:
+  score = m21.converter.parse(path)
+  #  print "Pickling score object"
+  #  scorefile = open(scorepath, 'wb')
+  #  score.setupPickleScaffold()
+  #  freezer = m21.converter.StreamFreezer(score)
+  #  freezer.writePickle(scorepath)
+  return score
 
 def getDeviation(composer, work, pianist, soundfont):
   path = getDeviationPath(composer, work, pianist, soundfont)

### 2011-07-02 Code/deviations.py

@@ -10,16 +10,20 @@ class Deviations:
       self.bpm = 120
       self.target = target
       self.init_silence = 0
+      self.tempo_deviations = {}
+      self.note_deviations = {}
+      self.measure_lengths = {}
       return
-    targetname = os.path.split(target)[1].split('.')[0]
-    if targetname in os.listdir('deviations'):
+    dir = target.split('/')[-2]
+    name = target.split('/')[-1].split('.')[0]
+    targetname = '{0}/{1}'.format(dir, name)
+    if os.path.exists('deviations/{0}'.format(targetname)):
       print "Stored deviations found, remove deviations/{0} to create new ones".format(targetname)
       f = open('deviations/{0}'.format(targetname), 'rb')
       self.data = pickle.load(f)
-      print "Loaded"
     else:
       self.data = self.parse(deviation, target)
-      self.store(targetname)
+      self.store(dir, name)
     self.bpm = self.data['tempo']
     self.tempo_deviations = self.data['tempo_deviations']
     self.note_deviations = self.data['note_deviations']
@@ -32,8 +36,10 @@ class Deviations:
     # When should be extracted from soup, default=100
     self.basedynamics = self.data['basedynamics']
 
-  def store(self, name):
-    location = 'deviations/{0}'.format(name)
+  def store(self, dir, name):
+    if not os.path.exists('deviations/{0}'.format(dir)):
+      os.mkdir('deviations/{0}'.format(dir))
+    location = 'deviations/{0}/{1}'.format(dir, name)
     print "Storing deviations in: {0} for future reference".format(location)
     f = open(location, 'wb')
     pickle.dump(self.data, f)
@@ -122,7 +128,9 @@ class Deviations:
       for beat in beats:
         if beat.find('tempo-deviation'):
           measurelength += 1
-          tempo_deviations[(int(float(tag['number'])), int(float(beat['beat']))-1)] = float(beat.find('tempo-deviation').contents[0])
+          key = (int(float(tag['number'])), int(float(beat['beat']))-1)
+          value = float(beat.find('tempo-deviation').contents[0])
+          tempo_deviations[key] = value
       measure_lengths[int(tag['number'])] = measurelength
 
     return {'tempo':tempo, 'tempo_deviations':tempo_deviations,\
@@ -161,13 +169,18 @@ class Deviations:
     return sum([beat_lengths[i] for i in range(0, int(beat))]) + \
         multiplier * beat_lengths[int(beat)]
 
[…]

### 2011-07-02 Code/expressionmodel.py

@@ -1,0 +1,10 @@
+import structure
+import database as db
+from representation import *
+
+def built(alignment):
+  parsetree = structure.parse(alignment.score)
+  constituents = structure.groupings(parsetree)
+  for constituent in constituents:
+    for note in constituent:
+

### 2011-07-02 Code/hmm.py

@@ -1,0 +1,53 @@
+# P(Expression|Features)
+# P(Ei|Ei-i, ..., Ei-n)
+# P(E0, E1, ..., En)
+
+
+class HMM:
+
+  def __init__(self, order=2):
+    self.order = order
+
+    self.coincedences = {}
+    self.observations = {}
+    self.states = []
+
+    self.ngrams = {}
+    self.coincedence_count = 0
+    self.ngram_count = 0
+    self.lessergrams = {}
+    self.lessergram_count = 0
+    self.starts = {}
+    self.start_count = 0
+
+  def learn(self, observations, states):
+    self.starts[states[0]] = self.starts.get(states[0], 0) + 1
+    self.start_count += 1
+    for observation, state in zip(observations, states):
+      self.coincedences[observation, state] = self.coincedences.get((observation, state), 0) + 1
+      self.coincedence_count += 1
+      self.observations[observation] = self.observations.get(observation, 0) + 1
+      self.states.append(state)
+    for i in range(len(states)-self.order):
+      ngram = []
+      lessergram = []
+      for j in range(self.order):
+        ngram.append(states[i+j])
+      for j in range(self.order-1):
+        lessergram.append(states[i+j])
+      self.ngram_count += 1
+      self.ngrams[tuple(ngram)] = self.ngrams.get(tuple(ngram), 0) + 1
+      self.lessergram_count += 1
+      self.lessergrams[tuple(lessergram)] = self.lessergrams.get(tuple(lessergram), 0) + 1
+
+  def joint_probability(self, observation, state):
+    return self.coincedences[observation, state] / float(self.coincedence_count)
+
+  def start_probability(self, state):
+    return self.starts[state] / float(self.start_count)
+
+  def emission_probability(self, state, observation):
+    return self.coincedences[observation, state] / float(self.observations[observation])
+
+  def transition_probability(self, lessergram, state):
+    return self.ngrams[tuple(lessergram + [state])] /  float(lessergrams[tuple(lessergram)])

### 2011-07-02 Code/melodyperformance.py

@@ -1,0 +1,15 @@
+import database as db
+from alignment import *
+from representation import *
+from sequencer import *
+from score import *
+import tools
+
+selection = db.select()
+score = db.getScore1(selection)
+notes = tools.parseScore(score)
+seq = Sequencer()
+seq.play(notes)
+#s = Score(alignment.score, alignment)
+
+

### 2011-07-02 Code/perform.py

@@ -1,0 +1,113 @@
+from representation import *
+from deviations import *
+from alignment import *
+from analysescore import *
+import structure, math
+
+# Expression vector:
+#       0       1           2             3           4       5
+# E = (ioi_r, loudness_r, articulation, duration_r, ioi_ch, loundness_ch)
+def widmerPerform(score, expression, average_loudness=80, tempo=80):
+  score.tempo = tempo
+  performance = NoteList()
+  time = 0
+  for i in range(len(score)):
+    on = time + structure.ioi(score, i) * math.exp(expression[i][0])
+    loudness = average_loudness * math.exp(expression[i][1])
+    off = on + structure.duration(score, i) * math.exp(expression[i][3])
+    performance.insert(Note(int(on), int(off), score[i].pitch, int(loudness), 0))
+    time = on
+  return performance
+
+def vanDerWeijPerformSimple(score, melody, segments, expression_vectors, average_loudness=100, bpm=120, converter=NoteList()):
+# converter is used to convert milliseconds to ticks and back
+  deviations = Deviations()
+  deviations.bpm = bpm
+  deviations.basedynamics = average_loudness
+  tempo_deviations = {}
+  alignment = {}
+
+ 
+  for segment, expression in zip(segments, expression_vectors):
+    average_tempo = expression[3]
+    for note in segment:
+      pointer = note.annotation
+      scorepart = melody[pointer[0]]
+      scoremeasure = scorepart[pointer[1]]
+      scorevoice = scoremeasure[pointer[2]]
+      scorenote = scorevoice[pointer[3]]
+      measure = scoremeasure.number
+      deviations.tempo_deviations[measure, int(scorenote.offset)] = average_tempo
+
+  # Fill in missing deviations
+  last = 1.0
+  for part in score.parts:
+    for measure in part:
+      if not isinstance(measure, m21.stream.Measure):
+        continue
+      deviations.measure_lengths[measure.number] = int(measure.barDuration.quarterLength)
+      for beat in range(int(measure.barDuration.quarterLength)):
+        if not (measure.number, beat) in deviations.tempo_deviations:
+          deviations.tempo_deviations[measure.number, beat] = last
+        else:
+          last = deviations.tempo_deviations[measure.number, beat]
+
+
+  for segment, expression in zip(segments, expression_vectors):
+    average_tempo = expression[3]
+    average_relative_loudness = expression[1]
+    articulation = expression[2]
+    for note in segment:
[…]

### 2011-07-02 Code/performancefeatures.py

@@ -26,12 +26,16 @@ def linear_fit(X, Y):
 
 
 
-def articulation(notes, i):
-  return structure.duration(notes, i)
-
 
 def vanDerWeijExpression(alignment, segments):
   performance = alignment.expressiveMelody()
+  scorenotes = []
+  average_loudness = 0
+  for s in segments:
+    for n in s:
+      scorenotes.append(n)
+      average_loudness += n.onvelocity
+  average_loudness /= float(len(scorenotes))
   # The two above should be guaranteed to be of equal length.
   # However, better be safe than sorry
   if len(performance) != sum([len(x) for x in segments]):
@@ -40,16 +44,13 @@ def vanDerWeijExpression(alignment, segments):
     print "This is good, performance length and score length match"
   
   expression = []
+  # Counter for position in notelists performance and scorenotes
   i = 0
-  index = 0
   for segment in segments:
     length = float(len(segment))
-    # Segments can't have a length of 1, the features don't work then
-    if length <=1:
-      print "WARNING: Don't know how to deal with atomic constituent, skipping, solve this!"
-      continue
-    average_articulation = 0
-    tempos = []
+    average_articulation = 0.0
+    average_ioi_ratio = 0.0
+    loudnesses = []
     for note in segment:
       pointer = note.annotation
       scorepart = alignment.melody()[pointer[0]]
@@ -58,16 +59,19 @@ def vanDerWeijExpression(alignment, segments):
       scorenote = scorevoice[pointer[3]]
       measure = scoremeasure.number
       
-      tempo = alignment.deviations.tempo_deviations[measure, int(scorenote.offset)]
-      if not tempo in tempos:
-        tempos.append(tempo)
 
       # Find out if the next note is a rest, if so use the ratio between expressive duration and duration
       useScoreDuration = False
-      # Last note of this
+      # Last note of this measure?
       if pointer[3] + 1 == len(scorevoice):
-        if pointer[1] + 1 == len(scorepart):
+        # Last measure?
+        if pointer[1] + 1 >= len(scorepart):
           useScoreDuration = True
+        # Don't know what to do if the voice in the next measure contains no notes
+        elif len(scorepart[pointer[1]+1].getElementById(scorevoice.id).notes) == 0:
+          measure = scorepart[pointer[1]+1]
+          useScoreDuration = True
+        # First note of this voice in the next measure is a rest?
[…]

### 2011-07-02 Code/performancerenderer.py

@@ -1,0 +1,112 @@
+# these files extract needed features and structural information
+import scorefeatures as sf
+import structure
+# *model contains statistical information
+# datahandling files
+from score import *
+from representation import *
+from hmm import *
+import database as db
+import alignment
+
+
+def trainHMM(features_set, expression_set):
+  hmm = HMM()
+  for work in features_set.keys():
+    features = features_set[work]
+    expression = expression_set[work]
+    obs = []
+    states = []
+    # Select the subset of features that will be used and discretise
+    for feature in features:
+      # Select subset
+      # Discretise
+      if feature[3] > 0: f[3] = 1
+      if feature[3] < 0: f[3] = -1
+      else: f[3] = 0
+      obs.append(((f[1], f[2], f[3])
+    for parameters in expression:
+      # Select subset
+      avg_rel_l = p[1]
+      avg_artic = p[2]
+      avg_tempo = p[3]
+      # Discretise
+      states.append((avg_rel_l, avg_artic, avg_tempo))
+    hmm.learn(obs, states)
+  return hmm
+
+def print_dptable(V):
+  print "    ",
+  for i in range(len(V)): print "%7s" % ("%d" % i),
+  print
+
+  for y in V[0].keys():
+      print "%.5s: " % y,
+      for t in range(len(V)):
+          print "%.7s" % ("%f" % V[t][y]),
+      print
+
+# Source: Wikipedia article on viterbi algorithm
+def viterbi(obs, hmm):
+  V = [{}]
+  path = {}
+
+  # Initialize base cases (t == 0)
+  for y in hmm.states:
+      V[0][y] = hmm.start_probability(y) * hmm.emission_probability(y, obs[0])
+      path[y] = [y]
+
+  # Run Viterbi for t > 0
+  for t in range(1,len(obs)):
+      V.append({})
+      newpath = {}
+
+      for y in hmm.states:
+          (prob, state) = max([(V[t-1][y0] * hmm.transition_probability(y0, y) *\
+              hmm.emission_probability(y, obs[t]), y0) for y0 in hmm.states])
+          V[t][y] = prob
+          newpath[y] = path[state] + [y]
+
+      # Don't need to remember the old paths
+      path = newpath
+
+  print_dptable(V)
+  (prob, state) = max([(V[len(obs) - 1][y], y) for y in states])
+  return (prob, path[state])
+
+def render(score, hmm):
+  melodyscore = Score(score).melody()
[…]

### 2011-07-02 Code/presentation.py

@@ -1,0 +1,86 @@
+import database as db
+from representation import *
+from score import *
+from alignment import *
+import util, tools, structure
+
+
+# Load all the scores and notelists
+
+#while True:
+#  print db.select()
+print ">>>> Loading materials....."
+
+schumann = ('Schumann', 'kdz007', 'ashke', 'GPO-Concert-Steinway-ver2.sf2')
+mozart1 = ('Mozart', 'snt331-3', 'nakam', 'GPO-Concert-Steinway-ver2.sf2')
+chopin1 = ('Chopin', 'wlz010', 'ashke', 'Bosendorfer PIANO / GIGA')
+mozart2 = ('Mozart', 'snt331-1', 'mo', None)
+lastig = ('Bach', 'wtc219-f', 'richt', 'GPO-Concert-Steinway-ver2.sf2')
+
+score = Score(db.getScore1(chopin1).stripTies())
+melody = score.melody()
+chopinnotes = tools.parseScore(melody)
+delta = structure.absolute_deltalist(structure.onset, chopinnotes)
+sodelta1 = structure.normalize(structure.second_order_deltalist(delta))
+
+score = Score(db.getScore1(mozart2).stripTies())
+melody = score.melody()
+mozartnotes = tools.parseScore(melody)
+delta2 = structure.square(structure.absolute_deltalist(structure.pitch, mozartnotes))
+sodelta2 = structure.normalize(structure.second_order_deltalist(delta2))
+
+s1 = structure.second_order_deltarule(chopinnotes, sodelta1, 0.1)
+s2 = structure.second_order_deltarule(mozartnotes, sodelta2, 0.1)
+
+while True:
+  choice = util.menu("Select", ['Schumann exp', 'Schumann score', 'Schumann noexp', 'chopin struct', 'lastig struct bach', 'struct moz'])
+  if choice == 0:
+    path = db.getPerformancePath1(schumann)
+    os.system("audacious {0}".format(path))
+  if choice == 1:
+    os.system("viewmxml {0}".format(db.getScorePath1(schumann)))
+  if choice == 2:
+    path = db.getScoreMidiPath1(schumann)
+    os.system("audacious {0}".format(path))
+  if choice == 3:
+    groups = structure.groupings(structure.list_to_tree(s1), 1)
+    structured_notes = []
+    loud = False
+    for group in groups:
+      if loud:
+        loud = False
+      else:
+        loud = True
+      for leaf in group:
+        if loud:
+          leaf.onvelocity = 80
+        else:
[…]

### 2011-07-02 Code/repetition.py

@@ -1,0 +1,90 @@
+from structure import *
+
+
+
+
+    
+
+def repetition(feature, notes, deltas=None):
+  if not deltas:
+    deltas = deltalist(feature, notes)
+  else:
+    print deltas
+  repetition = [0 for i in range(len(deltas))]
+  patterns = {}
+  for i in range(len(deltas)):
+    patterns[str(deltas[i])] = patterns.get(str(deltas[i]), []) + [i]
+
+  for offset in range(1, len(deltas)):
+    newpatterns = patterns.copy()
+    remove = []
+    #print "================="
+    #print patterns
+    for pattern, indices in patterns.iteritems():
+      if len(indices) < 2: continue
+      added = []
+      for index in indices[:]:
+        if index+offset >= len(deltas): continue
+          
+        new = ",".join([str(x) for x in deltas[index:index+offset+1]])
+        old = ",".join([str(x) for x in deltas[index:index+offset]])
+        #print "offset:{2} index:{3} New {0} Old {1}".format(new, old, offset, index)
+        if not index in newpatterns.get(new, []):
+          added.append(new)
+          newpatterns[new] = newpatterns.get(new, []) + [index]
+        if index in newpatterns.get(old, []) and len(newpatterns[new]) > 1:
+          newpatterns[old].remove(index)
+      # Clean up to keep the pattern dictionary small
+      for item in added:
+        if len(newpatterns[item]) < 2:
+          del newpatterns[item]
+
+    patterns = newpatterns
+
+  for key, value in patterns.iteritems():
+    if len(value) > 1:
+      if value[0] - 1 > 0:
+        p = ",".join([str(x) for x in deltas[value[0]-1:value[0] + len(key.split(","))]])
+        if patterns.get(p, []) == [x-1 for x in value]: continue
+      for index in value:
+        repetition[index] = max(repetition[index], len(key.split(",")))
+    
+  #print "Input: {0}".format(deltas)
+  #print "Output: {0}".format(repetition)
+  repetition = clean_deltalist(repetition)
+  #print "Postprocesed output: {0}".format(repetition)
+  #repetition.pop(0)
+  #repetition.append(0)
+  print repetition
+  return repetition
+      
+def clean_deltalist(deltas):
+  for i in range(0, len(deltas)):
+    current = deltas[i]
+    for j in range(i,i+current):
[…]

### 2011-07-02 Code/stats.py

@@ -1,0 +1,7 @@
+import scoremodel as sm
+import tools
+
+(features, expression) = tools.chooseFeatures()
+sm.scoremodel(features, expression)
+
+

### 2011-07-02 Code/structure.py

@@ -1,0 +1,303 @@
+from alignment import *
+from score import *
+from music21 import *
+from trees import *
+
+
+def onset(notes, i):
+  return notes[i].on
+
+def offset(notes, i):
+  return notes[i].off
+
+def pitch(notes, i):
+  return notes[i].pitch
+
+def loudness(notes, i):
+  return notes[i].onvelocity
+
+def duration(notes, i):
+  return notes[i].off - notes[i].on
+
+def virtual_silence(notes, i):
+  if i >= len(notes) - 1:
+    return 0
+  return notes[i+1].on - notes[i].off
+
+def virtual_duration(notes, i):
+  # This is not what Markwin suggests, the last notes' vd should be larger
+  if i >= len(notes) - 1:
+    return duration(notes, len(notes)-1) 
+  return notes[i+1].on - notes[i].on
+
+features = [onset, offset, pitch, loudness, duration, virtual_duration]
+
+def silence(notes, i, tolerance=1):
+  delta = notes[i].on - notes[i-1].off
+  return delta
+
+def ioi(notes, i):
+  if i == 0:
+    return 1
+  return bare_delta(notes, onset, i)
+
+def bare_delta(notes, feature, i, tolerance=1):
+  if feature == silence:
+    return silence(notes, i, tolerance)
+  else:
+    delta = feature(notes, i) - feature(notes, i-1)
+    return delta
+
+
+# Perhaps stochastic tolerance will work better?
+def absolute_delta(notes, feature, i, tolerance=1):
+  return abs(bare_delta(notes, feature, i, tolerance))
+
+def relative_delta(notes, feature, i, depth=4, tolerance=1):
+  avg_delta = 0
+  for x in range(1, depth+1):
+    if i-x <= 0: break
+    avg_delta += absolute_delta(notes, feature, i-x, tolerance)
+  if i == 1:
+    return 0
+  elif i <= depth:
+    avg_delta /= float(i-1)
+  else: 
+    avg_delta /= float(depth)
+  if avg_delta == 0: return 0
+  return abs(float(absolute_delta(notes, feature, i, tolerance)) - avg_delta)
+
+def relative_deltalist(feature, notes):
+    return [relative_delta(notes, feature, i) for i in range(1, len(notes))]
+
+def absolute_deltalist(feature, notes):
+    return [absolute_delta(notes, feature, i) for i in range(1, len(notes))]
+
+def second_order_deltalist(deltas):
+    result = []
+    for i in range(1, len(deltas)):
[…]

### 2011-07-02 Code/structure_test.py

@@ -1,0 +1,157 @@
+import database as db
+from alignment import *
+from representation import *
+from sequencer import *
+from score import *
+import tools, structure, sys
+import matplotlib.pyplot as plt
+
+
+def avg_list(lists, weights=None):
+  sumlist = []
+  if not weights:
+    weights = [1 for i in range(len(lists))]
+  for l in range(len(lists)):
+    index = 0
+    for item in lists[l]:
+      if index + 1 > len(sumlist):
+        sumlist.append(item)
+      else:
+        sumlist[index] += weights[l] * item
+      index += 1
+  result = []
+  for item in sumlist:
+    result.append(item / float(len(lists)))
+  return result
+
+def normalize(l):
+  m = float(max(l))
+  result = []
+  if(m == 0): 
+    m = 1
+  for i in l:
+    result.append(i/m)
+  return result
+
+def square(l):
+  result = []
+  for i in l:
+    result.append(i*i)
+  return result
+
+if len(sys.argv) > 1:
+  notes = sys.argv[1]
+  deltas = [int(x) for x in sys.argv[2]]
+  s3 = structure.second_order_deltarule(notes, deltas, 0)
+  print tools.recursive_print(s3)
+  sys.exit(0)
+
+
+selection = db.select()
+score = Score(db.getScore1(selection).stripTies())
+melody = score.melody()
+notes = tools.parseScore(melody)
+deltas = []
+
+
+print "COMBINED DELTA TREES"
+#for feature in [structure.pitch, structure.onset]:
+#  deltas.append(normalize(structure.repetition(feature, notes)))
+
+for feature in [structure.onset, structure.duration, structure.pitch]:
+  print ">>>> {0} absolute:".format(feature)
+  print structure.absolute_deltalist(feature, notes)
+  print ">>>> {0} relative:".format(feature)
+  print structure.relative_deltalist(feature, notes)
+
+
+#deltas.append(normalize(square(structure.absolute_deltalist(structure.onset, notes))))
+#deltas.append(normalize(structure.relative_deltalist(structure.pitch, notes)))
+
+deltas = []
+deltas.append(structure.absolute_deltalist(structure.onset, notes))
+deltas.append(square(structure.absolute_deltalist(structure.pitch, notes)))
+
+sodeltas = []
+sodeltas.append(normalize(structure.second_order_deltalist(deltas[0])))
[…]

### 2011-07-02 Code/test.py

@@ -4,16 +4,19 @@ from alignment import *
 import structure, tools
 import scorefeatures as sf
 import performancefeatures as pf
+import perform
 
 s = db.select()
 a = Alignment(db.getScore1(s), db.getDeviation1(s))
 melodyscore = a.melody()
 melody = tools.parseScore(melodyscore)
-onset = structure.bestgrouping(melody, structure.second_order_tree(structure.onset, melody, 0.1))
-pitch = structure.bestgrouping(melody, structure.second_order_tree(structure.pitch, melody, 0.1))
+onset = structure.groupings(structure.list_to_tree(structure.first_order_tree(structure.onset, melody, 0.1)), 1)
 score = sf.vanDerWeijFeatures(melodyscore, onset) 
 performance = pf.vanDerWeijExpression(a, onset)
 
 print score
 print performance
+
+seq = Sequencer()
+seq.play(perform.vanDerWeijPerformSimple(a.score, melodyscore, onset, performance, bpm=a.deviations.bpm, converter=melody))
 

### 2011-07-02 Code/tools.py

@@ -1,0 +1,121 @@
+import music21 as m21
+import os, pickle, util
+from representation import *
+
+def savePDF(score, name):
+  print "Saving to sheets/{0}.pdf".format(name)
+  ofile = open('sheets/score.xml', 'w')
+  ofile.write(score.musicxml)
+  ofile.close()
+  os.system('musescore sheets/score.xml -o sheets/{0}.pdf'.format(name))
+  
+def saveXML(score, name):
+  print "Saving to scores/{0}.xml".format(name)
+  ofile = open('scores/{0}.xml'.format(name), 'w')
+  ofile.write(score.musicxml)
+  ofile.close()
+
+def saveFeatures(features, expression):
+  while True:
+    print "Enter a name for the directory"
+    name = raw_input('')
+    print "Confirm to save to: data/{0}/features and data/{0}/expression (Y/n)".format(name)
+    answer = raw_input('')
+    if not answer == 'Y' and not answer == '':
+      continue
+    if not os.path.exists('data/{0}'.format(name)):
+      os.mkdir('data/{0}'.format(name))
+      break
+    else:
+      print "Directory exists, overwrite? (y/N)"
+      answer = raw_input('')
+      if answer == 'y':
+        break
+  f = open('data/{0}/features'.format(name), 'wb')
+  e = open('data/{0}/expression'.format(name), 'wb')
+  pickle.dump(features, f)
+  pickle.dump(expression, e)
+
+def saveCSV(features, expression):
+  while True:
+    print "Enter a name for the file"
+    name = raw_input('')
+    print "Confirm to save to: data/{0}.csv (Y/n)".format(name)
+    answer = raw_input('')
+    if not answer == 'Y' and not answer == '':
+      continue
+    if not os.path.exists('data/{0}'.format(name)):
+      break
+    else:
+      print "File exists, overwrite? (y/N)"
+      answer = raw_input('')
+      if answer == 'y':
+        break
+  out = open('data/{0}.csv'.format(name), 'w')
+  for key in features.keys():
+    f = features[key]
+    e = expression[key]
+    line = 'onset_arch, pitch_interval, duration_ratio, ioi_ratio, loudness_ratio, duration_ratio_exp\n'
+    out.write(line)
+    for fv, ev in zip(f,e):
+      line = '{0},{1},{2},{3},{4},{5}\n'.format(fv[0],fv[2],fv[3],ev[0],ev[1],ev[3])
+      out.write(line)
[…]

### 2011-07-02 Code/train.py

@@ -1,0 +1,44 @@
+import database as db
+from alignment import *
+from representation import *
+import structure, tools, scorefeatures, performancefeatures
+
+
+def trainset(composers=None, pianist=None):
+  queries = []
+  if not composers:
+    composers = db.composers
+  for composer in composers:
+    works = db.getWorks(composer)
+    for work in works:
+      if pianist:
+        if work[1] == pianist:
+          queries.append((composer, work[0], work[1], work[2]))
+      else:
+        queries.append((composer, work[0], work[1], work[2]))
+  return queries
+
+
+def train(trainset):
+  expression = {}
+  features = {}
+  print ">>>Loading scores and deviations, this will take hours and may eat all you memory"
+  for query in trainset:
+    print ">>>Loading: {0}".format(query)
+    score = db.getScore1(query)
+    deviations = db.getDeviation1(query)
+    alignment = Alignment(score, deviations)
+    melody = alignment.melody()
+    segments = structure.groupings(structure.list_to_tree(structure.first_order_tree(structure.onset, tools.parseScore(melody), 0)), 1)
+    expression[query] = performancefeatures.vanDerWeijExpression(alignment, segments)
+    features[query] = scorefeatures.vanDerWeijFeatures(melody, segments)
+
+
+  tools.saveFeatures(features, expression)
+#  tools.saveCSV(features, expression)
+
+if __name__ == "__main__":
+  import sys
+  train(trainset(sys.argv[1:]))
+
+

### 2011-07-02 Code/trees.py

@@ -1,0 +1,54 @@
+class Node:
+
+  def __init__(self, children, parent=None):
+    self.children = children
+    for child in self:
+      child.parent = self
+
+  def generateItems(self):
+    for n in self.children:
+      yield n
+
+  def __iter__(self):
+    return self.generateItems()
+
+  def __len__(self):
+    return len(self.children)
+
+  def __contains__(self, c):
+    for child in self.children:
+      if child is c:
+        return True
+    return False
+
+  def __getitem__(self, i):
+    return self.children[i]
+
+class MultiNode:
+
+  def __init__(self, nodes):
+    self.nodes = nodes
+
+  def generateItems(self):
+    for n in self.nodes:
+      yield n
+
+  def __iter__(self):
+    return self.generateItems()
+
+  def __len__(self):
+    return len(self.nodes)
+
+  def __contains__(self, n):
+    for node in self.nodes:
+      if node is n:
+        return True
+    return False
+
+  def __getitem__(self, i):
+    return self.nodes[i]
+
+class Leaf:
+
+  def __init__(self, contents):
+    self.contents = contents

### 2011-07-03 Code/hmm.py
Alignment and deviations working better, hmm implementation on the way

@@ -41,13 +41,57 @@ class HMM:
       self.lessergrams[tuple(lessergram)] = self.lessergrams.get(tuple(lessergram), 0) + 1
 
   def joint_probability(self, observation, state):
+    if not (observation, state) in self.coincedences: return 0.0
     return self.coincedences[observation, state] / float(self.coincedence_count)
 
   def start_probability(self, state):
+    if not state in self.starts: return 0.0
     return self.starts[state] / float(self.start_count)
 
   def emission_probability(self, state, observation):
+    if not (observation, state) in self.coincedences: return 0.0
     return self.coincedences[observation, state] / float(self.observations[observation])
 
   def transition_probability(self, lessergram, state):
-    return self.ngrams[tuple(lessergram + [state])] /  float(lessergrams[tuple(lessergram)])
+    if not tuple(list(lessergram) + [state]) in self.ngrams: return 0.0
+    return self.ngrams[tuple(list(lessergram) + [state])] /  float(lessergrams[tuple(lessergram)])
+
+  def print_dptable(self, V):
+    print "    ",
+    for i in range(len(V)): print "{0}".format("{0}".format(i)),
+    print
+
+    for y in V[0].keys():
+        print "{0}: ".format(y),
+        for t in range(len(V)):
+            print "{0}".format("{0}".format(V[t][y])),
+        print
+
+  # Source: Wikipedia article on viterbi algorithm
+  def viterbi(self, obs):
+    V = [{}]
+    path = {}
+
+    # Initialize base cases (t == 0)
+    for y in self.states:
+      V[0][y] = self.start_probability(y) * self.emission_probability(y, obs[0])
+      path[y] = [y]
+
+    # Run Viterbi for t > 0
+    for t in range(1,len(obs)):
+      V.append({})
+      newpath = {}
+
+      for y in self.states:
+        (prob, state) = max([(V[t-1][y0] * self.transition_probability(y0, y) *\
+          self.emission_probability(y, obs[t]), y0) for y0 in self.states])
+        V[t][y] = prob
+        newpath[y] = path[state] + [y]
+
+      # Don't need to remember the old paths
+      path = newpath
+
+    self.print_dptable(V)
+    (prob, state) = max([(V[len(obs) - 1][y], y) for y in self.states])
+    return (prob, path[state])
+

### 2011-07-03 Code/performancefeatures.py

@@ -83,7 +83,7 @@ def vanDerWeijExpression(alignment, segments):
         ioi = performance.microseconds_to_ticks(alignment.deviations.getExpressiveDuration(scoremeasure.number, scorenote.offset, scorenote.duration.quarterLength))
       else: 
         ioi = structure.ioi(performance, i+1)
-        # Ugly hack to ix something
+        # Ugly hack to fix something
         if ioi == 0:
           ioi = performance.microseconds_to_ticks(\
               alignment.deviations.getExpressiveDuration(scoremeasure.number, scorenote.offset, scorenote.duration.quarterLength))
@@ -92,7 +92,9 @@ def vanDerWeijExpression(alignment, segments):
       #    performance.microseconds_to_ticks(alignment.deviations.getExpressiveDuration(scoremeasure.number, scorenote.offset,\
       #    scorenote.duration.quarterLength)), structure.duration(performance, i), scoremeasure.number, scorenote.offset)
 
-      average_ioi_ratio += math.log(structure.ioi(performance, i) / float(structure.ioi(scorenotes, i)))
+      # Results in zero division error in Mozart: 28 (not needed anyway so let's kick it out)
+      #average_ioi_ratio += math.log(structure.ioi(performance, i) / float(structure.ioi(scorenotes, i)))
+      average_ioi_ratio += 0
       average_articulation += structure.duration(performance, i) / float(ioi)
       loudnesses.append(performance[i].onvelocity)
       i += 1

### 2011-07-03 Code/performancerenderer.py

@@ -6,8 +6,50 @@ import structure
 from score import *
 from representation import *
 from hmm import *
+from sequencer import *
 import database as db
-import alignment
+import alignment, math
+import perform
+
+
+def discretize_features(features):
+  # Select subset
+  avg_dur = features[1]
+  avg_pitch = features[2]
+  pitch_direction = features[3]
+  # Discretise
+  print avg_dur
+  avg_dur = int(math.log(avg_dur) * 20)
+  avg_pitch = int(avg_pitch / 4.0)
+  if pitch_direction > 0: pitch_direction = 1
+  if pitch_direction < 0: pitch_direction = -1
+  else: pitch_direction = 0
+  return (avg_dur, avg_pitch, pitch_direction)
+  
+
+def discretize_expression(parameters):
+  # Select subset
+  avg_rel_l = parameters[1]
+  avg_artic = parameters[2]
+  avg_tempo = parameters[3]
+  # Discretise
+  # Perhaps it is better to take this logarithm into performancefeatures
+  avg_rel_l = int(math.log(avg_rel_l) * 20)
+  avg_artic = int(avg_artic * 20)
+  avg_tempo = int(math.log(avg_tempo) * 20)
+  return (avg_rel_l, avg_artic, avg_tempo)
+
+def undiscretize(state):
+  # Select subset
+  avg_rel_l = state[0]
+  avg_artic = state[1]
+  avg_tempo = state[2]
+  # Discretise
+  # Perhaps it is better to take this logarithm into performancefeatures
+  avg_rel_l = math.exp(avg_rel_l / 20.0)
+  avg_artic = avg_artic / 20.0
+  avg_tempo = math.exp(avg_tempo / 20.0)
+  return (avg_rel_l, avg_artic, avg_tempo)
 
 
 def trainHMM(features_set, expression_set):
@@ -18,74 +60,37 @@ def trainHMM(features_set, expression_set):
     obs = []
     states = []
     # Select the subset of features that will be used and discretise
-    for feature in features:
-      # Select subset
-      # Discretise
-      if feature[3] > 0: f[3] = 1
-      if feature[3] < 0: f[3] = -1
-      else: f[3] = 0
-      obs.append(((f[1], f[2], f[3])
-    for parameters in expression:
-      # Select subset
-      avg_rel_l = p[1]
-      avg_artic = p[2]
-      avg_tempo = p[3]
-      # Discretise
-      states.append((avg_rel_l, avg_artic, avg_tempo))
+    for f in features:
[…]

### 2011-07-03 Code/scorefeatures.py

@@ -27,13 +27,13 @@ def vanDerWeijFeatures(melodyscore, segments):
     length = len(segments[i]) 
     # Calculate averages:
     # What if the segment is len 1?
-    abs_interval = sum([1/float(length - 1 ) * \
-      structure.absolute_delta(segments[i], structure.pitch, x) for x in range(1, length)])
-    duration = sum([1/float(length) * \
-      structure.duration(segments[i], x) for x in range(length)])
-    pitch = sum([1/float(length) * \
-      structure.pitch(segments[i], x) for x in range(length)])
+    abs_interval = 1/float(length) * sum([structure.absolute_delta(segments[i], structure.pitch, x) for x in range(1, length)])
+    duration = 1/float(length) * sum([structure.duration(segments[i], x) for x in range(length)])
+    pitch = 1/float(length) * sum([structure.pitch(segments[i], x) for x in range(length)])
     pitch_direction = (segments[i][length-1].pitch - segments[i][0].pitch ) / float(length)
+    # Average of deltafunctions?
+    # Average of second order deltafunctions?
+    # Average of log relatives?
     features.append((pitch, duration, abs_interval, pitch_direction))
 
   return features

### 2011-07-06 Code/alignment.py
Improved scorefeatures, improved renderer usability

@@ -18,35 +18,12 @@ class Alignment:
       self.score = scorepath
     s = Score(self.score)
     m = s.melody()
-    self.m = self.removeTies(m)
-    self.score = self.removeTies(self.score)
+    self.score = Score(score.tieless())
     if not noAlign:
       self.alignment = None
       self.align()
 
     print "Done"
-
-  def removeTies(self, score):
-    # This doesn't actually remove ties, but it does fix the duration of tied notes
-    score.stripTies(inPlace=True)
-    # Now remove every note that is not the start of a tie
-    for part in score:
-      if not isinstance(part, m21.stream.Part):
-        continue
-      for measure in part:
-        if not isinstance(measure, m21.stream.Measure):
-          continue
-        for voice in measure:
-          if not isinstance(voice, m21.stream.Voice):
-            continue
-          for note in voice:
-            if isinstance(note, m21.chord.Chord) or isinstance(note, m21.note.Note):
-              if hasattr(note,'tie') and note.tie is not None:
-                if not note.tie.type == 'start':
-                  voice.remove(note)
-                else:
-                  note.tie = None
-    return score
 
 
   def storeAlignment(self):

### 2011-07-06 Code/hmm.py

@@ -21,6 +21,8 @@ class HMM:
     self.start_count = 0
     self.startstate = ('start',)
     self.endstate = ('end',)
+    # This doesn't really belong here, it is just used as convenient storage
+    self.normalizations = []
 
   def learn(self, observations, states):
     parameters = len(states[0])
@@ -83,6 +85,30 @@ class HMM:
             print "{0}".format("{0}".format(V[t][y])),
         print
 
+  def max_emission(self, obs):
+    states = []
+    p = 1.0
+    for o in obs:
+      probs = [self.emission_probability(state, o) for state in self.states]
+      m = max(probs)
+      if m == 0.0:
+        print o
+      p *= m
+      states.append(self.states[probs.index(m)])
+    return (p, states)
+
+  def sequence_probability(self, obs, states):
+    states = [self.startstate] + states + [self.endstate]
+    obs = obs + [self.endstate]
+    p = 1.0
+    for i in range(len(states)-2):
+      cp = self.emission_probability(states[i+1], obs[i]) * self.transition_probability([states[i]], states[i+1])
+      if cp == 0.0:
+        print '{0}. {1}, {2}'.format(states[i], states[i+1], self.emission_probability(states[i+1], obs[i]))
+      p *= cp
+    return p
+
+
   # Source: Wikipedia article on viterbi algorithm
   def viterbi(self, obs):
     obs = obs + [self.endstate]
@@ -103,8 +129,6 @@ class HMM:
         #print [self.transition_probability(y0, y) for y0 in self.states]
         (prob, state) = max([(V[t-1][y0] * self.transition_probability([y], y0) *\
           self.emission_probability(y, obs[t]), y0) for y0 in self.states])
-        #if prob > 0:
-          #print self.transition_probability([y], state, True)
         V[t][y] = prob
         newpath[y] = path[state] + [y]
 

### 2011-07-06 Code/performancerenderer.py

@@ -10,20 +10,17 @@ import database as db
 import alignment, math
 import perform, util, structure
 
-def discretize_features(features, multiplication=10, division=5.0):
-  division = float(division)
+#(dPitch, abs_dPitch, dDuration, abs_dDuration, silence, ddPitch, abs_ddPitch, pitch_direction)
+def discretize_features(features, normalizations, discretization=10, subset=None):
+  f = []
+  r = range(len(features))
+  if subset:
+    r = subset
+  for i in r:
+    f.append(int((features[i] / float(normalizations[i])) * discretization))
+  return tuple(f)
   # Select subset
-  avg_dur = features[1]
-  avg_pitch = features[2]
-  pitch_direction = features[3]
   # Discretise
-# avg dur klopt niet
-  avg_dur = int(avg_dur / multiplication)
-  avg_pitch = int(avg_pitch / division)
-  if pitch_direction > 0: pitch_direction = 1
-  if pitch_direction < 0: pitch_direction = -1
-  else: pitch_direction = 0
-  return (avg_dur, avg_pitch, pitch_direction)
   
 
 def discretize_expression(parameters, multiplication=30):
@@ -38,7 +35,7 @@ def discretize_expression(parameters, multiplication=30):
   avg_tempo = int(math.log(avg_tempo) * multiplication)
   return (avg_rel_l, avg_artic, avg_tempo)
 
-def undiscretize(state, multiplication=30):
+def undiscretize(state, multiplication=30.0):
   multiplication = float(multiplication)
   # Select subset
   avg_rel_l = state[0]
@@ -52,7 +49,19 @@ def undiscretize(state, multiplication=30):
   return (avg_rel_l, avg_artic, avg_tempo)
 
 
-def trainHMM(hmm, features_set, expression_set, f_discretization=10, e_discretization=30):
+def trainHMM(hmm, features_set, expression_set, f_discretization=10, e_discretization=30, subset=None):
+  l = 0
+  for work in features_set.keys():
+    l = len(features_set[work][0])
+    break
+  normalizations = [0 for i in range(l)]
+  for work in features_set.keys():
+    m = []
+    for i in range(l):
+      m = max([f[i] for f in features_set[work]])
+      if m > normalizations[i]:
+        normalizations[i] = m
+  hmm.normalizations = normalizations
   for work in features_set.keys():
     features = features_set[work]
     expression = expression_set[work]
@@ -60,34 +69,44 @@ def trainHMM(hmm, features_set, expression_set, f_discretization=10, e_discretiz
     states = []
[…]

### 2011-07-06 Code/score.py

@@ -7,18 +7,31 @@ class Score:
     self.score = score
     # Assume each part has the same number of measures and every score has two parts
     parts = []
-    for part in self.score:
-      if isinstance(part,m21.stream.Part):
-        parts.append(part)
-    self.part1 = parts[0]
-    self.part2 = parts[1]
-
-  def notesOn(self,measure,beat):
-    m1 = self.part1[measure]
-    m2 = self.part2[measure]
   
   def getNoteList():
     return tools.parseScore(self.score)
+
+  def tieless(self):
+    # This doesn't actually remove ties, but it does fix the duration of tied notes
+    score = self.score.stripTies()
+    # Now remove every note that is not the start of a tie
+    for part in score:
+      if not isinstance(part, m21.stream.Part):
+        continue
+      for measure in part:
+        if not isinstance(measure, m21.stream.Measure):
+          continue
+        for voice in measure:
+          if not isinstance(voice, m21.stream.Voice):
+            continue
+          for note in voice:
+            if isinstance(note, m21.chord.Chord) or isinstance(note, m21.note.Note):
+              if hasattr(note,'tie') and note.tie is not None:
+                if not note.tie.type == 'start':
+                  voice.remove(note)
+                else:
+                  note.tie = None
+    return score
 
   #Highest note in part one at any moment
   def melody(self):
@@ -63,5 +76,5 @@ class Score:
                       n.duration = note.duration
                       n.id = note.id
                       voice.append(n)
-    return melody
+    return Score(melody).tieless()
 

### 2011-07-06 Code/scorefeatures.py

@@ -1,13 +1,8 @@
 import tools, structure, math
+import performancefeatures as pf
 
 
 # Take a score, split it in melody and harmony, do structure analysis and output features per note.
-
-def discretize(min, max, value, bins=10):
-  if value < min or value > max:
-    print "Discretize error: value outside range, returning 0"
-    return 0
-  return int((value - min) * (bins / float(max - min)))
 
 def pitch_interval(notes, i):
   if i == 0: return 0
@@ -27,14 +22,30 @@ def vanDerWeijFeatures(melodyscore, segments):
     length = len(segments[i]) 
     # Calculate averages:
     # What if the segment is len 1?
-    abs_interval = 1/float(length) * sum([structure.absolute_delta(segments[i], structure.pitch, x) for x in range(1, length)])
-    duration = 1/float(length) * sum([structure.duration(segments[i], x) for x in range(length)])
-    pitch = 1/float(length) * sum([structure.pitch(segments[i], x) for x in range(length)])
-    pitch_direction = (segments[i][length-1].pitch - segments[i][0].pitch ) / float(length)
+    pitch_int = structure.bare_deltalist(structure.pitch, segments[i])
+    abs_pitch_int = structure.absolute_deltalist(structure.pitch, segments[i])
+    dPitch = 1/float(length) * sum(pitch_int)
+    abs_dPitch = 1/float(length) * sum(abs_pitch_int)
+    dDuration = 1/float(length) * sum(structure.bare_deltalist(structure.duration, segments[i]))
+    abs_dDuration = 1/float(length) * sum(structure.absolute_deltalist(structure.duration, segments[i]))
+    silence = 1/float(length) * sum([structure.silence(segments[i], j+1) for j in range(len(segments[i])-1)])
+    ddPitch = 1/float(length) * sum(structure.second_order_deltalist(pitch_int))
+    abs_ddPitch = 1/float(length) * sum(structure.second_order_deltalist(abs_pitch_int))
+    
+    #print [structure.silence(segments[i], j) for j in range(len(segments[i]))]
+
+    onsets = structure.normalize([n.on for n in segments[i]])
+    pitches = structure.normalize([n.pitch for n in segments[i]])
+    if len(onsets) == 1:
+      pitch_direction = 0
+    else:
+      pitch_direction = pf.linear_fit(onsets, pitches)[1]
+
     # Average of deltafunctions?
     # Average of second order deltafunctions?
     # Average of log relatives?
[…]

### 2011-07-06 Code/structure.py

@@ -81,6 +81,9 @@ def relative_deltalist(feature, notes):
 def absolute_deltalist(feature, notes):
     return [absolute_delta(notes, feature, i) for i in range(1, len(notes))]
 
+def bare_deltalist(feature, notes):
+    return [bare_delta(notes, feature, i) for i in range(1, len(notes))]
+
 def second_order_deltalist(deltas):
     result = []
     for i in range(1, len(deltas)):

### 2011-07-07 Paper/Paper.tex
HMM now works as it should. Last commit broke that #%@! alignment again

@@
 
+Although certainly not all expression is at structural level we may plausibly assume that \emph{something} is happening at structural level.
+The goal is to capture this dimension of expression in a performance rendering system.
+The resulting performances will demonstrate that expression at structural level is can be extracted and is meaningful
+
 \subsection{Musical structure}

### 2011-07-07 Code/alignment.py

@@ -16,9 +16,8 @@ class Alignment:
       self.score = m21.converter.parse(scorepath)
     else:
       self.score = scorepath
-    s = Score(self.score)
-    m = s.melody()
-    self.score = Score(score.tieless())
+    s = Score(Score(self.score).tieless())
+    self.m = s.melody()
     if not noAlign:
       self.alignment = None
       self.align()

### 2011-07-07 Code/hmm.py

@@ -8,9 +8,12 @@ class HMM:
   def __init__(self, order=2):
     self.order = order
 
+    self.startstate = ('start',)
+    self.endstate = ('end',)
+
     self.coincedences = {}
     self.observations = {}
-    self.states = []
+    self.states = [self.startstate, self.endstate]
 
     self.ngrams = {}
     self.coincedence_count = 0
@@ -19,8 +22,6 @@ class HMM:
     self.lessergram_count = 0
     self.starts = {}
     self.start_count = 0
-    self.startstate = ('start',)
-    self.endstate = ('end',)
     # This doesn't really belong here, it is just used as convenient storage
     self.normalizations = []
 
@@ -29,12 +30,9 @@ class HMM:
     observations = [self.startstate] + observations + [self.endstate]
     states = [self.startstate] + states + [self.endstate]
     for observation, state in zip(observations, states):
-      if state == self.startstate or state == self.endstate:
-        state = [state for i in range(parameters)]
-      for i in range(parameters):
-        self.coincedences[i, observation, state[i]] = self.coincedences.get((i, observation, state[i]), 0) + 1
-      self.observations[observation] = self.observations.get((i, observation), 0) + 1
-      if not state in self.states and not (state[0] == self.startstate or state[0] == self.endstate):
+      self.coincedences[observation, state] = self.coincedences.get((observation, state), 0) + 1
+      self.observations[observation] = self.observations.get((observation), 0) + 1
+      if not state in self.states and not (state == self.startstate or state == self.endstate):
         self.states.append(state)
     for i in range(len(states)-self.order + 1):
       ngram = []
@@ -68,11 +66,37 @@ class HMM:
     return self.coincedences[observation, state] / float(self.observations[observation])
 
   def transition_probability(self, lessergram, state, verbose=False):
-    if verbose:
-      print '{0}, {1}, {2}, {3}'.format(tuple(list(lessergram) + [state]), self.ngrams.get(tuple(list(lessergram) + [state]), 0), tuple(lessergram), self.lessergrams.get(tuple(lessergram), 0))
     if self.order == 1: return 1.0
     if not tuple(list(lessergram) + [state]) in self.ngrams: return 0.0
-    return self.ngrams[tuple(list(lessergram) + [state])] /  float(self.lessergrams[tuple(lessergram)])
[…]

### 2011-07-07 Code/performancerenderer.py

@@ -1,3 +1,4 @@
+#!/usr/bin/env python
 # these files extract needed features and structural information
 import scorefeatures as sf
 # *model contains statistical information
@@ -95,6 +96,7 @@ def render(score, segmentation, hmm, f_discretization=10, e_discretization=30, s
   # Find the best expressive explanation for these features
   print "Finding best fitting expression"
   (p, states) = hmm.viterbi(observations)
+  print states
   (prob, simple) = hmm.max_emission(observations)
   simple_p = hmm.sequence_probability(observations, simple)
   hmm.sequence_probability(observations, states)

### 2011-07-11 Paper/Paper.tex
Updated paper, added more functionality to performancerenderer.py (started on supporting music from the music21 corpus)

--- 2011-06-21 Paper/Paper.tex
@@
 \begin{abstract}
-This paper introduces a system for expressively performing transcribed music (scores) using statistical learning from a dataset.
+Both machine learning and rule based techniques have been extensively applied to performance rendering.
+However, relatively few systems make explicit use of musical structure.
+This paper introduces a performance rendering approach that tries to learn expression exclusively at a structural level.
+The approach attempts to to be complementary to approaches that learn expression at note level.
+Musical structure is extracted automatically.
+A dataset containing performances aligned to scores is then used to learn how structure relates to expression.
+Finally, the model can be applied to a a new score to produce an expressive performance of the score.
+
+%This paper introduces a system for expressively performing transcribed music (scores) using statistical learning from a dataset.
 The dataset that is used contains performances aligned to the corresponding score.
@@
 
-A music performance involves a performer and an audience.
-The performer tries his best to perform music to the audience in a way that pleases the audience.
-In western art music tradition it is conventional that the performer provides his rendition of a musical score: a detailed transcription of the music in musical notation.
-A performer should be skilled enough to accurately reproduce the music in the score, but what sets a good performer apart from other performers and drives people to pay money to see a musical performance is his ability deviate from the score and interpret it in a way that pleases his audience.
-In fact, it is generally agreed on that for most western art music, a perfect reproduction of the score sounds very unpleasant.
-The performer makes the performance sound `natural' and expresses his musical knowledge and perhaps feelings in the performance, the resulting performance is expressive.
+It is customary in Western art music tradition that composers write down their compositions in scores.
+The task of a performer is to some extend to accurately reproduce the score, however, a perfect reproduction of a score generally sounds robotic and unpleasant.
+What makes a performance appealing is that the performer deviates from the score, altering timing, articulation and loudness, creating an expressive performance.
 
-It should be clear that performing a musical piece is not a trivial task.
-Theoretically it would be possible to construct robotic performers with perfect mastery of their instruments as well as perfect ability to reproduce scores.
-Yet there is still large gap between a perfect score reproduction by robots and a good sounding performance (the perfect reproduction sounds, indeed, robotic).
-Performance rendering systems attempt to bridge this gap.
+YQX is state of the art
 
-Deviations from a score in a performance can be described as the end result of a number of processes, the most important of which is called expression.
-It is not immediately clear where expression comes for or even what it is.
-It is not the within the scope of this thesis to elaborate widely on the complex phenomenon of expression.
-Instead, it will be assumed that expression is strongly linked to musical structure and that a computational model using this structure can to some extend adequately model expression.
-The quality of
-the system should be an indication of the correctness of this assumption.
+Rencon
 
-Early performance rendering systems used rule-based approaches to recognise small local structures that are used as indications to add certain kind of expression.
+%Early performance rendering systems used rule-based approaches to recognise small local structures that are used as indications to add certain kind of expression.
 A prominent ongoing rule based system, \textit{Director Musices} uses 30 rules with weighting parameters that take music features as input and produce expressive actions as output.
@@
 
-As datasets become more available, machine learning approaches are gaining popularity.
+%As datasets become more available, machine learning approaches are gaining popularity.
 However, very few approaches make extensive use of structure.
@@
 
-This thesis introduces machine learning system that makes of local structure as well as higher level structure to generate expressive performances of polyphonic piano music.
+\subsection{Motivation}
+
+The YQX system, as widmer admitted, tended to sometimes produce nervous sounding changes in expression.
+He presents two extensions that seek to generate smoother performances.
+The problem with these extensions is, as widmer admits himself, that the increased smoothness comes at the expense of expressitivity.
+To counter this, he adds three explicit rules that he uses to postprocess the performances.
+We think the reason that Widmer stumbled upon this tradeoff between expressiveness and nervousness because the nervousness is inherent to the way performances are generated at note-level.
+The system simply misses one component of expression, namely structure, that Widmer had to artificially introduce using rules.
+
+Structure seems to be directly related to expression and it seems logical that a performance rendering system should have some notion of usical structure.
+The system presented here will use structure exclusively to generate performances and will completely ignore note level expression.
+The expressiveness of the performances it generates is therefore limited.
+In section XX we will describe how this system could be integrated with a note level performance rendering system to hopefully produce some variant of YQX that doesn't need explicit rules to produce expressive performances.
+
+%This thesis introduces machine learning system that makes of local structure as well as higher level structure to generate expressive performances of polyphonic piano music.
 Section X will introduce Y, section P will elaborate on Q and finally section Z will discuss the results.
-
-\paragraph*{Evaluation.} Performance context, adaptability, creativity. Subjective listening, correlation.
-
-\subsection{Related Work}
-Move part of the introduction here?
 
@@
 \subsection{Musical structure}
-When listening to music, the listener assigns a certain hierarchical structure to the music perceived: Notes make up phrases, phrases make up themes, themes make up parts and parts make up the piece.
-This structure is usually accentuated in a performance.
-The most obvious example of this is slowing down at the end of a piece.
-On a lower level, transitions between themes may be marked by slowing down, changing volume or changing speed.
-Although structure can be accentuated in a performance, listeners, and trained listeners even more so, will also be able to recognise structure in an expressionless reproduction of the score, or when quietly singing the music in their heads.
-This suggests that it is possible to extract structure from a merely score.
-It also seems likely that knowing the structure will help creating an expressive performance.
+When listening to music, the listener's musical intuition assigns a certain hierarchical structure to the music: Notes make up phrases, phrases make up themes and themes make up a piece.
+In a performance, this structure may be accentuated in different ways.
+Accentuation happens at different levels, at note level performers may slow down at the end a phrase or introduce small pauses in the performance at phrase transitions.
+At constituent level one phrase may be played very loud, fast or staccato, while the next may be played slow, soft and legato.
+
+To formally describe musical structure, we can look at music in a way similar to the way we look at natural language processing(NLP).
+In this analogy we see a piece of music as a sentence, which consists of constituents that individually can be made of constituents as well.
+We can recursively subdivide constituents in more constituents until we reach a terminal symbol.
+In NLP this may be a word, in music, this may be a note.
+We could represent musical structure as a parse tree.
+This paradigm corresponds to the intuition that a melody is not simply a sequence of notes but that notes form phrases and we can look at phrases at different levels.
+
+It must be noted that the resulting parse tree can be highly ambiguous and even experienced listeners may not agree on the correct parsetree of a piece.
+Quite often there may simply be more than one parse tree that makes musical sense.
+This should not be a problem for a performance rendering system: different expressive interpretations of one piece can be very diverse and still be accepted as sensible interpretations of the piece.
+As long as the parse tree does make at least some musical sense, a performance rendering system should be able to use it.
+
+Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, it still only looks at note level expression.
+The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''.
+We see this as a symptom of a note level expression based system.
+
+In this thesis, we propose a structure based performance rendering (SBPR) system.
+The system presented here ignores note level expression.
+Instead we will only try to predict constituent level expression.
+The assumption is that this kind of expression really exists in performances and that is different and independend from note level structure.
+We think that a constituent level system also corresponds better to how actual human performers play music.
+A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.
+
+The system will be similar to YQX in a number of ways, but instead of predicting expression per note, it will predict expression per constituent.
+Every constituent will be played with consistend expression.
+A number of score features defined per constituent will be used to predict epressive parameters of the constituent.
+
+The simplification of ignoring note level expression is of course unjustified and severly limits the expressiveness of the performances that can be generated.
+However, if succesful, the resulting performances will clearly demonstrate the phenomenon of constituent level expression.
+It should be possible to extend the system to incorporate note level expression, section \ref{integration} will further explore this possibility.
+
+The succes of a structure based performance rendering (SBPR) system depends largely on two factors.
+The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents.
+The following two sections address these issues.
+
+%The following section will introduce an application of the delta framework \ref{markwin} that intends to segment a score into musically meaningfull constituents.
+Section \ref{scorefeatures} will deal with the characterization of individual constituents, section \ref{expressionfeatures} will explain how expressive parameters of a constituent are determined and translated into expressioin.
+Section \ref{performancerendering} will explain how performances are
+
+\section{The Delta Framework}
+
+In his Phd thesis \ref{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework.
+He relates his work to the work of Lerdahl and Jackendoff \ref{lerdahl1996generative} but claims to have found a more general approach.
+
+The delta framework is based on the intuition that differences between notes indicate splits between constituents.
+The higher the difference, the higher the level of split in the parse tree (where the root note is at the highest level).
+Van der Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parsetree following this intuition.
+
+The differences between notes are defined as the difference in value of a certain note feature. More formally, we can look at a piece of music as a sequence of notes, ordered by onset time:
+
+\[M = n_i, n_{i+1}, \cdots, n_j\]
+
+We can assign features to these notes.
+
+\subsection{Extracting structure}
+
+\begin{itemize}
+\item We only need a segmentation for now
+\item Using the deltaframework to define segmentation
+\end{itemize}
+
+\section{Constituent Features}
+
+We can now convert a piece of music into a serie of constituents.
+These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed.
+Clearly, expressive markings in the score play a role in how the constituent should be played.
+If the segmentation is specific enough  there will hopefully be at most expressive marking within the constituent.
+
+\section{Performance Rendering}
+\section{Dealing with polyphony}
+\subsection{Dataset and representation}
+
+Distinction of expressive tempo and attack and release.
 
@@
 
-The delta framework is based on the idea that small differences indicate low level structural transitions and large differences indicate higher level structural transitions.
-
-\paragraph*{Combining delta trees}
-
-In his MSc dissertation, van den Berg suggests using \textit{yeild rules} to combine trees generated by different delta functions.
-The yeild rule states that nodes in a set of trees can be connected if they are \textit{yeild equal}, that is, if the their set of leaf nodes they recursively contain is equal.
-The result is a multi-tree, in which the top note is always connected and intermediate nodes may or may not be connected.
-
-Unfortunately, a multi tree is hard to interpret since
-
-\paragraph*{Using repetition to improve structural analysis}
-
-Repetition be formulated in terms of the delta framework to provide greater generality.
-It seems desirable that we consider a transposed repetition of an earlier theme to be a repetition nevertheless.
-The same holds for rhythmic repetition: it may be that a particular rhythm is repeated, we want to treat this as a repetition too.
-The delta framework allows this intuition to be implemented easily: we only look at repetition in the deltas between notes.
-If we take for example the inter
----------------------------
-Consider the musical sequence $S = \{n_0, n_1, n_2, n_3\}$ is ambiguous and has two possible assignments of constituents: $S_0 = \{n_0\}, S_1 = \{n_1, n_2, n_3\}$ or $S'_0 = \{n_0, n_1\}, S'_1 = \{n_2, n_3\}$.
-The notion off repetition can help us here.
-Consider the case that $S'_1$ is a repetition of $S'_0$ but $S_1$ is not a repetition of $S_0$, this tells us that the split $S = \{S'_0, S'_1\}$ is more likely than the split $S = \{S_0, S_1\}$.
-
-\paragraph*{Using repetition to create constituent classes}
-
-Repetition can also be used to make the constituent assignment more interesting. What follows is a method that identifies which consituents are repetitions of which other constituents.
-
-\begin{algorithm}
-\begin{algorithmic}
-\FOR{$\texttt{depth} \in \{\texttt{depth}(\texttt{tree}), \cdots, 0\}$}
-  \FOR{$c \in \{c | \texttt{depth}(c, \texttt{tree}) = \texttt{depth}\}$}
-    \FOR{$\texttt{class} \in C$}
-      \IF{$\texttt{similar}(c, \texttt{class})$}
-        \STATE c.class = class
-      \ENDIF
-    \ENDFOR
-    \IF{c.class = null}
-      \STATE C.append(createClass(c))
-      \STATE c.class = createClass(c)
-    \ENDIF
-  \ENDFOR
-\ENDFOR
-\end{algorithmic}
-\caption{Bottom up tree traversal and similarity relations}
-\end{algorithm}
-
-A resulting tree could look like figure \ref{similtree}
-
-\begin{figure}
-\begin{center}
-\Tree 
-[
-.{$M$}
-	[ .{$d$} 
-		[ .{$a$} 
-			[ .{$x$} ]
-			[ .{$y$} ]
-			[ .{$x$} ]
-		]
-		[ .{$a$} 
-			[ .{$x$} ]
-			[ .{$y$} ]
-			[ .{$x$} ]
-		]
-	]	
-	[ .{$e$} 
-		[.{$c$} 
-			[ .{$b$} 
-				[ .{$p$} ]
-				[ .{$q$} ]
-			]
-			[ .{$r$} ]
-		]		
-		[.{$a$} 
-			[ .{$x$} ]
-			[ .{$y$} ]
-			[ .{$x$} ]
-		]
-	]
-]
-\end{center}
-\caption{Example tree with similarities}
-\label{similtree}
-\end{figure}
-
 \subsection{Performance Model}
-\subsubsection{Smoothness of expression}
-To avoid sudden changes in expression that sound weird, you would want a model that describes what a sensible sequence of expressive actions is.
-It may be sensible to train this model on all available data since this model should not describe what expressive actions are appropriate for one specific pianist or composer, it should rather describe what sequences of expressive actions are acceptable in general.
-To achieve this, a continuous hidden markov model, trained on all performances in the dataset, will be used.
-
-Let $D$ be a collection of compositions.
-Let a composition, $c$ be a collection of expressive actions, $a$ and let an expressive action be a tuple $(l,t)$ describing the loudness deviation, timing deviation (and/or local tempo?) of the note.
-The hidden markov model can be described as follows:
-\begin{align*}
-D &= \{c_0, c_1, \cdots, c_n\}\\
-c_i &= \{a_0, a_1, \cdots, a_n\}\\
-a_i &= (l,t)\\
-P(c) &= P(a_1 | a_0) * P(a_2|a_1, a_0) * \cdots * P(a_n|a_{n-1}, a_{n-2}, \cdots, a_0))
-\end{align*}
 
@@
 \section{Evaluation and Results}
+The proposed system actually requires more training data to do meaningful statistics. However, since expressive interpretation can be 
 \subsection{Subjective Listening}
 \subsection{Correlation}
+\section{Integration with Note Level Performance Rendering}
+
+The YQX system defines expressive tempo implicitly by predicting the logarithmic ratio of the IOI in a performance and the IOI in a score.
+Timing alterations of notes are always defined relative to the base tempo.
+This does not correspond to the intuition that the \textit{the tempo itself} is altered during the performance and that rhythmic changes should be seen relative to the local tempo.
+The same applies to the way YQX looks at loudness.
+This is specified as the logarithmic ratio between the notes loudness and the average loudness of the performance.
+
+An intergration of constituent level expression and note level expression can provide a solution to this problem.
+We can define expressive tempo relative and dynamics relative to the expression parameters of the constituent
+
 \section{Conclusion}
+\cite{markwin}
+\section{Discussion}
+
+Unfortunately we do not have acces to the large dataset that YQX uses.
+The dataset we use is smaller.
+Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger.
+We have discretize into rather large bins for this reason, resulting in cartoonistisc performances.
+
+We think we can afford to do this because:...
+
 \section{Future Work}
+\subsection{Repetition}
+A notion of repetition and similarity would certainly improve the system.
+Repetition is a very good indicator of constituent breaks.
+Repetition and similarity could also be used to improve expressiveness of performances.
+It is probably telling when a phrase is repeated three times and then slightly altered the fourth time.
+Although finding similarity and musically significant repetition is a subject of its own the delta framework could help to define repetition arbitrarily of transposition or rhythm.
+A list of pitch deltas can for example be used to detect repetition independend of transposition and a list of duration deltas can be used to detect repetition of rhythm independend of the notes used.
 

### 2011-07-11 Code/alignment.py

@@ -16,14 +16,16 @@ class Alignment:
       self.score = m21.converter.parse(scorepath)
     else:
       self.score = scorepath
-    s = Score(Score(self.score).tieless())
+    import score
+
+    s = score.Score(self.score)
     self.m = s.melody()
+    self.m = score.Score(self.m).tieless()
+    self.s = s.tieless()
     if not noAlign:
       self.alignment = None
       self.align()
-
     print "Done"
-
 
   def storeAlignment(self):
     location = 'alignments/{0}'.format(self.deviations.target.split('.')[0])
@@ -237,7 +239,7 @@ class Alignment:
               if query in deviations.note_deviations: break 
             # This situation is usually caused by fucking TIED NOTES, it can mess up indexing
             if not query in deviations.note_deviations:
-              print "Query: {0} not found in deviations".format(query)
+              print "WARNING: Query: {0} not found in deviations".format(query)
               n_dev = None
             else:
               n_dev = deviations.note_deviations[query]

### 2011-07-11 Code/database.py

@@ -32,7 +32,10 @@ def getPerformancePath1(t):  return getPerformancePath(t[0], t[1], t[2], t[3])
 def getScorePath1(t):  return getScorePath(t[0], t[1], t[2], t[3])
 def getDeviationPath1(t):  return getDeviationPath(t[0], t[1], t[2], t[3])
 def getPerformance1(t):  return getPerformance(t[0], t[1], t[2], t[3])
-def getScore1(t):  return getScore(t[0], t[1], t[2], t[3])
+def getScore1(t):
+  if t == 'm21':
+    return selectM21Score()
+  return getScore(t[0], t[1], t[2], t[3])
 def getDeviation1(t):  return getDeviation(t[0], t[1], t[2], t[3])
 
 def getAllDirectories():
@@ -184,10 +187,22 @@ def sampleScore(): return m21.converter.parse("./samples/score.xml")
 def sampleDeviationPath(): return "./samples/deviation.xml"
 def sampleDeviation(): return Deviations("./samples/deviation.xml")
 
+def selectM21Score():
+  composers = ['bach', 'beethoven', 'ciconia', 'coltrane', 'handel', 'haydn', 'josquin',\
+      'luca', 'monteverdi', 'mozart', 'pachelbel', 'schoenberg', 'schubert', 'schumann']
+  path = -1
+  while(path == -1):
+    comp = util.menu('Choose a composer', composers)
+    path = util.menu('Choose a work', m21.corpus.getComposer(composers[comp]), cancel=True)
+  return m21.corpus.parse(m21.corpus.getComposer(composers[comp])[path])
+
 def select():
   work = -1
   while(work == -1):
-    comp = util.menu('Choose a composer', getComposers())
+    comp = util.menu('Choose a composer', getComposers() + ['MUSIC21 Corpus'])
+    print len(getComposers()), comp
+    if comp == len(getComposers()):
+      return 'm21'
     work = util.menu('Choose a work and performer', getWorks(getComposers()[comp]), cancel=True)
 
   selected = getWorks(getComposers()[comp])[work]

### 2011-07-11 Code/hmm.py

@@ -13,7 +13,7 @@ class HMM:
 
     self.coincedences = {}
     self.observations = {}
-    self.states = [self.startstate, self.endstate]
+    self.states = []
 
     self.ngrams = {}
     self.coincedence_count = 0
@@ -84,11 +84,8 @@ class HMM:
     return (p, states)
 
   def sequence_probability(self, obs, states):
-    print '----------------'
-    print len(states), len (obs)
     states = [self.startstate] + states + [self.endstate]
     obs = obs + [self.endstate]
-    print len(states), len (obs)
     p = 1.0
     for i in range(len(states)-1):
       cp = self.emission_probability(states[i+1], obs[i]) * self.transition_probability([states[i]], states[i+1])
@@ -114,12 +111,10 @@ class HMM:
     obs = obs + [self.endstate]
     V = [{}]
     path = {}
-    print self.states
 
     # Initialize base cases (t == 0)
     for y in self.states:
       V[0][y] = self.transition_probability([self.startstate], y) * self.emission_probability(y, obs[0])
-      if V[0][y] > 0: print '{0} {1}'.format(y, V[0][y])
       path[y] = [y]
 
     # Run Viterbi for t > 0
@@ -127,13 +122,9 @@ class HMM:
       V.append({})
       newpath = {}
 
-      for y in self.states:
+      for y in self.states + [self.endstate]:
         (prob, state) = max([(V[t-1][y0] * self.transition_probability([y0], y) *\
           self.emission_probability(y, obs[t]), y0) for y0 in self.states])
-        # If everything is zero, we can only assume to keep playing like we were, can't we?
-        # This is ugly, better use the self loops Remco was talking about but I don't know how!
-        if prob == 0:
-          state = y
         V[t][y] = prob
         newpath[y] = path[state] + [y]
 
@@ -141,7 +132,7 @@ class HMM:
       path = newpath
 
     #self.print_dptable(V)
-    (prob, state) = max([(V[len(obs) - 1][y], y) for y in self.states])
+    (prob, state) = max([(V[len(obs) - 1][y], y) for y in self.states + [self.endstate]])
     # Exclude the last node from the path (this is the end state that slipped in)
     return (prob, path[state][:-1])
 
@@ -195,10 +186,6 @@ class HMM_indep(HMM):
   def __str__(self):
[…]

### 2011-07-11 Code/performancefeatures.py

@@ -39,9 +39,7 @@ def vanDerWeijExpression(alignment, segments):
   # The two above should be guaranteed to be of equal length.
   # However, better be safe than sorry
   if len(performance) != sum([len(x) for x in segments]):
-    print 'This shouldn\'t happen: melodyscore and performance lengths don\'t match: {0} and {1}'.format(len(sum([len(x) for x in segments])), len(performance))
-  else:
-    print "This is good, performance length and score length match"
+    print 'WARNING: Melodyscore and performance are different (should not happen): {0} and {1}'.format(len(sum([len(x) for x in segments])), len(performance))
   
   expression = []
   # Counter for position in notelists performance and scorenotes

### 2011-07-11 Code/performancerenderer.py

@@ -12,16 +12,39 @@ import alignment, math
 import perform, util, structure
 
 #(dPitch, abs_dPitch, dDuration, abs_dDuration, silence, ddPitch, abs_ddPitch, pitch_direction)
-def discretize_features(features, normalizations, discretization=10, subset=None):
+def discretize_features(features, discretization=10):
   f = []
   r = range(len(features))
-  if subset:
-    r = subset
-  for i in r:
-    f.append(int((features[i] / float(normalizations[i])) * discretization))
+  for i in range(len(features)):
+    f.append(int(features[i] * discretization))
   return tuple(f)
   # Select subset
   # Discretise
+
+def preprocess(features, normalizations, discretization=10, subset=None):
+  observations = []
+  lastPitch = 0
+  lastDuration = 0
+  for f in features:
+    # Normalize
+    f = [f[i] / float(normalizations[i]) for i in range(len(f))]
+    pitch = f[sf.featureset.index('avg_pitch')]
+    duration = f[sf.featureset.index('avg_duration')]
+    pitch_interval = 0
+    duration_ratio = 0 
+    if lastPitch != 0:
+      pitch_interval = pitch - lastPitch
+      duration_ratio = math.log(duration/float(lastDuration))
+    f[sf.featureset.index('avg_pitch')] = pitch_interval
+    f[sf.featureset.index('avg_duration')] = duration_ratio
+    lastPitch = pitch
+    lastDuration = duration
+    obs = f
+    if subset:
+      obs = [obs[i] for i in subset]
+    observations.append(discretize_features(obs, discretization))
+  return observations
+    
   
 
 def discretize_expression(parameters, multiplication=30):
@@ -66,55 +89,80 @@ def trainHMM(hmm, features_set, expression_set, f_discretization=10, e_discretiz
   for work in features_set.keys():
     features = features_set[work]
     expression = expression_set[work]
-    obs = []
+    obs = preprocess(features, normalizations, f_discretization, subset)
     states = []
     # Select the subset of features that will be used and discretise
-    for f in features:
-      obs.append(discretize_features(f, normalizations, f_discretization, subset=subset))
-    if work == ('Mozart', 'snt545-2', 'gould', 'GPO-Concert-Steinway-ver2.sf2'):
-      print obs
-
-    lastexp = 'start'
-    for p in expression:
-      #  Multiplication for expression and features doesn't neccesarily need to be the same!
[…]

### 2011-07-11 Code/score.py

@@ -13,9 +13,9 @@ class Score:
 
   def tieless(self):
     # This doesn't actually remove ties, but it does fix the duration of tied notes
-    score = self.score.stripTies()
+    self.score.stripTies(inPlace=True)
     # Now remove every note that is not the start of a tie
-    for part in score:
+    for part in self.score:
       if not isinstance(part, m21.stream.Part):
         continue
       for measure in part:
@@ -31,7 +31,7 @@ class Score:
                   voice.remove(note)
                 else:
                   note.tie = None
-    return score
+    return self.score
 
   #Highest note in part one at any moment
   def melody(self):

### 2011-07-11 Code/scorefeatures.py

@@ -2,7 +2,10 @@ import tools, structure, math
 import performancefeatures as pf
 
 
-# Take a score, split it in melody and harmony, do structure analysis and output features per note.
+featureset = ['avg_pitch', 'dPitch', 'abs_dPitch', 'ddPitch', 'abs_ddPitch',\
+    'avg_duration', 'dDuration', 'abs_dDuration', 'ddDuration',\
+    'abs_ddDuration', 'silence', 'pitch_direction']
+version = '0.02'
 
 def pitch_interval(notes, i):
   if i == 0: return 0
@@ -12,6 +15,9 @@ def duration_ratio(notes, i):
   if i == 0: return 1
   return math.log((notes[i].off - notes[i].on) / float(notes[i-1].off - notes[i-1].on))
 
+def getFeature(name, features):
+  return features[featureset.index(name)]
+
 def vanDerWeijFeatures(melodyscore, segments):
   # This is radically different, we only look at features at constituent level
   melody = tools.parseScore(melodyscore)
@@ -19,21 +25,30 @@ def vanDerWeijFeatures(melodyscore, segments):
   features = []
   index = 0
   for i in range(len(segments)):
+    #if i > 0:
+    #  segments[i] = [segments[i-1][len(segments[i-1])-1]] + segments[i]
     length = len(segments[i]) 
-    # Calculate averages:
-    # What if the segment is len 1?
+    # Calculate features:
+
     pitch_int = structure.bare_deltalist(structure.pitch, segments[i])
     abs_pitch_int = structure.absolute_deltalist(structure.pitch, segments[i])
+    pitches = [structure.pitch(segments[i], j) for j in range(len(segments[i]))]
+    durations = [structure.duration(segments[i], j) for j in range(len(segments[i]))]
+
+    avg_pitch = 1/float(length) * sum(pitches)
     dPitch = 1/float(length) * sum(pitch_int)
     abs_dPitch = 1/float(length) * sum(abs_pitch_int)
-    dDuration = 1/float(length) * sum(structure.bare_deltalist(structure.duration, segments[i]))
-    abs_dDuration = 1/float(length) * sum(structure.absolute_deltalist(structure.duration, segments[i]))
-    silence = 1/float(length) * sum([structure.silence(segments[i], j+1) for j in range(len(segments[i])-1)])
     ddPitch = 1/float(length) * sum(structure.second_order_deltalist(pitch_int))
[…]

### 2011-07-11 Code/tools.py

@@ -1,5 +1,5 @@
 import music21 as m21
-import os, pickle, util
+import os, pickle, util, scorefeatures
 from representation import *
 
 def savePDF(score, name):
@@ -38,8 +38,12 @@ def saveFeatures(features, expression):
         break
   f = open('data/{0}/features'.format(name), 'wb')
   e = open('data/{0}/expression'.format(name), 'wb')
+  m = open('data/{0}/metadata'.format(name), 'wb')
+
+  metadata = {'version':scorefeatures.version, 'featureset':scorefeatures.featureset}
   pickle.dump(features, f)
   pickle.dump(expression, e)
+  pickle.dump(metadata, m)
 
 def saveCSV(features, expression):
   while True:
@@ -77,11 +81,16 @@ def loadPerformance(name=None):
 def loadFeatures(name):
   f = open('data/{0}/features'.format(name), 'rb')
   e = open('data/{0}/expression'.format(name), 'rb')
-  return (pickle.load(f), pickle.load(e))
+  m = open('data/{0}/metadata'.format(name), 'rb')
+
+  return (pickle.load(f), pickle.load(e), pickle.load(m))
   
 def chooseFeatures():
   choice = util.menu('Pick a dataset', os.listdir('data'))
   return loadFeatures(os.listdir('data')[choice])
+
+def datasets():
+  return os.listdir('data')
 
 # This doesn't handle polyfony very well but it is used for single voices only anyway
 def newParseScore(score):

### 2011-07-11 Code/train.py

@@ -25,9 +25,10 @@ def train(trainset):
   const = 0
   Max = 0
   Min = None
-  print ">>>Loading scores and deviations, this will take hours and may eat all you memory"
+  count = 0
+  print ">>> Loading scores and deviations, this will take hours and may eat all you memory"
   for query in trainset:
-    print ">>>Loading: {0}".format(query)
+    print ">>> Loading: {0}".format(query)
     score = db.getScore1(query)
     deviations = db.getDeviation1(query)
     alignment = Alignment(score, deviations)
@@ -43,8 +44,11 @@ def train(trainset):
       Min = mi
     if mi < Min:
       Min = mi
+    print '>>> Extracting features'
     expression[query] = performancefeatures.vanDerWeijExpression(alignment, segments)
     features[query] = scorefeatures.vanDerWeijFeatures(melody, segments)
+    count += 1
+    print '{0}/{1} done'.format(count, len(trainset))
 
   print "Done, {0} segments found with an average length of: {1} (min: {2} max: {3})".format(const, lengths / float(const), Min, Max)
   tools.saveFeatures(features, expression)

### 2011-07-11 Code/util.py

@@ -1,21 +1,20 @@
 def menu(question, options, cancel=False):
   while True:
-    c = 0
+    c = 1
     print question
+    if cancel:
+      print "\t0: Cancel"
     for option in options:
       print "\t{0}: {1}".format(c,option)
       c += 1
-    if cancel:
-      print "\t{0}: Cancel".format(c)
     inp = raw_input("?: ")
-    max = len(options)
-    if cancel: max += 1
     try:
       choice = int(inp)
     except:
       continue
-    if choice < 0 or choice >= max:
+    if choice < 0 or choice > len(options):
+      continue
+    if choice == 0 and not cancel:
       continue
     break
-  if choice == c: choice = -1
-  return choice
+  return choice - 1

### 2011-07-11 Paper/Paper.tex
Updated paper. Added support for never occuring observations although not sure if it is working. Also: logistic discretisation! And more relative features

@@
 Both machine learning and rule based techniques have been extensively applied to performance rendering.
-However, relatively few systems make explicit use of musical structure.
+However, relatively few systems make explicit use of musical structure, combined with machine learning.
 This paper introduces a performance rendering approach that tries to learn expression exclusively at a structural level.
@@
 
-It is customary in Western art music tradition that composers write down their compositions in scores.
+In Western art music tradition it is customary that composers write down their compositions in scores.
 The task of a performer is to some extend to accurately reproduce the score, however, a perfect reproduction of a score generally sounds robotic and unpleasant.
 What makes a performance appealing is that the performer deviates from the score, altering timing, articulation and loudness, creating an expressive performance.
+
+Given a hypothetical perfect synthesizer, performing music with computers is a trivial task.
+However \textit{expressively} performing music is not and much research has focussed on this issue.
+Computers are widely used in music production.
+Since support for automatic expression is mostly absent or very poor in music editing software, some genres of music have evolved to sound acceptable even without expression.
+If automatic expression better in this software, computer music production could greatly widen its field of application.
+
+Some music transcription
+Performance rendering
 
@@
 We think the reason that Widmer stumbled upon this tradeoff between expressiveness and nervousness because the nervousness is inherent to the way performances are generated at note-level.
-The system simply misses one component of expression, namely structure, that Widmer had to artificially introduce using rules.
+The system simply misses one component of expression, namely structure level expression.
+(that Widmer had to artificially introduce using rules.
+) ?
 
@@
 These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed.
-Clearly, expressive markings in the score play a role in how the constituent should be played.
-If the segmentation is specific enough  there will hopefully be at most expressive marking within the constituent.
+Analogous to YQX we are looking for the \textit{context} of the constituent.
+
+(Clearly, expressive markings in the score play a role in how the constituent should be played.
+If the segmentation is specific enough  there will hopefully be at most expressive marking within each constituent.
+)?
+
+\section{Targets}
+
+\begin{itemize}
+\item Use average expression parameters per constituent
+\end{itemize}
 
 \section{Performance Rendering}
-\section{Dealing with polyphony}
+\begin{itemize}
+\item Extract melody
+\item Extract scorefeatures from melody
+\item lookup notes in deviation, ignore missing notes
+\item Extract expressive and non-expressive melody and extract expression features
+\item Perform notes other than melody notes the same as the nearest melody note
+\end{itemize}
 \subsection{Dataset and representation}
+\begin{itemize}
+\item attack release
+\item tempo curves
+\end{itemize}
+\subsection{Dealing with polyphony}
+\subsection{Discretization}
 
@@
 
+\begin{itemize}
+\item Top notes is not always the melody, musical attention
+\item Bass and harmony shouldn't be played with the same expression as melodynotes
+\end{itemize}
+
 \section{Future Work}
+\begin{itemize}
+\item Use loudness and tempo direction instead of averages (requires more data)
+\item Generalize approach to incorporate hierarchical structure, let structure extend to note level, so note level expression and structurelevel expression become integrated
+\end{itemize}
 \subsection{Repetition}

### 2011-07-11 Code/database.py

@@ -216,36 +216,24 @@ if __name__ == "__main__":
   import sequencer
   seq = sequencer.Sequencer()
   while(True):
-    comp = util.menu('Choose a composer', getComposers())
-    work = util.menu('Choose a work and performer', getWorks(getComposers()[comp]), cancel=True)
-    if work == -1: continue
-    selected = getWorks(getComposers()[comp])[work]
+    selection = select()
     choice = 0
     while choice != 8:
-      choice = util.menu('What should I load?', ['Expressive performance', 'Score without expression'])
-      if choice == 0:
-        path = getPerformancePath(getComposers()[comp], selected[0], selected[1], selected[2])
-      else:
-        path = getScoreMidiPath(getComposers()[comp], selected[0], selected[1], selected[2])
-
-      selection = (getComposers()[comp], selected[0], selected[1], selected[2])
-      print path
 
       choice = util.menu('Choose action', \
           ['Play with internal sequencer', 'Play with audacious', 'View score', 'View midi info', 'View score info',\
           'Export deviation data to CSV', 'Performance', 'Extract melody','Cancel'])
       if choice == 0:
-        seq.play(NoteList(path))
+        seq.play(NoteList(getScoreMidiPath1(selection)))
       elif choice == 1:
-        os.system("audacious {0}".format(path))
+        os.system("audacious {0}".format(getScorePath1(selection)))
       elif choice == 2:
-        os.system("viewmxml {0}".format(getScorePath(getComposers()[comp], selected[0], selected[1], selected[2])))
-        pass
+        getScore1(selection).show()
       elif choice == 3:
-        nlist = NoteList(path)
+        nlist = NoteList(getScoreMidiPath1(selection))
         nlist.printinfo()
       elif choice == 4:
-        score = getScore(getComposers()[comp], selected[0], selected[1], selected[2])
+        score = getScore1(selection)
         parts = 0
         notes = 0
         for part in score:
@@ -261,13 +249,13 @@ if __name__ == "__main__":
           parts += 1
         print "{0} Parts in piece, total notes: {1}".format(parts, notes)
       elif choice == 5:
-        os.system("cmx dev2csv {0} > deviation.csv".format(getDeviationPath(getComposers()[comp], selected[0], selected[1], selected[2])))
+        os.system("cmx dev2csv {0} > deviation.csv".format(getDeviationPath1(selection)))
         os.system("vim deviation.csv")
       elif choice == 6:
[…]

### 2011-07-11 Code/hmm.py

@@ -10,6 +10,7 @@ class HMM:
 
     self.startstate = ('start',)
     self.endstate = ('end',)
+    self.repeatstate = ('repeat',)
 
     self.coincedences = {}
     self.observations = {}
@@ -62,15 +63,20 @@ class HMM:
     return self.starts[state] / float(self.start_count)
 
   def emission_probability(self, state, observation):
-    if not (observation, state) in self.coincedences: return 0.0
+    if not (observation, state) in self.coincedences: 
+      return 0.0
     return self.coincedences[observation, state] / float(self.observations[observation])
 
   def transition_probability(self, lessergram, state, verbose=False):
     if self.order == 1: return 1.0
-    if not tuple(list(lessergram) + [state]) in self.ngrams: return 0.0
-    p = self.ngrams[tuple(list(lessergram) + [state])] /  float(self.lessergrams[tuple(lessergram)])
+    p = self.ngrams.get(tuple(list(lessergram) + [state]), 0) /  float(self.lessergrams[tuple(lessergram)])
     if verbose:
       print '{0}, {1}, {2}, {3} {4}'.format(tuple(list(lessergram) + [state]), self.ngrams.get(tuple(list(lessergram) + [state]), 0), tuple(lessergram), self.lessergrams.get(tuple(lessergram), 0), p)
+    # If this ngram was not found and the following state is the same as the last state (expression remains the same) do not return zero
+    # These are the self-loops
+    if p==0 and lessergram[len(lessergram)-1] == state:
+      minimum = 1 /  float(self.lessergrams[tuple(lessergram)])
+      return minimum
     return p
 
   def max_emission(self, obs):
@@ -108,6 +114,15 @@ class HMM:
 
   # Source: Wikipedia article on viterbi algorithm
   def viterbi(self, obs):
+    # If an observation is not in the corpus we would like either find the nearest observation, or keep using the same expression
+    # So what do we do? We could add a special entry for this observation that emission_probabilty propagates to transition_probability
+    # Who in turn propagates it to viterbi who uses the last state
+    # But this is ugly
+    # Better idea: add a specal repeat state and make this the maximum probability for the unfound states! 
+    for o in obs:
+      if not o in self.observations:
+        print self.emission_probability(self.repeatstate, o)
+        print "{0} this observation is not in the corpus :(".format(o)
     obs = obs + [self.endstate]
     V = [{}]
[…]

### 2011-07-11 Code/performancerenderer.py

@@ -11,69 +11,98 @@ import database as db
 import alignment, math
 import perform, util, structure
 
-#(dPitch, abs_dPitch, dDuration, abs_dDuration, silence, ddPitch, abs_ddPitch, pitch_direction)
-def discretize_features(features, discretization=10):
+def sigmoid(x, p=1.0):
+  return 1/(1+math.exp(-p*x))
+
+def inv_sigmoid(y, p=1.0):
+  return -math.log((1/float(y) - 1))/float(p)
+
+# Given normalized features use some function to discretize.
+# Possibilities: linear or sigmoid
+# Use a logistic function to get a larger resolution for small differences than for larger ones
+def discretize(features, discretization=10, function='sigmoid', logarithmic=False):
   f = []
-  r = range(len(features))
-  for i in range(len(features)):
-    f.append(int(features[i] * discretization))
-  return tuple(f)
-  # Select subset
-  # Discretise
+  # WORKAROUND
+  if logarithmic:
+    for i in range(len(features)):
+      if features[i] < 0: 
+        features[i] = -features[i]
+        print "Warning: invalid featurevalue found, {0} in {1}".format(features[i], features)
+    features = [math.log(x) for x in features]
+  if function == 'linear':
+    for i in range(len(features)):
+      f.append(int(round(features[i] * discretization)))
+  elif function == 'sigmoid':
+    for i in range(len(features)):
+      f.append(int(round(sigmoid(features[i], 5) * discretization)))
+  return f
+
+def undiscretize(features, discretization=10, function='sigmoid', logarithmic=False):
+  features = list(features)
+  discretization = float(discretization)
+  f = []
+  if function == 'linear':
+    for i in range(len(features)):
+      f.append(features[i] / discretization)
+  elif function == 'sigmoid':
+    for i in range(len(features)):
+      # Discretization may have made a features the minumum or max value, cap this to 1 and 9
+      if features[i] == 0:
+        features[i] = 1        
+      if features[i] == discretization:
+        features[i] = 9        
+      f.append(inv_sigmoid(features[i] / discretization, 5))
+  if logarithmic:
+    f = [math.exp(x) for x in f]
+  return f
+
+def discretize_expression(parameters, discretization=10, subset=[1,2,3]):
+  if subset != []:
+    parameters = [parameters[i] for i in subset]
[…]

### 2011-07-11 Code/scorefeatures.py

@@ -4,7 +4,8 @@ import performancefeatures as pf
 
 featureset = ['avg_pitch', 'dPitch', 'abs_dPitch', 'ddPitch', 'abs_ddPitch',\
     'avg_duration', 'dDuration', 'abs_dDuration', 'ddDuration',\
-    'abs_ddDuration', 'silence', 'pitch_direction']
+    'abs_ddDuration', 'silence', 'pitch_direction', 'const_length']
+virtual_features = ['d_avg_pitch', 'd_avg_duration', 'abs_pitch_interval_ratio', 'd_abs_dDuration']
 version = '0.02'
 
 def pitch_interval(notes, i):

### 2011-07-12 Paper/Paper.tex
Updated paper, reformulated relative features to correspond to Widmer

@@
 
-Structure seems to be directly related to expression and it seems logical that a performance rendering system should have some notion of usical structure.
 The system presented here will use structure exclusively to generate performances and will completely ignore note level expression.
+Section \ref{approach} will describe the system as a whole.
+What we mean by structure level expression is clarified in section \ref{structure}.
+The approach for extracting structure is described in section \ref{deltaframework}.
+
 The expressiveness of the performances it generates is therefore limited.
@@
 Section X will introduce Y, section P will elaborate on Q and finally section Z will discuss the results.
-
-\section{Approach}
 
@@
 
-Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, it still only looks at note level expression.
+Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, expression is still only predicted per note.
 The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''.
@@
 
+\section{Approach}
+
 In this thesis, we propose a structure based performance rendering (SBPR) system.
 The system presented here ignores note level expression.
-Instead we will only try to predict constituent level expression.
+Instead we will try to predict only constituent level expression.
 The assumption is that this kind of expression really exists in performances and that is different and independend from note level structure.
@@
 
+%\subsection{Structure Based Performance Rendering}
+
 The system will be similar to YQX in a number of ways, but instead of predicting expression per note, it will predict expression per constituent.
-Every constituent will be played with consistend expression.
-A number of score features defined per constituent will be used to predict epressive parameters of the constituent.
+Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.
+
+We use a corpus that contains performances and corresponding scores of Western classical piano music.
+Every note in every performance has been associated with the corresponding score note so we can see exactly how every note in the score was played.
+The performances are of high quality and played by famous pianists.
+See section \ref{corpus} for more details on the corpus.
+
+A structural analysis is used to derrive hierarchical structure of every score in the corpus, however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents.
+After segmentation, four score features, two of which are direct generalizations of YQX's score features, are extracted for each constituent.
+
+So far, we have only used the score.
+Since we have a segmentation and every score note is associated with a performance note\footnote{In reality, not every score note is associated with a performance note since the pianist may have missed some notes.
+These notes will be ignored} we can also define expression per constituent.
+Three parameters, analogous to YQX's targets, will be used to describe expression per constituent.
+
+The segmentation, score features and expression parameters are based only on the \textit{melody} of the piece. In this case, melody is defined to be the highest notes in the top staff.
+
+The resulting data is used to train a first order hidden Markov model.
+The system uses this model to generate performances given a score.
+To do this, the score is segmented into constituents, score features are extracted for each constituent.
+Finally viterbi decoding is used to find the sequence of expression parameters that best explains the observed score features.
 
 The simplification of ignoring note level expression is of course unjustified and severly limits the expressiveness of the performances that can be generated.
-However, if succesful, the resulting performances will clearly demonstrate the phenomenon of constituent level expression.
-It should be possible to extend the system to incorporate note level expression, section \ref{integration} will further explore this possibility.
+However, if succesful, the resulting performances will clearly demonstrate the phenomenon of structure level expression.
+A succesful performance rendering system should incorporate both structure and note level expression, section \ref{integration} will explore two possibilities for an integration of SBPR and note level performance rendering.
 
-The succes of a structure based performance rendering (SBPR) system depends largely on two factors.
-The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents.
-The following two sections address these issues.
+The succes of a SBPR system depends largely on two factors.
+The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents in score features.
+The following section address these issues.
+
+\begin{itemize}
+\item Mention GTTM
+\item Give short overview of the system?
+\end{itemize}
 
@@
 
-\section{The Delta Framework}
+\section{Method}
+
+This section will describe individual components of the system sketched in section \ref{approach} in more detail.
+The structural analysis of is based on the delta framework, which will be described in section \ref{deltaframework}.
+Section \ref{segmentation} will discuss how the delta framework is applied to get a segmentation.
+Sections \ref{features} and \ref{targets} will describe how the scorefeatures and expression parameters are calculated.
+Section \ref{model} will describe how a hidden Markov model is trained on our data.
+
+\subsection{The Delta Framework}
 
@@
 He relates his work to the work of Lerdahl and Jackendoff \ref{lerdahl1996generative} but claims to have found a more general approach.
+Below, I will shortly describe the deltaframework as proposed by Van der Berg.
 
@@
 
-\[M = n_i, n_{i+1}, \cdots, n_j\]
+\[M_{ij} = [n_i, n_{i+1}, \cdots, n_j]\]
 
-We can assign features to these notes.
+A set of basic features, $\Phi$, is assigned to each note.
+These are: \textbf{onset}, \textbf{pitch}, \textbf{loudness} and \textbf{release} (called offset by Van den Berg).
+From these two other features can be derived: \textbf{duration} and \textbf{virtual duration}.
+Duration is defined as \textbf{release($n_i$)} - \textbf{onset($n_i$)} while virtual duration is defined as \textbf{onset($n_{i+1}$)} - \textbf{onset($n_i$)}.
 
-\subsection{Extracting structure}
+The basic note features can be used to define delta functions, for example $\Delta\textbf{Pitch} = \textbf{Pitch}(n_i) - \textbf{Pitch}(n_{i-1})$.
+In general, a delta function $\delta(i)$ is defined as the difference of two notes in some feature $\phi$: $\delta(i) = \phi(n_i) - \phi(n_{i-1})$.
+We can apply a delta function to every pair of succeeding notes in a sequence to get a list of deltas:
+\[\Delta M_{ij} = [\delta(n_{i+1}), \delta(n_{i+2}), \cdots, \delta(n_{j})]\]
+
+A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure.
+The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed, is defined as follows\footnote{The version here is a slightly reformulated version of Van den Berg's deltarule}:
+\begin{algorithm}
+\begin{algorithmic}
+\STATE{$D_{i+1,j} \leftarrow \Delta M_{ij}$}
+\STATE{$A \leftarrow []$}
+\STATE $m \leftarrow \textbf{ max}(D)$
+\FOR{$\delta \textbf{ in } D$}
+\IF{$\delta = m$}
+\STATE $p = \textbf{ index of }\delta \textbf{ in } D$
+\STATE $q = \textbf{ index of the next occurence of m in }D \textbf{ or } j$
+\STATE $\textbf{append DeltaRule(}M_{pq}\textbf{) to } A$
+\ENDIF
+\ENDFOR
+\RETURN{$A$}
+\end{algorithmic}
+\caption{The delta rule}
+\label{deltarule}
+\end{algorithm}
+
+It also possible to define higher order delta function (deltas of deltas).
+Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches.
+Van den Berg notes that this generates ambigous grouping structures as for example a second order deltarule needs at least three notes to be specified.
+For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$.
+Because we want to use a second order deltafunction for segmentation only, we choose to only use one interpretation: $[n_1, n_2], n_3]$.
+EXPLAIN WHY?
+\subsection{Segmentation}
+\label{segmentation}
 
@@
 
-\section{Constituent Features}
+\subsection{Constituent Features}
+\label{scorefeatures}
+
+\begin{itemize}
+\item We are talking about SCOREFEATURES here, mention this
+\end{itemize}
 
@@
 These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed.
-Analogous to YQX we are looking for the \textit{context} of the constituent.
+Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.
+
+YQX uses a set of three score features: pitch interval, duration ratio and I-R arch.
+The pitch interval is simply the difference in pitch between the current note and the next note.
+The duration ratio is the logarithmic ratio of the duration of the current note and the duration of the next note.
+The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis.
+(Too literally?)
+
+We can generalize pitch interval and duration ratio per note to constituent features: \textit{average pitch interval} and \textit{average duration ratio}.
+Definitions can be found in table \ref{features}.
+Since I-R arch is related to note-level expression it does not generalize well to a constituent level feature.
+
+The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in average pitch and average duration.
+At note level there is not much to say about the current note besides the pitch and duration.
+However at constituent we would also like to say something about the constituent itself.
+For this purpose \textit{average delta pitch} and \textit{average delta duration} are used.
+These features say something about the amount change in pitch and the amount of change in note duration.
+
+The complete set of score features consists of two \textit{context features} and two \textit{constituent features}.
+
+\begin{table}
+\begin{itemize}
+\item[\textbf{Average pitch interval}] 
+The difference between the average pitch of the current constituent and the average pitch of the next constituent, zero if the current is the last constituent
+\item[\textbf{Average duration ratio}]
+The logarithmic ratio between the average note duration of the current constituent and the average note duration of the next constituent
+\end{itemize}
+\caption{Context features}
+\label{features}
+\end{table}
+
+\begin{table}
+\begin{itemize}
+\item[\textbf{Average delta pitch}] The average of all absolute pitch intervals within one constituent. 
+\item[\textbf{Average delta duration}] The average of all absolute differences in duration of succeeding notes within one constituent
+\end{itemize}
+\caption{Constituent Features}
+\label{features}
+\end{table}
 
@@
 
-\section{Targets}
+\subsection{Expression Parameters}
+\label{targets}
 
@@
 
+\subsection{Model}
+
+\section{Implementation}
 \section{Performance Rendering}
@@
 
-\subsubsection*{The delta framework}
-
-\subsection{Performance Model}
-
-\subsubsection{Expressive timing} Honing and Desain, Honing and timmers, Honing.
-\subsection{Learning}
-
-\section{Method}
 \section{Evaluation and Results}
@@
 Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger.
-We have discretize into rather large bins for this reason, resulting in cartoonistisc performances.
+We have discretize into rather large bins for this reason, resulting in cartoonistic performances.
 
@@
 \item Bass and harmony shouldn't be played with the same expression as melodynotes
+\item Dataset size
+\item Targets for expression level expression
 \end{itemize}
@@
 \begin{itemize}
+\item Combine deltatrees
 \item Use loudness and tempo direction instead of averages (requires more data)

### 2011-07-12 Code/hmm.py

@@ -17,7 +17,7 @@ class HMM:
     self.states = []
 
     self.ngrams = {}
-    self.coincedence_count = 0
+    self.observation_count = 0
     self.ngram_count = 0
     self.lessergrams = {}
     self.lessergram_count = 0
@@ -33,6 +33,7 @@ class HMM:
     for observation, state in zip(observations, states):
       self.coincedences[observation, state] = self.coincedences.get((observation, state), 0) + 1
       self.observations[observation] = self.observations.get((observation), 0) + 1
+      self.observation_count += 1
       if not state in self.states and not (state == self.startstate or state == self.endstate):
         self.states.append(state)
     for i in range(len(states)-self.order + 1):
@@ -119,10 +120,11 @@ class HMM:
     # Who in turn propagates it to viterbi who uses the last state
     # But this is ugly
     # Better idea: add a specal repeat state and make this the maximum probability for the unfound states! 
+    unseen = 0
     for o in obs:
       if not o in self.observations:
-        print self.emission_probability(self.repeatstate, o)
-        print "{0} this observation is not in the corpus :(".format(o)
+        unseen += 1
+    print "{0} out of {1} observations do not occur in the corpus!".format(unseen, len(obs))
     obs = obs + [self.endstate]
     V = [{}]
     path = {}
@@ -183,6 +185,7 @@ class HMM_indep(HMM):
       for i in range(parameters):
         self.coincedences[i, observation, state[i]] = self.coincedences.get((i, observation, state[i]), 0) + 1
       self.observations[observation] = self.observations.get(observation, 0) + 1
+      self.observation_count += 1
       if not state in self.states and not (state[0] == self.startstate or state[0] == self.endstate):
         self.states.append(state)
     for i in range(len(states)-self.order + 1):
@@ -212,7 +215,7 @@ class HMM_indep(HMM):
     p = 1.0
     # This must be the absolute minimum probability
     if state == self.repeatstate:
-      return 1 / float(len(self.observations))
+      return 1 / float(self.observation_count)
     if state == self.startstate:
       if observation == self.startstate:
         return 1.0

### 2011-07-12 Code/performancerenderer.py

@@ -66,43 +66,49 @@ def undiscretize_expression(state, discretization=10):
 
 def preprocess(features, normalizations, discretization=10, subset=None):
   observations = []
-  lastPitch = 0
-  lastDuration = 0
-  last_dPitch = 0
-  last_dDuration = 0
-  for f in features:
+  for i in range(len(features)):
     # Normalize
-    f = [f[i] / float(normalizations[i]) for i in range(len(f))]
-    pitch = f[sf.featureset.index('avg_pitch')]
-    duration = f[sf.featureset.index('avg_duration')]
-    dPitch = f[sf.featureset.index('abs_dPitch')]
-    dDuration = f[sf.featureset.index('abs_dDuration')]
-    pitch_interval = 0
-    duration_ratio = 0 
-    dPitch_interval = 0
-    dDuration_ratio = 0
-    if lastPitch != 0:
-      pitch_interval = pitch - lastPitch
-      duration_ratio = math.log(duration/float(lastDuration))
-      dPitch_interval = dPitch - last_dPitch
-      #duration_ratio = math.log(dDuration/float(last_dDuration))
-    f[sf.featureset.index('avg_pitch')] = pitch_interval
-    f[sf.featureset.index('avg_duration')] = duration_ratio
-    lastPitch = pitch
-    lastDuration = duration
-    last_dPitch = dPitch
-    last_dDuration = dDuration
-    obs = f
+    features[i] = [features[i][j] / float(normalizations[j]) for j in range(len(features[i]))]
+  for f0, f1 in zip(features[:-1], features[1:]):
+    pitch0 = f0[sf.featureset.index('avg_pitch')]
+    duration0 = f0[sf.featureset.index('avg_duration')]
+    dPitch0 = f0[sf.featureset.index('abs_dPitch')]
+    dDuration0 = f0[sf.featureset.index('abs_dDuration')]
+    pitch1 = f1[sf.featureset.index('avg_pitch')]
+    duration1 = f1[sf.featureset.index('avg_duration')]
+    dPitch1 = f1[sf.featureset.index('abs_dPitch')]
+    dDuration1 = f1[sf.featureset.index('abs_dDuration')]
+    # Difference in average pitch
+    pitch_interval = (pitch0 - pitch1) * 10 # 10 is a rather arbitrary number to make this feature visible in discretization
+    # Relative difference in average note duration
+    duration_ratio = math.log(duration0/float(duration1))
+    # Do the next two features make any musical sense?
[…]

### 2011-07-12 Code/scorefeatures.py

@@ -42,6 +42,7 @@ def vanDerWeijFeatures(melodyscore, segments):
     ddPitch = 1/float(length) * sum(structure.second_order_deltalist(pitch_int))
     abs_ddPitch = 1/float(length) * sum(structure.second_order_deltalist(abs_pitch_int))
 
+    avg_duration_ratio = 0
     avg_duration = 1/float(length) * sum(durations)
     dDuration = 1/float(length) * sum(structure.bare_deltalist(structure.duration, segments[i]))
     abs_dDuration = 1/float(length) * sum(structure.absolute_deltalist(structure.duration, segments[i]))

### 2011-07-12 Code/tools.py

@@ -86,11 +86,30 @@ def loadFeatures(name):
   return (pickle.load(f), pickle.load(e), pickle.load(m))
   
 def chooseFeatures():
-  choice = util.menu('Pick a dataset', os.listdir('data'))
+  choice = util.menu('Pick a dataset', ['Name:\t[{0}]\tInfo:\t{1}'.format(x, corpusInfo(x)) for x in os.listdir('data')])
   return loadFeatures(os.listdir('data')[choice])
 
 def datasets():
   return os.listdir('data')
+
+def corpusInfo(name):
+  (f, e, m) = loadFeatures(name)
+  avg_notes = 0
+  size = 0
+  version = m['version']
+  featureset = m['featureset']
+  for work in f:
+    size += len(f[work])
+    # This only works for corpora that have a const length feature
+    if 'const_length' in featureset:
+      index = featureset.index('const_length')
+      avg_notes += sum([x[index] for x in f[work]])
+    else:
+      # Some featureset strings miss the const_lenght index
+      # This code crashes if the featureset truly doesn't have this feature
+      avg_notes += sum([x[12] for x in f[work]])
+  avg_notes /= float(size)
+  return '[Works: {0}\tEntries: {1}\tFeatures version: {2}\tNotes per entry: {3}]'.format(len(f), size, version, avg_notes)
 
 # This doesn't handle polyfony very well but it is used for single voices only anyway
 def newParseScore(score):

### 2011-07-13 Paper/Paper.tex
Paper updated

@@
 \begin{Huge}
-Expressive Performance of Polyphonic Piano Music using Hierarchical Structure
+Performance Rendering using Structure Level Expression
 \end{Huge}
@@
 Both machine learning and rule based techniques have been extensively applied to performance rendering.
-However, relatively few systems make explicit use of musical structure, combined with machine learning.
-This paper introduces a performance rendering approach that tries to learn expression exclusively at a structural level.
-The approach attempts to to be complementary to approaches that learn expression at note level.
+However, relatively few systems make explicit use of machine learning combined with musical structure.
+Systems that use machine learning usually learn expression at note level.
+This paper introduces a performance rendering system that learns expression exclusively at a structural level.
+The system can be seen as complementary to systems that learn expression at note level.
+
 Musical structure is extracted automatically.
 A dataset containing performances aligned to scores is then used to learn how structure relates to expression.
-Finally, the model can be applied to a a new score to produce an expressive performance of the score.
+A hidden markov model is used to model this relation.
+Finally, the system can be applied to a a new score to produce an expressive performance.
 
@@
 A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure.
-The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed, is defined as follows\footnote{The version here is a slightly reformulated version of Van den Berg's deltarule}:
+The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed and $A$ is the resulting analysis, is defined as \footnote{The version here is a slightly reformulated, although functionally the same, version of Van den Berg's deltarule.
+See \cite{markwin} for his original version.}:
 \begin{algorithm}
@@
 
+The delta rule converts a sequence of notes into a nested list structure that can easily be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.
+
 It also possible to define higher order delta function (deltas of deltas).
@@
 For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$.
-Because we want to use a second order deltafunction for segmentation only, we choose to only use one interpretation: $[n_1, n_2], n_3]$.
-EXPLAIN WHY?
+Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[n_1, n_2], n_3]$.
+This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note.
+Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right, for our purposes it is important that at least one of these right interpretations is chosen.
+
+(The delta rule needs to be slightly modified to apply it to second order deltafunction, is this trivial?) ?
+
+Having defined first and second order deltafunctions and the deltarule, we can parse a piece of music into a delta tree or parse tree.
+Since we have a set of six basic note features we have a set of six interpretations(parse trees) available for a piece of music.
+Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent.
+Van den Berg captures this intuition in the form of a \textit{yield rule}.
+A node in the parse tree is said to `yield' a group of notes if the node recursively contains this group of notes.
+If two or more nodes in different trees yield the same group of notes they should be connected according to the yield rule.
+
+Unfortunately, the yield rule addresses the problem of how to interpret multiple parse trees quite poorly.
+We would like to have some way of combining parsetrees into one `consensus tree'.
+This is not what the yield rule does.
+The yield rule generates a set of trees in which some nodes may be interconnected but there does not seem to be a logical way to interpret this set of interpretations and connected nodes.
+Van den Berg does not further address this issue.
+In the next section we circumvent the problem completely by not using more than one interpretation at a time.
+
 \subsection{Segmentation}
 \label{segmentation}
+
+A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music.
+However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, derrived from deltatrees.
+
+The goal is to find a `safe' method to segment music, that is, a method that does not generate segmentations that clash too much with human musical intuition.
+For this purpose the delta trees of note onset and pitch seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions.
 
@@
 \begin{itemize}
-\item We are talking about SCOREFEATURES here, mention this
+\item We are talking about SCOREfeatures here, mention this ( has now been mentioned in approach)
 \end{itemize}
@@
 \begin{table}
+\caption{Context features}
+\label{features}
 \begin{itemize}
@@
 \end{itemize}
-\caption{Context features}
-\label{features}
 \end{table}
@@
 \begin{table}
+\caption{Constituent Features}
+\label{features}
 \begin{itemize}
@@
 \end{itemize}
-\caption{Constituent Features}
-\label{features}
 \end{table}

### 2011-07-13 Paper/Paper.tex
Updated paper, draft getting there but a lot still needs to be written

@@
 \usepackage{algorithmic}
+
+% Multiline tables
+\usepackage{array}
+\usepackage{booktabs}
 
@@
 
+NOTE(
 Some music transcription
@@
 
-Rencon
+Rencon)
 
@@
 The system simply misses one component of expression, namely structure level expression.
-(that Widmer had to artificially introduce using rules.
-) ?
+NOTE(that Widmer had to artificially introduce using rules. ) ?
 
@@
 The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''.
-We see this as a symptom of a note level expression based system.
+The only way to overcome this problem was to introduce methodologies that limited the expressiveness of performances.
+We see consider this trade-off to be inherent to note-level expression based systems.
+To solve it, some notion of structure level expression is required.
+
+NOTE(THE MOTIVATION SECTION SAYS MORE OR LESS THE SAME, MAKE IT COMPATIBLE)
 
@@
 
+NOTE
 \begin{itemize}
@@
 Having defined first and second order deltafunctions and the deltarule, we can parse a piece of music into a delta tree or parse tree.
-Since we have a set of six basic note features we have a set of six interpretations(parse trees) available for a piece of music.
+Since we have a set of six basic note features, we have a set of six interpretations(parse trees) available for a piece of music.
 Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent.
@@
 The goal is to find a `safe' method to segment music, that is, a method that does not generate segmentations that clash too much with human musical intuition.
-For this purpose the delta trees of note onset and pitch seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions.
+For this purpose the parse trees of delta onset (inter-onset interval, or simply IOI) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions.
+In figure \ref{pitchonsettrees} a pitch and onset parse tree of the first few bars of etude 23 of Chopin are displayed.
 
-\begin{itemize}
-\item We only need a segmentation for now
-\item Using the deltaframework to define segmentation
-\end{itemize}
+NOTE(Show pitch and onset trees here)
+
+To translate a delta tree into a segmentation we simply take all the nodes at a particular level and consider the notes it yields as one group.
+However, as can be seen in figure \ref{pitchonsettrees} the trees are quite deep and it is not clear what level we should choose.
+We can make the trees shallower by slightly modifying the delta rule.
+We introduce a small tolerance parameter $\epsilon$ and instead of looking for the maximum in of a list of deltas, we look for deltas within $m - \epsilon$ and $m + \epsilon$, where $m$ is the maximum delta within a list of deltas.
+
+Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes.
+These groupings are safe, but also quite large.
+Some music, like Bach's inventionen uses almost no differentiation in IOI.
+Hence, a segmentation based solely on inter onset interval will not serve our purposes.
+
+To overcome this problem it we will use pitch.
+While Bach's inventionen use almost no differentiation in IOI they can be segmented nonetheless.
+We will not make any attempt to combine pitch interval trees and IOI trees.
+Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.
+
+The sort of changes in pitch that indicate constituent transitions or not characterized very well by a first order pitch interval tree.
+For example a melody that has four notes that are chromatically ascending followed by four notes that have octave intervals between them should be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals.
+A first order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments.
+For this reason we use a second order pitch interval tree that does not suffer from this.
+
+NOTE (SECOND ORDER PITCH INTERVAL AND "SMOOTHED" IOI TREES FIGURES plus DISCUSSION)
+
+The pitch interval segmentations are less reliable than the IOI segmentations but since we use pitch only to subdivide IOI segments we can still be sure that some constituent transitions are based on the more reliable IOI segmentation.
 
@@
 \label{scorefeatures}
-
-\begin{itemize}
-\item We are talking about SCOREfeatures here, mention this ( has now been mentioned in approach)
-\end{itemize}
 
@@
 The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis.
-(Too literally?)
+NOTE(Too literally?)
 
-We can generalize pitch interval and duration ratio per note to constituent features: \textit{average pitch interval} and \textit{average duration ratio}.
+We can generalize pitch interval and duration ratio per note to constituent features: \textit{mean pitch interval} and \textit{mean duration ratio}.
 Definitions can be found in table \ref{features}.
@@
 
-The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in average pitch and average duration.
+The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in mean pitch and mean duration.
 At note level there is not much to say about the current note besides the pitch and duration.
 However at constituent we would also like to say something about the constituent itself.
-For this purpose \textit{average delta pitch} and \textit{average delta duration} are used.
+For this purpose \textit{mean delta pitch} and \textit{mean delta duration} are used.
 These features say something about the amount change in pitch and the amount of change in note duration.
 
-The complete set of score features consists of two \textit{context features} and two \textit{constituent features}.
+The complete set of score features consists of two \textit{context features} and two \textit{constituent features}:
+\begin{description}
+\item[\textbf{Mean pitch interval}] 
+The difference between the mean pitch of the current constituent and the mean pitch of the next constituent, zero if the current is the last constituent
+\item[\textbf{Mean duration ratio}]
+The logarithmic ratio between the mean note duration of the current constituent and the mean note duration of the next constituent
+\item[\textbf{Mean delta pitch}] The mean of all absolute pitch intervals within one constituent. 
+\item[\textbf{Mean delta duration}] The mean of all absolute differences in duration of succeeding notes within one constituent
+\end{description}
 
-\begin{table}
-\caption{Context features}
-\label{features}
-\begin{itemize}
-\item[\textbf{Average pitch interval}] 
-The difference between the average pitch of the current constituent and the average pitch of the next constituent, zero if the current is the last constituent
-\item[\textbf{Average duration ratio}]
-The logarithmic ratio between the average note duration of the current constituent and the average note duration of the next constituent
-\end{itemize}
-\end{table}
-
-\begin{table}
-\caption{Constituent Features}
-\label{features}
-\begin{itemize}
-\item[\textbf{Average delta pitch}] The average of all absolute pitch intervals within one constituent. 
-\item[\textbf{Average delta duration}] The average of all absolute differences in duration of succeeding notes within one constituent
-\end{itemize}
-\end{table}
-
-(Clearly, expressive markings in the score play a role in how the constituent should be played.
+NOTE(Clearly, expressive markings in the score play a role in how the constituent should be played.
 If the segmentation is specific enough  there will hopefully be at most expressive marking within each constituent.
@@
 
-\begin{itemize}
-\item Use average expression parameters per constituent
-\end{itemize}
+Every constituent will be assigned expression parameters that indicate how the constituent is played expressively in a performance.
+These parameters define what we mean by structure level expression and should therefore be chosen carefully.
+
+Some concepts that we think fall under structure level expression are \textit{crescendo} or \textit{decrescendo}, \textit{ritardando} and \textit{piano} or \textit{forte} etc.
+Concepts like crescendo, decrescendo and ritardando arguably fall under note level expression and for simplicity we will not consider them structure level expression.
+In fact, we will only look at the mean tempo, the mean articulation and the mean loudness of a constituent.
+
+YQX defines expression per note in three parameters: \textit{IOI ratio}, \textit{articulation} and \textit{loudness}.
+These parameters are defined as the logarithmic ratio between the performance IOI and the IOI notated in the score (calculated using some base tempo), the silence after a score note divided by the silence after the performed note and the logarithmic ratio between the loudness of the performed note and the mean loudness of the performance.
+
+We are going to define our own expression parameters in a similar way but let us first look at some issues with the definitions YQX uses.
+The definition of articulations seems to be awkward and inconsistend.
+Awkward because if a score note is not followed by a rest, the notated silence after it is zero, rendering articulation undefined.
+Inconsisted because all the other features use logarithmic ratios and we see no reason to not use a logarithmic ratio for articulation as well.
+
+IOI ratio and loudness is defined relative to mean performance tempo and loudness.
+However, to capture micro expression in the form of small changes in onset relative to the beat, it seems more logical to define this feature relative to the local tempo, instead of relative to the global tempo.
+The same argument can be made for defining loudness relative to local loudness instead of global loudness.
+Structure level expression may help to define the concepts of local tempo and local loudness as we can simply take the mean tempo and loudness within one constituent and take this to be the local tempo and loudness.
+
+The expression parameters that we will use are:
+\begin{description}
+\item[Mean Tempo Ratio] The logarithmic ratio between the mean tempo within the constituent and the base tempo of the performance.
+\item[Mean Articulation] The logarithmic ratio between performance IOI and the score IOI if the next note is not a rest.
+If the next note is a rest we use the note duration calculated with the score duration of the note and the local expressive tempo instead of the performance IOI.
+IOI is the onset of the next note minus the onset of the current note.
+\item[Mean Loudness Ratio] The logarithmic ratio of the mean loudness within the constituent and the base loudness of the performance
+\end{description}
+
+Admittedly, we still use mean loudness and tempo to calculate tempo and loudness ratio.
+This is a symptom of using a segmentation instead of hierarchical structure.
+When using hierarchical structure, these parameters could be defined relative to the parent constituent.
 
 \subsection{Model}
+
+We can now reduce a piece of music, represented as a sequence of notes $N_ij$, of which we have a score and a performance, to a sequence of score feature vectors $F$ and expression parameter vectors $E$.
+First we segment the score into constituents:
+
+\[\texttt{segment}(N_{ij}) = \{c_1, c_2, \cdots, c_n\}\]
+We can then extract feature and parameter values from the score and the performance:
+\begin{align*}
+f_i &= (p_i, d_i, \Delta p_i, \Delta d_i)^T\\
+F &= \{f_1, f_2, \cdots, f_n\}
+\end{align*}
+where $f_i$ is a feature vector, $p_i$ is the the mean pitch interval, $d_i$ is the mean duration ratio, $\Delta p_i$ is the mean delta pitch and $\Delta d_i$ is the mean delta duration.
+And the expression parameters: 
+\begin{align*}
+e_i &= (t_i, a_i, l_i)^T\\
+E &= \{e_1, e_2, \cdots, e_n\}
+\end{align*}
+where $e_i$ is a parameter vector, $t_i$ is the mean tempo ratio, $a_i$ is the mean articulation and $l_i$ is the mean loudness ratio.
+
+We can now train a hidden Markov model on this data. The
 
@@
 The same applies to the way YQX looks at loudness.
-This is specified as the logarithmic ratio between the notes loudness and the average loudness of the performance.
+This is specified as the logarithmic ratio between the notes loudness and the mean loudness of the performance.
 
@@
 \section{Conclusion}
-\cite{markwin}
+We have introduced structure level expression and argued that a good performance rendering systems needs some notion of structure level expression.
+
+We have shown that we can create sensible segmentations of music based on an application of the delta framework. \cite{markwin}
+
+We have suggested a possible intergration of SBPR with note level performance rendering
 \section{Discussion}
@@
 \item Combine deltatrees
-\item Use loudness and tempo direction instead of averages (requires more data)
+\item Use loudness and tempo direction instead of means (requires more data)
 \item Generalize approach to incorporate hierarchical structure, let structure extend to note level, so note level expression and structurelevel expression become integrated

### 2011-07-13 Code/database.py

@@ -224,9 +224,9 @@ if __name__ == "__main__":
           ['Play with internal sequencer', 'Play with audacious', 'View score', 'View midi info', 'View score info',\
           'Export deviation data to CSV', 'Performance', 'Extract melody','Cancel'])
       if choice == 0:
-        seq.play(NoteList(getScoreMidiPath1(selection)))
+        seq.play(NoteList(getPerformancePath1(selection)))
       elif choice == 1:
-        os.system("audacious {0}".format(getScorePath1(selection)))
+        os.system("audacious {0}".format(getPerformancePath1(selection)))
       elif choice == 2:
         getScore1(selection).show()
       elif choice == 3:

### 2011-07-13 Paper/Paper.tex
Updated paper, minor fix to code

@@
 
-NOTE(
-Some music transcription
-Performance rendering
+\begin{itemize}
+\item Introduce performance rendering
+\item Introduce YQX (What is it and why is it good)
 
-YQX is state of the art
-
-Rencon)
+\end{itemize}
 
@@
 
-The YQX system, as widmer admitted, tended to sometimes produce nervous sounding changes in expression.
+NOTE(Widmer didn't write the YQX paper, find out how to address the authors)
+
+The YQX system, as Widmer admitted, tended to sometimes produce nervous sounding changes in expression.
 He presents two extensions that seek to generate smoother performances.
-The problem with these extensions is, as widmer admits himself, that the increased smoothness comes at the expense of expressitivity.
+The problem with these extensions is, as Widmer admits himself, that the increased smoothness comes at the expense of expressitivity.
 To counter this, he adds three explicit rules that he uses to postprocess the performances.
@@
 
-(The delta rule needs to be slightly modified to apply it to second order deltafunction, is this trivial?) ?
+NOTE(The delta rule needs to be slightly modified to apply it to second order deltafunction, is this trivial?) ?
 
@@
 \subsection{Dealing with polyphony}
+\begin{itemize}
+\item Use only melody
+\end{itemize}
 \subsection{Discretization}
 
-Distinction of expressive tempo and attack and release.
+\begin{itemize}
+\item Normalize every feature
+\item Discretization: int($\phi * d$) where d is the discretization parameter (the higher, the smaller the bins)
+\end{itemize}
+
+Discretization of expression is a delicate subject, we want to capture small changes in dynamics and tempo precisely, but outliers may fall in large bins.
+A sigmoid function is very useful for this purpose.
+
+All expression parameters are ratios: their values vary around 1.0.
+To get a linear scale we use the logarithmic ratio.
+The tempo ratio is defined as: $\log(\frac{\texttt{performancetempo}}{\texttt{scoretempo}})$.
+To get the desired properties of discretization we use a sigmoid function.
+$D(p) = d * \frac{1}{1+e^{-p}}$ where $D(p)$ is the discretization of $p$, $e$ is the expression parameter value and $p$ is the discretization parameter.
 

### 2011-07-13 Code/performancerenderer.py

@@ -122,7 +122,7 @@ def trainHMM(hmm, features_set, expression_set, f_discretization=10, e_discretiz
         normalizations[i] = m
   hmm.normalizations = normalizations
   for work in features_set.keys():
-    if work == ignore:
+    if work[0] == ignore[0] and work[1] == ignore[1]:
       continue
     features = features_set[work]
     expression = expression_set[work]
@@ -217,7 +217,8 @@ def test(f_discretization=10, e_discretization=30, indep=False, selection=None, 
   else:
     hmm = HMM(2)
 
-  trainHMM(hmm, f, e, f_discretization, e_discretization, subset=subset, ignore=selection)
+  #trainHMM(hmm, f, e, f_discretization, e_discretization, subset=subset, ignore=selection)
+  trainHMM(hmm, f, e, f_discretization, e_discretization, subset=subset)
   hmm.storeInfo('hmm2.txt')
   print "Loading score"
   melodyscore = Score(score).melody()

### 2011-07-15 Code/hmm.py
Set unseens transition probabilities to zero. Fixed discretisation, should now be good

@@ -211,11 +211,11 @@ class HMM_indep(HMM):
   def __str__(self):
     return 'States:\n{0}\nNgrams:\n{1}\nLessergrams:\n{2}'.format(self.states, self.ngrams, self.lessergrams)
 
-  def emission_probability(self, state, observation):
+  def emission_probability(self, state, observation, smoothing=False):
     p = 1.0
     # This must be the absolute minimum probability
     if state == self.repeatstate:
-      return 1 / float(self.observation_count)
+      return 0.1 / float(self.observation_count)
     if state == self.startstate:
       if observation == self.startstate:
         return 1.0
@@ -227,11 +227,11 @@ class HMM_indep(HMM):
       else:
         return 0.0
     for i in range(len(state)):
-      p *= self.coincedences.get((i, observation, state[i]), 0.0) / float(self.observations.get(observation, 1))
+      p *= (self.coincedences.get((i, observation, state[i]), 0.0)) / float(self.observations.get(observation, 1))
       if p == 0.0: break
     return p
 
-  def transition_probability(self, lessergram, state, verbose=False):
+  def transition_probability(self, lessergram, state, smoothing=False, verbose=False):
     p = 1.0
     #lessergrams = [[v[j] for v in lessergram] for j in len(state)] 
     if state == self.endstate:
@@ -255,7 +255,7 @@ class HMM_indep(HMM):
 )
     # If this ngram was not found and the following state is the same as the last state (expression remains the same) do not return zero
     # These are the self-loops
-    if p == 0.0 and lessergram[len(lessergram)-1] == state:
+    if p == 0.0 and lessergram[len(lessergram)-1] == state and smoothing:
       minimum = 1.0
       for i in range(parameters):
         minimum *= 1 / float(self.lessergrams.get(tuple([i] + lg[i]), 1))

### 2011-07-15 Code/performancerenderer.py

@@ -31,10 +31,10 @@ def discretize(features, discretization=10, function='sigmoid', logarithmic=Fals
     features = [math.log(x) for x in features]
   if function == 'linear':
     for i in range(len(features)):
-      f.append(int(round(features[i] * discretization)))
+      f.append(int(round(features[i] * float(discretization))))
   elif function == 'sigmoid':
     for i in range(len(features)):
-      f.append(int(round(sigmoid(features[i], 5) * discretization)))
+      f.append(int(round(sigmoid(features[i], 5) * float(discretization))))
   return f
 
 def undiscretize(features, discretization=10, function='sigmoid', logarithmic=False):
@@ -64,11 +64,15 @@ def discretize_expression(parameters, discretization=10, subset=[1,2,3]):
 def undiscretize_expression(state, discretization=10):
   return tuple(undiscretize(state, discretization, function='sigmoid', logarithmic=True))
 
-def preprocess(features, normalizations, discretization=10, subset=None):
-  observations = []
+def normalize(features, normalizations):
   for i in range(len(features)):
     # Normalize
     features[i] = [features[i][j] / float(normalizations[j]) for j in range(len(features[i]))]
+  return features
+
+# Make relative features and select subset
+def preprocess(features, subset=None):
+  preprocessed = []
   for f0, f1 in zip(features[:-1], features[1:]):
     pitch0 = f0[sf.featureset.index('avg_pitch')]
     duration0 = f0[sf.featureset.index('avg_duration')]
@@ -79,7 +83,7 @@ def preprocess(features, normalizations, discretization=10, subset=None):
     dPitch1 = f1[sf.featureset.index('abs_dPitch')]
     dDuration1 = f1[sf.featureset.index('abs_dDuration')]
     # Difference in average pitch
-    pitch_interval = (pitch0 - pitch1) * 10 # 10 is a rather arbitrary number to make this feature visible in discretization
+    pitch_interval = (pitch0 - pitch1)# * 10 # 10 is a rather arbitrary number to make this feature visible in discretization
     # Relative difference in average note duration
     duration_ratio = math.log(duration0/float(duration1))
     # Do the next two features make any musical sense?
@@ -90,43 +94,49 @@ def preprocess(features, normalizations, discretization=10, subset=None):
     f0[sf.featureset.index('avg_duration')] = duration_ratio
     #f0[sf.featureset.index('abs_dPitch')] = dPitch_interval
[…]

### 2011-07-15 Code/tools.py

@@ -98,8 +98,10 @@ def corpusInfo(name):
   size = 0
   version = m['version']
   featureset = m['featureset']
+  unique = {}
   for work in f:
     size += len(f[work])
+    unique[(work[0], work[1])] = 0
     # This only works for corpora that have a const length feature
     if 'const_length' in featureset:
       index = featureset.index('const_length')
@@ -109,7 +111,7 @@ def corpusInfo(name):
       # This code crashes if the featureset truly doesn't have this feature
       avg_notes += sum([x[12] for x in f[work]])
   avg_notes /= float(size)
-  return '[Works: {0}\tEntries: {1}\tFeatures version: {2}\tNotes per entry: {3}]'.format(len(f), size, version, avg_notes)
+  return '[Works: {4}\tPerformances: {0}\tEntries: {1}\tFeatures version: {2}\tNotes per entry: {3}]'.format(len(f), size, version, avg_notes, len(unique))
 
 # This doesn't handle polyfony very well but it is used for single voices only anyway
 def newParseScore(score):

### 2011-07-15 Code/hmm.py
last commit before overhauling hmm.py

@@ -1,12 +1,14 @@
 # P(Expression|Features)
 # P(Ei|Ei-i, ..., Ei-n)
 # P(E0, E1, ..., En)
-
+import math, tools
 
 class HMM:
 
-  def __init__(self, order=2):
+  # Smoothing options: good-turing
+  def __init__(self, order=2, smoothing=None):
     self.order = order
+    self.smoothing = smoothing
 
     self.startstate = ('start',)
     self.endstate = ('end',)
@@ -23,6 +25,8 @@ class HMM:
     self.lessergram_count = 0
     self.starts = {}
     self.start_count = 0
+    # Used for good-turing smoothing:
+    self.nr = {}
     # This doesn't really belong here, it is just used as convenient storage
     self.normalizations = []
 
@@ -54,6 +58,30 @@ class HMM:
         self.ngrams[tuple(ngram)] = self.ngrams.get(tuple(ngram), 0) + 1
         self.lessergram_count += 1
         self.lessergrams[tuple(lessergram)] = self.lessergrams.get(tuple(lessergram), 0) + 1
+    if self.smoothing=='good-turing':
+      for obs in self.observations:
+        nr = {}
+        for state in self.states:
+          if (obs, state) in self.coincedences:
+            nr[self.coincedences[obs, state]] = nr.get(self.coincedences[obs, state], 0) + 1
+        x,y = [], []
+        for n,count in nr.iteritems():
+          x.append(math.log(n))
+          y.append(count)
+        print nr
+        # Find a least squares fit to of the frequency counts to nr = a + b*log(x)
+        (a,b) = y[0], 0
+        if len(x) > 1:
+          (a,b) = tools.linear_fit(x,y)
+        self.nr[obs] = (a,b)
+
+
+
+
+        
+
+      # Print a table of how much each number of states occurs
+
 
   def joint_probability(self, observation, state):
     if not (observation, state) in self.coincedences: return 0.0
@@ -231,7 +259,7 @@ class HMM_indep(HMM):
       if p == 0.0: break
     return p
 
-  def transition_probability(self, lessergram, state, smoothing=False, verbose=False):
+  def transition_probability(self, lessergram, state, smoothing=None, verbose=False):
     p = 1.0
     #lessergrams = [[v[j] for v in lessergram] for j in len(state)] 
     if state == self.endstate:

### 2011-07-15 Code/performancefeatures.py

@@ -235,3 +235,15 @@ def expressionOLD(alignment):
     lastdeviation = deviation
   return expression
 
+if __name__ == '__main__':
+  x = [math.log(1), math.log(3), math.log(4), math.log(5)]
+  y = [1, 2, 2.1, 2.2]
+  (a, b) = linear_fit(x, y)
+  xx = [i for i in range(1,10)]
+  fx = [a+b*math.log(i) for i in xx]
+
+  import matplotlib.pyplot as plt
+  plt.plot(xx, fx)
+  plt.show()
+
+

### 2011-07-15 Code/performancerenderer.py

@@ -203,7 +203,7 @@ def analyseCorpus(discret):
       print '{0}-{1}: {2}'.format(work[1], work[2], p)
   
 
-def test(f_discretization=10, e_discretization=30, indep=False, selection=None, subset=None, corpus=None):
+def test(f_discretization=10, e_discretization=30, indep=False, selection=None, subset=None, corpus=None, smoothing=None):
   if not selection:
     selection = (db.select())
   score = db.getScore1(selection)
@@ -224,9 +224,9 @@ def test(f_discretization=10, e_discretization=30, indep=False, selection=None, 
   print ']'
   print '\tScorefeatures discretization: {0}\n\tExpression discretization: {1}\n'.format(f_discretization, e_discretization)
   if indep:
-    hmm = HMM_indep(2)
+    hmm = HMM_indep(2, smoothing)
   else:
-    hmm = HMM(2)
+    hmm = HMM(2, smoothing)
 
   trainHMM(hmm, f, e, f_discretization, e_discretization, subset=subset, ignore=selection)
   #trainHMM(hmm, f, e, f_discretization, e_discretization, subset=subset)
@@ -301,6 +301,7 @@ if __name__ == '__main__':
   f_discretization = 10
   e_discretization = 10
   corpus = None
+  smoothing = None
   if len(a) > 1:
     if a[1] == 'load':
       loadperformance()
@@ -398,6 +399,13 @@ if __name__ == '__main__':
         except IndexError, ValueError:
           print 'Error parsing command line arguments'
           exit(0)
+      elif a[i] == '-smoothing':
+        try:
+          smoothing = a[i+1]
+          i += 1
+        except IndexError, ValueError:
+          print 'Error parsing command line arguments'
+          exit(0)
       elif a[i] == '-corpus':
         try:
           corpus = a[i+1]
@@ -410,7 +418,7 @@ if __name__ == '__main__':
       i += 1
 
 
-  test(f_discretization, e_discretization, indep, selection=selection, subset=subset, corpus=corpus)
+  test(f_discretization, e_discretization, indep, selection=selection, subset=subset, corpus=corpus, smoothing=smoothing)
 
   
 

### 2011-07-15 Code/tools.py

@@ -160,3 +160,24 @@ def recursive_print(l):
     else:
       return l
 
+# Least squares fit
+def linear_fit(X, Y):
+  n = len(X)
+  if len(X) != len(Y):
+    print "linear fit: WARNING: lengths of data don't match: {0} and {1}".format(len(X), len(Y))
+  sum_x=0.0
+  sum_y=0.0
+  sum_xx=0.0
+  sum_xy=0.0
+  for x,y in zip(X,Y):
+    sum_x=sum_x+x
+    sum_y=sum_y+y
+    xx=pow(x,2)
+    sum_xx=sum_xx+xx
+    xy=x*y
+    sum_xy=sum_xy+xy
+
+  #Calculating the coefficients
+  a=(-sum_x*sum_xy+sum_xx*sum_y)/float(n*sum_xx-sum_x*sum_x)
+  b=(-sum_x*sum_y+n*sum_xy)/float(n*sum_xx-sum_x*sum_x)
+  return a, b

### 2011-07-15 Code/hmm.py
HMM has is now a proper model, reversed the conditional probabilities and added good turing smoothing for emission probabilities. Result: beautiful expressive performances! Still the performance renderer has some problems rendering mozart on chopin corpus and what's very weird is that sometimes the generated deviations results in errors because tempos are not found. This happens for Srjkjabin

@@ -9,6 +9,7 @@ class HMM:
   def __init__(self, order=2, smoothing=None):
     self.order = order
     self.smoothing = smoothing
+    print self.smoothing
 
     self.startstate = ('start',)
     self.endstate = ('end',)
@@ -16,17 +17,18 @@ class HMM:
 
     self.coincedences = {}
     self.observations = {}
-    self.states = []
+    self.states = {}
 
     self.ngrams = {}
     self.observation_count = 0
+    self.state_count = 0
     self.ngram_count = 0
     self.lessergrams = {}
     self.lessergram_count = 0
     self.starts = {}
     self.start_count = 0
     # Used for good-turing smoothing:
-    self.nr = {}
+    self.nr = None
     # This doesn't really belong here, it is just used as convenient storage
     self.normalizations = []
 
@@ -38,8 +40,10 @@ class HMM:
       self.coincedences[observation, state] = self.coincedences.get((observation, state), 0) + 1
       self.observations[observation] = self.observations.get((observation), 0) + 1
       self.observation_count += 1
-      if not state in self.states and not (state == self.startstate or state == self.endstate):
-        self.states.append(state)
+      # We allow start and end state to be in the statecounts
+      if not state in self.states:
+        self.states[state] = self.states.get(state, 0) + 1
+        self.state_count += 1
     for i in range(len(states)-self.order + 1):
       ngram = []
       lessergram = []
@@ -58,30 +62,25 @@ class HMM:
         self.ngrams[tuple(ngram)] = self.ngrams.get(tuple(ngram), 0) + 1
         self.lessergram_count += 1
         self.lessergrams[tuple(lessergram)] = self.lessergrams.get(tuple(lessergram), 0) + 1
-    if self.smoothing=='good-turing':
+
+  # Initialize Good-Turing smoothing
+  def init_GT_smoothing(self):
+    self.nr = {}
+    for state in self.states:
+      nr = {}
       for obs in self.observations:
-        nr = {}
-        for state in self.states:
-          if (obs, state) in self.coincedences:
-            nr[self.coincedences[obs, state]] = nr.get(self.coincedences[obs, state], 0) + 1
-        x,y = [], []
-        for n,count in nr.iteritems():
-          x.append(math.log(n))
-          y.append(count)
-        print nr
-        # Find a least squares fit to of the frequency counts to nr = a + b*log(x)
-        (a,b) = y[0], 0
[…]

### 2011-07-15 Code/performancerenderer.py

@@ -20,7 +20,7 @@ def inv_sigmoid(y, p=1.0):
 # Given normalized features use some function to discretize.
 # Possibilities: linear or sigmoid
 # Use a logistic function to get a larger resolution for small differences than for larger ones
-def discretize(features, discretization=10, function='sigmoid', logarithmic=False):
+def discretize(features, discretization=10, function='sigmoid', logarithmic=False, sensitivity=5):
   f = []
   # WORKAROUND
   if logarithmic:
@@ -34,10 +34,10 @@ def discretize(features, discretization=10, function='sigmoid', logarithmic=Fals
       f.append(int(round(features[i] * float(discretization))))
   elif function == 'sigmoid':
     for i in range(len(features)):
-      f.append(int(round(sigmoid(features[i], 5) * float(discretization))))
+      f.append(int(round(sigmoid(features[i], sensitivity) * float(discretization))))
   return f
 
-def undiscretize(features, discretization=10, function='sigmoid', logarithmic=False):
+def undiscretize(features, discretization=10, function='sigmoid', logarithmic=False, sensitivity=5.0):
   features = list(features)
   discretization = float(discretization)
   f = []
@@ -51,18 +51,18 @@ def undiscretize(features, discretization=10, function='sigmoid', logarithmic=Fa
         features[i] = 1        
       if features[i] == discretization:
         features[i] = 9        
-      f.append(inv_sigmoid(features[i] / discretization, 5))
+      f.append(inv_sigmoid(features[i] / discretization, sensitivity))
   if logarithmic:
     f = [math.exp(x) for x in f]
   return f
 
-def discretize_expression(parameters, discretization=10, subset=[1,2,3]):
+def discretize_expression(parameters, discretization=10, subset=[1,2,3], sensitivity=5):
   if subset != []:
     parameters = [parameters[i] for i in subset]
-  return tuple(discretize(parameters, discretization, function='sigmoid', logarithmic=True))
+  return tuple(discretize(parameters, discretization, function='sigmoid', logarithmic=True, sensitivity=sensitivity))
 
-def undiscretize_expression(state, discretization=10):
-  return tuple(undiscretize(state, discretization, function='sigmoid', logarithmic=True))
+def undiscretize_expression(state, discretization=10, sensitivity=5):
[…]

### 2011-07-17 Paper/Paper.tex
last commit before overhauling hmm.py
Fixed HMM. Smoothing now seems to work, results not too impressive. Independent still doesn't seem to work. Updated Paper

@@
 
-Musical structure is extracted automatically.
+%Musical structure is extracted automatically.
 A dataset containing performances aligned to scores is then used to learn how structure relates to expression.
@@
 
-We can now reduce a piece of music, represented as a sequence of notes $N_ij$, of which we have a score and a performance, to a sequence of score feature vectors $F$ and expression parameter vectors $E$.
-First we segment the score into constituents:
+We can now reduce a piece of music, represented as a sequence of notes $N_{ij}$, of which we have a score and a performance, to a sequence of score feature vectors $F$ and expression parameter vectors $E$.
+First, we segment the score into constituents:
 
@@
 
-We can now train a hidden Markov model on this data. The
+If we parse a number of pieces is this manner we can use it to do some statistics.
+We can for example calculate the features likely-hood which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$.
+
+\begin{equation}
+\label{emission}
+P(f_i|e_i) = \frac{c(f_i, e_i)}{c(e_i)}
+\end{equation}
+
+Where $c(x)$ is the number of occurences of $x$ in the corpus.
+The expression transition probability is defined by
+\begin{equation}
+\label{transition}
+P(e_i|e_{i-1}) = \frac{c(e_{i-1}, e_i)}{c(e_{i-1})}
+\end{equation}
+where we make the simplification that expression in one constituent is only dependend on the expression in the previous constituent which is of course not true.
+
+The problem of creating a performance of a score is now reduced to finding a suitable sequence of expression vectors given a sequence of feature vectors.
+We can calculate the probability of one expression vector of a performance as the product the probabilities defined in equations \ref{emission} and \ref{transition}.
+The probability of the entire performance is approximated by the product of the individual expression vector probabilities.
+
+\begin{equation}
+P(E|F) = \displaystyle\prod_{i=1}^{n}P(f_i|e_i)P(e_i|e_{i-1})
+\end{equation}
+
+The resulting model is analogous to a stochastic part of speech tagger based on a hidden Markov model where we have substituted words for score feature vectors and parts of speech for expression parameter vectors.
+Rendering a performance is like finding the most likely part of speech tags for a sequence of words.
+In our case:
+
+\[E^* = \texttt{argmax}_E P(E|F)\]
+
+which can be found using Viterbi decoding.
 
 \section{Implementation}
-\section{Performance Rendering}
+
 \begin{itemize}
@@
 \end{itemize}
-\subsection{Dataset and representation}
-\begin{itemize}
-\item attack release
-\item tempo curves
-\end{itemize}
+\subsection{Corpus and Representation}
+We were lucky to find and receive permission to use the CrestMusePEDB \ref{crestmuse} corpus, which is a corpus that contains expressive performances of Western classical music by famous pianists.
+The corpus contains about
+
+Every performance is accompanied by an XML file containing information on how every note from the score is performed.
+This information consists of a loudness deviation, attack deviation and release deviation.
+The loudness is defined relative to a base loudness.
+The attack and release deviation are defined as the portion of a local beat duration that the attack and release deviates from the score.
+
+Local tempo is defined for every beat in every measure as the ratio of the tempo in that beat and the base tempo.
+
+To prepare a score and performance from the corpus for use in our system we use the score to extract melody notes.
+We do this simply by taking notes in the highest voice in the top staff.
+From now on when we talk about pieces and notes we mean melodies and melody notes.
+
+Attack and release are clearly note level parameters so we do not use them.
+We only need tempo deviations and loudness deviations.
+The average tempo of a constituent is determined by the average tempo deviation of all the beats that fall within the constituent.
+The average loudness is determined by the average loudness deviation of every note within the constituent.
+
+The deltafunctions that are used in the segmentation process use pitch intervals and duration ratios.
+The pitch intervals use MIDI note numbers which range from 21 to 108 and go up one semi-tone with each step.
+Durations are in milliseconds, calculated from the score and a standard tempo, set at 120 beats per minute.
+
 \subsection{Dealing with polyphony}
@@
 $D(p) = d * \frac{1}{1+e^{-p}}$ where $D(p)$ is the discretization of $p$, $e$ is the expression parameter value and $p$ is the discretization parameter.
+\subsection{Performance rendering}
+\begin{itemize}
+\item Extract melody
+\item Extract scorefeatures from melody
+\item lookup notes in deviation, ignore missing notes
+\item Extract expressive and non-expressive melody and extract expression features
+\item Perform notes other than melody notes the same as the nearest melody note
+\end{itemize}
 
@@
 \item Targets for expression level expression
+\item Expressive markings
+\item Staccato/articulation statistics
 \end{itemize}
@@
 \subsection{Repetition}
-A notion of repetition and similarity would certainly improve the system.
+The system would certainly benefit from a notion of repetition and similarity.
 Repetition is a very good indicator of constituent breaks.
@@
 
+\section{Acknowledgements}
+
  \bibliographystyle{plain} % plain, nature, jbact

### 2011-07-17 Code/alignment.py

@@ -169,6 +169,9 @@ class Alignment:
 
               on = performance.microseconds_to_ticks(on_ms)
               off = performance.microseconds_to_ticks(off_ms)
+              if off < on:
+                print "WARNING: release before onset! Check your deviations or performancerenderer"
+                off = on
               
               pnote = Note(on, off, n.pitch.midi, int(onvel), int(offvel), 
                   annotation=(score.index(part), part.index(measure), measure.index(voice), voice.index(note), note.id))

### 2011-07-17 Code/database.py

@@ -35,6 +35,8 @@ def getPerformance1(t):  return getPerformance(t[0], t[1], t[2], t[3])
 def getScore1(t):
   if t == 'm21':
     return selectM21Score()
+  elif t == 'sd':
+    return selectScore()
   return getScore(t[0], t[1], t[2], t[3])
 def getDeviation1(t):  return getDeviation(t[0], t[1], t[2], t[3])
 
@@ -196,17 +198,44 @@ def selectM21Score():
     path = util.menu('Choose a work', m21.corpus.getComposer(composers[comp]), cancel=True)
   return m21.corpus.parse(m21.corpus.getComposer(composers[comp])[path])
 
+def selectScore():
+  scores = os.listdir('scores')
+  path = 'scores/{0}'.format(scores[util.menu('Select score', scores)])
+  return m21.converter.parse(path)
+
 def select():
   work = -1
   while(work == -1):
-    comp = util.menu('Choose a composer', getComposers() + ['MUSIC21 Corpus'])
+    comp = util.menu('Choose a composer', getComposers() + ['MUSIC21 Corpus'] + ['Score Directory'])
     print len(getComposers()), comp
     if comp == len(getComposers()):
       return 'm21'
+    elif comp == len(getComposers())+1:
+      return 'sd'
     work = util.menu('Choose a work and performer', getWorks(getComposers()[comp]), cancel=True)
 
   selected = getWorks(getComposers()[comp])[work]
   return getComposers()[comp], selected[0], selected[1], selected[2]
+
+def selectWork():
+  works = {}
+  allworks = []
+  for i in range(len(getComposers())):
+    allworks += [(getWorks(getComposers()[i])[j], i, j) for j in range(len(getWorks(getComposers()[i])))]
+  for work, i, j in allworks:
+    s = getWorks(getComposers()[i])[j]
+    works[getComposers()[i], s[0]] = (getComposers()[i], s[0], s[1], s[2])
+
+  sortedkeys = sorted(works.keys())
+
+  choice = util.menu('Choose a score', sortedkeys + ['Music21 Corpus'] + ['Score Directory'])
+  if choice == len(sortedkeys):
+    return 'm21'
+  elif choice == len(sortedkeys)+1:
+    return 'sd'
+  return works[sortedkeys[choice]]
+
+
 
 def byIndexes(i, j):
   selected = getWorks(getComposers()[i])[j]

### 2011-07-17 Code/hmm.py

@@ -7,6 +7,8 @@ class HMM:
 
   # Smoothing options: good-turing
   def __init__(self, order=2, smoothing=None):
+    # Order is a relic from when I had the illusion of making this class generic
+    # Now it's just bigrams
     self.order = order
     self.smoothing = smoothing
     print self.smoothing
@@ -19,18 +21,21 @@ class HMM:
     self.observations = {}
     self.states = {}
 
-    self.ngrams = {}
+    self.bigrams = {}
     self.observation_count = 0
-    self.state_count = 0
-    self.ngram_count = 0
-    self.lessergrams = {}
-    self.lessergram_count = 0
+    self.bigram_count = 0
+    self.unigrams = {}
+    self.unigram_count = 0
     self.starts = {}
     self.start_count = 0
     # Used for good-turing smoothing:
     self.nr = None
+    self.unseen_bigrams = 0
+    self.unseen_observations = 0
     # This doesn't really belong here, it is just used as convenient storage
     self.normalizations = []
+
+    self.state_counts = {}
 
   def learn(self, observations, states):
     parameters = len(states[0])
@@ -41,64 +46,66 @@ class HMM:
       self.observations[observation] = self.observations.get((observation), 0) + 1
       self.observation_count += 1
       # We allow start and end state to be in the statecounts
-      if not state in self.states:
-        self.states[state] = self.states.get(state, 0) + 1
-        self.state_count += 1
+      self.states[state] = self.states.get(state, 0) + 1
     for i in range(len(states)-self.order + 1):
-      ngram = []
-      lessergram = []
+      bigram = []
+      unigram = []
       for offset in range(self.order):
         if states[i+offset] == self.startstate or states[i+offset] == self.endstate:
-          ngram.append(states[i+offset])
+          bigram.append(states[i+offset])
         else:
-          ngram.append(states[i+offset])
+          bigram.append(states[i+offset])
       for offset in range(self.order-1):
         if states[i+offset] == self.startstate or states[i+offset] == self.endstate:
-          lessergram.append(states[i+offset])
+          unigram.append(states[i+offset])
         else:
-          lessergram.append(states[i+offset])
-        self.ngram_count += 1
-        # Add the index of this parameter to the ngram
-        self.ngrams[tuple(ngram)] = self.ngrams.get(tuple(ngram), 0) + 1
[…]

### 2011-07-17 Code/performancerenderer.py

@@ -34,7 +34,7 @@ def discretize(features, discretization=10, function='sigmoid', logarithmic=Fals
       f.append(int(round(features[i] * float(discretization))))
   elif function == 'sigmoid':
     for i in range(len(features)):
-      f.append(int(round(sigmoid(features[i], sensitivity) * float(discretization))))
+      f.append(int((sigmoid(features[i], sensitivity) * float(discretization))))
   return f
 
 def undiscretize(features, discretization=10, function='sigmoid', logarithmic=False, sensitivity=5.0):
@@ -156,6 +156,7 @@ def render(score, segmentation, hmm, f_discretization=10, e_discretization=30, s
   # Find the best expressive explanation for these features
   print "Finding best fitting expression"
   (p, states) = hmm.viterbi(observations)
+  print p, states
   expression = []
   for state in states:
     expression.append(undiscretize_expression(state, e_discretization, sensitivity=sensitivity))
@@ -202,12 +203,71 @@ def analyseCorpus(discret):
       print '{0}-{1}: {2}'.format(work[1], work[2], [s[i] for i in subset])
       print '{0}-{1}: {2}'.format(work[1], work[2], p)
 
-def playPerformance(performance, expression, name='ExpressivePerformance'):
+      
+
+def loadperformance():
+  (selection, expression) = tools.loadPerformance()
+  score = db.getScore1(selection)
+  melodyscore = Score(score).melody()
+  melody = tools.parseScore(melodyscore)
+  segmentation = structure.reasonableSegmentation(melody)
+  performance = perform.perform(score, Score(score).melody(), segmentation, expression, converter=melody)
+  seq = Sequencer()
+  seq.play(performance)
+
+def visualize(segmentation, expression, visualize=None, store=None, title=None):
+  notes = []
+  onsets = []
+  values = []
+  if not visualize:
+    visualize = selectSubset(['Dynamics', 'Articulation', 'Tempo'])
+  for segment, expr in zip(segmentation, expression):
+    for note in segment:
+      onsets.append(note.on)
+      values.append([expr[i] for i in visualize])
+  import matplotlib.pyplot as plt
+  fig = plt.figure()
+  plt.plot(onsets, values)
+  plt.ylabel('Deviation')
+  plt.xlabel('Score time')
+  if title:
+    plt.title(title)
+  #dplot = fig.add_subplot(111)
[…]

### 2011-07-18 Code/hmm.py
Safety commit before once again overhauling hmm.py

@@ -31,7 +31,7 @@ class HMM:
     # Used for good-turing smoothing:
     self.nr = None
     self.unseen_bigrams = 0
-    self.unseen_observations = 0
+    self.unseen_coincedences = 0
     # This doesn't really belong here, it is just used as convenient storage
     self.normalizations = []
 
@@ -71,19 +71,22 @@ class HMM:
     self.nr = {}
     for state in self.states:
       nr = {}
-      unseen = 0
+      unseen = 0 #self.unseen_coincedences
+      sum = 0
       for obs in self.observations:
         if (obs, state) in self.coincedences:
           nr[self.coincedences[obs, state]] = nr.get(self.coincedences[obs, state], 0) + 1
-        else:
-          unseen += 1
+        else: unseen += 1
       # Things that never occur
+      # Variable should be set when decoding is called
       x,y = [1], [unseen]
+      #x, y = [], []
       for n,count in nr.iteritems():
         x.append(math.log(n+1))
         y.append(count)
-      x.append(math.log(100000))
-      y.append(0)
+      #x.append(math.log(max(nr.keys())+100))
+      #y.append(1)
+      #nr[0] = self.unseen_coincedences * len(self.states)
       nr[0] = unseen
       #print nr
       # Find a least squares fit to of the frequency counts to nr = a + b*log(x)
@@ -92,32 +95,77 @@ class HMM:
       (a,b) = tools.linear_fit(x,y)
       self.nr[state] = (a,b, nr)
 
-  def smoothed_frequency_count(self, state, r):
-    if not self.nr:
-      self.init_GT_smoothing()
+  def testModel(self):
+    print "Sum of smoothed probabilities of all coincedences starting with:"
+    for state in self.states:
+      p = sum([self.emission_probability(state, obs) for obs in self.observations])
+      #if p > 2 or p < 0:
+      print 'Emission\t{0}: {1}'.format(state, p)
+      #p = sum([self.transition_probability([state], state1) for state1 in self.states])
+      #print 'Transition\t{0}: {1}'.format(state, p)
+    system.exit(0)
+
+  # Turing estimate
+  def smoothed_frequency_count(self, state, c):
     (a,b,nr) = self.nr[state]
     # If we cannot find this frequency we back off to a logarithmic fit
     # But perhaps interpolation  is better?
-    if r in nr:
-      smoothed = nr[r]
+    if c in nr:
[…]

### 2011-07-18 Code/performancerenderer.py

@@ -156,7 +156,6 @@ def render(score, segmentation, hmm, f_discretization=10, e_discretization=30, s
   # Find the best expressive explanation for these features
   print "Finding best fitting expression"
   (p, states) = hmm.viterbi(observations)
-  print p, states
   expression = []
   for state in states:
     expression.append(undiscretize_expression(state, e_discretization, sensitivity=sensitivity))

### 2011-07-18 Paper/Paper.tex
Small paper update. Hmm working better, performance rendering broken. Probabilities sum more to one

@@
 We were lucky to find and receive permission to use the CrestMusePEDB \ref{crestmuse} corpus, which is a corpus that contains expressive performances of Western classical music by famous pianists.
-The corpus contains about
+The music includes works by Bach, Beethoven and Chopin.
 
@@
 \subsection{Discretization}
+To calculate probabilities, we must discretize the feature and parameter vectors.
 
-\begin{itemize}
-\item Normalize every feature
-\item Discretization: int($\phi * d$) where d is the discretization parameter (the higher, the smaller the bins)
-\end{itemize}
+\paragraph*{Expression} Discretization of expression parameters is a delicate subject, we want to capture small changes in dynamics and tempo precisely, but outliers may fall in large bins.
+A sigmoid function is very useful for this purpose.
+The expression parameters are all logarithmic ratios so they theoretically vary from $-\infty$ to $\infty$.
+Therefore we first normalize the parameters by dividing through the mizimum absolute value found in the corpus.
+Discretization of a normalized expression parameter $p$ into $d$ bins is now done as follows:
 
-Discretization of expression is a delicate subject, we want to capture small changes in dynamics and tempo precisely, but outliers may fall in large bins.
-A sigmoid function is very useful for this purpose.
+\begin{equation}
+D(p) = \texttt{floor}\left(\frac{d}{1+e^{-sp}}\right)
+\end{equation}
 
-All expression parameters are ratios: their values vary around 1.0.
-To get a linear scale we use the logarithmic ratio.
-The tempo ratio is defined as: $\log(\frac{\texttt{performancetempo}}{\texttt{scoretempo}})$.
-To get the desired properties of discretization we use a sigmoid function.
-$D(p) = d * \frac{1}{1+e^{-p}}$ where $D(p)$ is the discretization of $p$, $e$ is the expression parameter value and $p$ is the discretization parameter.
+Where $D(p)$ is the discretization of $p$, $s$ is a special sensitivity parameter indicating how small the changes in tempo that the discretization captures can be; a larger $s$ means a more sensitive discretization.
+Undiscretization is the reverse operation:
+
+\begin{equation}
+D^{-1}(p) = s^{-1} d^{-1}(-\log(p^{-1}) - 1)
+\end{equation}
+
+\paragraph*{Features}
+
+To discretize the features, we simply normalize every feature dividing through the maximum absolute value of that feature found in the corpus.
+After normalization we multiply the feature by a discretization parameter that determines the number of bins and take the floor.
+
+\begin{equation}
+D(f) = \texttt{floor}(f*d)
+\end{equation}
+
+Where $f$ is the normalized feature value.
+
 \subsection{Performance rendering}

### 2011-07-18 Code/hmm.py

@@ -31,6 +31,7 @@ class HMM:
     # Used for good-turing smoothing:
     self.nr = None
     self.unseen_bigrams = 0
+    self.N = 0
     self.unseen_coincedences = 0
     # This doesn't really belong here, it is just used as convenient storage
     self.normalizations = []
@@ -67,12 +68,23 @@ class HMM:
         self.unigrams[tuple(unigram)] = self.unigrams.get(tuple(unigram), 0) + 1
 
   # Initialize Good-Turing smoothing
+  def new(self):
+    self.nr = {}
+    unseen = 0 #self.unseen_coincedences
+    for state in self.states:
+      for obs in self.observations:
+        if (obs, state) in self.coincedences:
+          self.nr[self.coincedences[obs, state]] = self.nr.get(self.coincedences[obs, state], 0) + 1
+        else: unseen += 1
+    self.nr[0] = unseen + len(self.states) * self.unseen_coincedences
+    self.N = sum([self.smoothed_frequency_count(None, i) for i in range(0, max(self.nr.keys())+1)])
+
+    
   def init_GT_smoothing(self):
     self.nr = {}
     for state in self.states:
       nr = {}
-      unseen = 0 #self.unseen_coincedences
-      sum = 0
+      unseen = self.unseen_coincedences
       for obs in self.observations:
         if (obs, state) in self.coincedences:
           nr[self.coincedences[obs, state]] = nr.get(self.coincedences[obs, state], 0) + 1
@@ -83,17 +95,16 @@ class HMM:
       #x, y = [], []
       for n,count in nr.iteritems():
         x.append(math.log(n+1))
-        y.append(count)
+        y.append(math.log(count))
       #x.append(math.log(max(nr.keys())+100))
       #y.append(1)
       #nr[0] = self.unseen_coincedences * len(self.states)
-      nr[0] = unseen
-      #print nr
+      print state, nr
       # Find a least squares fit to of the frequency counts to nr = a + b*log(x)
       #(a,b) = y[0], 0
       #if len(x) > 1:
       (a,b) = tools.linear_fit(x,y)
-      self.nr[state] = (a,b, nr)
+      self.nr[state] = (a,b,unseen)
 
   def testModel(self):
     print "Sum of smoothed probabilities of all coincedences starting with:"
@@ -103,55 +114,31 @@ class HMM:
[…]

### 2011-07-18 Code/structure.py

@@ -321,6 +321,7 @@ if __name__ == '__main__':
   level = int(raw_input(""))
   groups = segmentation(trees[tree], level)
   groups = subsegmentation(groups, second_order_tree, pitch, 1)
+  groups = reasonableSegmentation(melody)
   structured_notes = []
   loud = False
   for group in groups:

### 2011-07-20 Paper/Paper.tex
Updated paper. Final tweaks to the system

@@
 \usepackage{amsmath, amsthm, amssymb}
+%\usepackage{fullpage}
 
@@
 \usepackage{qtree}
+
+% Images
+\usepackage{graphicx}
 
@@
 
-NOTE(Show pitch and onset trees here)
+\begin{figure}
 
-To translate a delta tree into a segmentation we simply take all the nodes at a particular level and consider the notes it yields as one group.
+\end{figure}
+NOTE(we now use only onset on the mozart corpus, explain adaptive segmentation)
+
+In the tree in figure \ref{pitchonsettrees} we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes.
+To translate the tree into a segmentation we do the following for for every node in the tree starting with the root node:
+\begin{algorithm}
+\label{alg:segmentation}
+\begin{enumerate}
+\item If the node is a note, at it as a singleton constituent to the segmentation
+\item Expand the node.
+\item If the node contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
+\item If the node contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
+\end{enumerate}
+\end{algorithm}
+
+The value of $x$ represents the tolerance of singleton constituents.
+We need this parameter because sometimes very long notes cause top level splits in the onset tree.
+We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart.
+
+%To translate a delta tree into a segmentation we simply take all the nodes at a particular level and consider the notes it yields as one group.
 However, as can be seen in figure \ref{pitchonsettrees} the trees are quite deep and it is not clear what level we should choose.
@@
 
-Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes.
-These groupings are safe, but also quite large.
-Some music, like Bach's inventionen uses almost no differentiation in IOI.
-Hence, a segmentation based solely on inter onset interval will not serve our purposes.
+%Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes.
 
-To overcome this problem it we will use pitch.
+For some music like Mozart, using only a delta onset tree works very well.
+However, some music, like Bach's inventionen uses almost no differentiation in IOI.
+Hence, a segmentation based solely on inter onset interval will not serve our purposes for this music.
+
+To overcome this problem we will use pitch.
 While Bach's inventionen use almost no differentiation in IOI they can be segmented nonetheless.
@@
 For this reason we use a second order pitch interval tree that does not suffer from this.
+
+\begin{figure}
+\includegraphics[scale=3]{img/melodysegments}
+\caption{First four segments of piano sonata KV331 III by Mozart (grace notes are not shown)}
+\end{figure}
 
@@
 
-\subsection{Dealing with polyphony}
-\begin{itemize}
-\item Use only melody
-\end{itemize}
 \subsection{Discretization}
@@
 To discretize the features, we simply normalize every feature dividing through the maximum absolute value of that feature found in the corpus.
-After normalization we multiply the feature by a discretization parameter that determines the number of bins and take the floor.
+After normalization we multiply the feature by a discretization parameter $d$ that determines the number of bins and take the floor.
 
@@
 
+\subsection{Smoothing}
+
+Despite having discretized our feature and observation vectors we often find that we observe feature vectors in a new score that we had never seen during training.
+We do not want the conditional probabilities from equation \ref{emission} to become zero so we have to smooth these probabilities.
+We use a smoothing technique known as simple Good-Turing smoothing.
+The idea is that we use the probabilities of things we have seen once for the things we have never seen.
+Recall how we calculate the conditional probability of a feature vector:
+\[
+P(f_i|e_i) = \frac{c(f_i, e_i)}{c(e_i)}
+\]
+
+Let us call the $c(f_i, e_i)$ the coincedence count.
+Let $N_c$ be the number of things with frequency $c$ in the corpus.
+For example if there are five coincedences $(f_x, e_x)$ nd no other pair of $f$ and $e$ occurs five times, then $N_{5} = 20$.
+Good-Turing re\"estimates counts according to this formula:
+
+\begin{equation}
+\label{count}
+c^* = (c+1) \frac{N_{c+1}}{N_c}
+\end{equation}
+
+The smoothed probability of some event $x$ is
+
+\[P(x) = \frac{c^*}{N}\]
+
+In our case, the probability we seek is $P(f_i|e_i)$, the sample size $N$ is therefore the number of times we have seen $e_i$: $c(e_i)$.
+$N_c$ is the number of coincedences with $e_i$ that we have seen $c$ times.
+The count $c$ is $c(f_i, e_i)$ and $c^*$ is derived using equation \ref{count}.
+The
+
+\[P(f_i|e_i) = \frac{c^*}{c(e_i)} \mbox{ for all } c > 0 \]
+
+By applying this formula to all counts larger than zero we reserve approximately $\frac{N_1}{N}$ probability mass for things that we have never seen.
+We assign an equal probability for all unseen feature vectors, namely:
+\[
+P(f_{\texttt{unseen}}|e_i) = \frac{1}{U} \frac{N_1}{c(e_i)} 
+\]
+Where U is the number of unseen coincedences, which is determined by the number of different observations in the corpus plus the new observations from the score minus the number of coincedences with $e_i$.
+
+In order to make this work we cannot use $N_c$ directly since it will not be defined for every $c$. We use a least square apprixmation using the following function.
+
+\[\log(N_c) = a + b \log(c)\]
+
+This completes the definition of simple Good-Turing smoothing.
+However, even now there may still be some expression parameter vectors that occur with unique feature vectors, so for that expression only $N_1$ is defined.
+We cannot fit a function to one sample, so if we have only one $Nc$ sample, we simply turn off Good-Turing smoothing and accept the fact that some probabilities will be zero.
+
 \subsection{Performance rendering}
+
+Our model is able to generate expressively performed melodies but does not handle polyphony.
+During training, the bass and harmony notes were stripped off.
+After rendering an expressive performance we can simply put them back in and estimate their expressive parameters.
+We do this by this by by giving each  bass and harmony note the expressive parameters of the last played melody note.
+
 \begin{itemize}
@@
 \section{Evaluation and Results}
-The proposed system actually requires more training data to do meaningful statistics. However, since expressive interpretation can be 
+
+Using the method that we now have extensively described we have tried to generate a few performances of piano works by Mozart.
+The system was trained on 42 performances of 13 works by Mozart, see appendix \ref{appendix:corpus} for a more details.
+The choice for Mozart was motivated by the fact that our segmentation works best when phrase transitions are marked by long notes or silences, which is often the case in the work of Mozart.
+Another reason for choosing Mozart in favor of for example Chopin is that we think structure level expression is more outspoken in performances of Mozart's music.
+
+%The system was tested on works from Mozart as we do not expect our system to generalize well to other composers.  We used
+
+We discretized the score features using a discretization parameter of five.
+This means the constituent features (see section \ref{sec:features}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well.
+The expression features where discretized into ten bins, with a sensitivity of five.
+These settings made sure we recognised most of the score features of a new score.
+The low number of bins does give the performances bit of a cartoonistic quality, but that does not make them any less interesting.
+
 \subsection{Subjective Listening}
@@
 \section{Conclusion}
+In this thesis we have critisized performance rendering systems that only use note level expression.
+
 We have introduced structure level expression and argued that a good performance rendering systems needs some notion of structure level expression.
@@
 \item Staccato/articulation statistics
+\item Backoff smoothing (doesn't assign equal probabilities to unseen events)
 \end{itemize}
@@
 
+\appendix
+\section{Corpus}
+\label{appendix:corpus}
+\begin{tabular}{ll}
+\textbf{Work} & \textbf{Performer}\\
+\hline
+Sonata KV331 \MakeUppercase{\romannumeral 1}. & Hiroko Nakamura\\
+& Glenn Gould\\
+& Christoph Eschenbach\\
+& Ingrid Haebler\\
+& Lili Kraus\\
+& Maria Jo\~ao Pires\\
+& Alicia De Larrocha\\
+& kn?\\
+& yi?\\
+& tn?\\
+& Norio Shimizu\\
+& mo\\
+& mk?\\
+& ea?\\
+& nm?\\
+& kt?\\
+& tm?\\
+\hline
+Sonata KV331 \MakeUppercase{\romannumeral 2}. & Hiroko Nakamura\\
+& Maria Jo\~ao Pires\\
+& Alicia De Larrocha\\
+& ea?\\
+\hline
+Sonata KV331 \MakeUppercase{\romannumeral 3}. & Hiroko Nakamura\\
+& Maria Jo\~ao Pires\\
+& Christoph Eschenbach\\
+\hline
+Sonata KV545 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
+& Glenn Gould\\
+\hline
+Sonata KV545 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
+& Glenn Gould\\
+\hline
+Sonata KV545 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
+& Glenn Gould\\
+\hline
+Sonata KV279 \MakeUppercase{\romannumeral 1}. & Glenn Gould\\
+& Maria Jo\~ao Pires\\
+\hline
+Sonata KV279 \MakeUppercase{\romannumeral 2}. & Glenn Gould\\
+& Maria Jo\~ao Pires\\
+\hline
+Sonata KV279 \MakeUppercase{\romannumeral 3}. & Glenn Gould\\
+& Maria Jo\~ao Pires\\
+\hline
+Sonata KV310 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
+& Glenn Gould\\
+& Hiroko Nakamura\\
+Sonata KV310 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
+Sonata KV310 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
+Sonata KV570 \MakeUppercase{\romannumeral 3}. & nm?\\
+\end{tabular}
+
 \end{document}

### 2011-07-20 Code/hmm.py

@@ -82,6 +82,7 @@ class HMM:
     
   def init_GT_smoothing(self):
     self.nr = {}
+    undef = 0
     for state in self.states:
       nr = {}
       unseen = self.unseen_coincedences
@@ -89,43 +90,71 @@ class HMM:
         if (obs, state) in self.coincedences:
           nr[self.coincedences[obs, state]] = nr.get(self.coincedences[obs, state], 0) + 1
         else: unseen += 1
+
+      # We can't use these
+      if len(nr) == 1:
+        self.nr[state] = 'undefined'
+        undef += 1
+        continue
       # Things that never occur
       # Variable should be set when decoding is called
-      x,y = [1], [unseen]
-      #x, y = [], []
+      x, y = [], []
       for n,count in nr.iteritems():
-        x.append(math.log(n+1))
+        x.append(math.log(n))
         y.append(math.log(count))
+      if len(x) < 2:
+        x.append(math.log(n+1))
+        y.append(math.log(1))
       #x.append(math.log(max(nr.keys())+100))
       #y.append(1)
       #nr[0] = self.unseen_coincedences * len(self.states)
-      print state, nr
+      #print state, x, y, nr
       # Find a least squares fit to of the frequency counts to nr = a + b*log(x)
       #(a,b) = y[0], 0
       #if len(x) > 1:
       (a,b) = tools.linear_fit(x,y)
+      #if len(nr) == 1:
+    #    print '{0} unseen: {1} Nr: {2}'.format(state, unseen, nr)
+      if len(nr) > 5:
+        import matplotlib.pyplot as plt
+        print [nr[i] for i in nr.keys()]
+        print [math.exp(a+b*math.log(i+1)) for i in nr.keys()]
+        #plt.plot([1] + [i+1 for i in sorted(nr.keys())], [math.exp(i) for i in y])
+        #plt.show()
+        #raw_input("Press enter to continue...")
+        #plt.plot(range(1, 1000), [math.exp(a+b*math.log(i)) for i in range(1, 1000)])
+        #raw_input("Press enter to continue...")
       self.nr[state] = (a,b,unseen)
+    print "Smoothing disabled on {0} out of {1} states".format(undef, len(self.states))
 
   def testModel(self):
     print "Sum of smoothed probabilities of all coincedences starting with:"
+    #for state in self.states:
+    #  print self.transition_probability([self.startstate], state),
+    #  print self.emission_probability(state, self.observations.values()[2])
     for state in self.states:
[…]

### 2011-07-20 Code/performancerenderer.py

@@ -25,6 +25,9 @@ def discretize(features, discretization=10, function='sigmoid', logarithmic=Fals
   # WORKAROUND
   if logarithmic:
     for i in range(len(features)):
+      if features[i] == 0:
+        print "WARNING zero expressionfeature"
+        features[i] = 1
       if features[i] < 0: 
         features[i] = -features[i]
         print "Warning: invalid featurevalue found, {0} in {1}".format(features[i], features)
@@ -73,6 +76,8 @@ def normalize(features, normalizations):
 # Make relative features and select subset
 def preprocess(features, subset=None):
   preprocessed = []
+  pitch_interval = 0
+  duration_ratio = 0
   for f0, f1 in zip(features[:-1], features[1:]):
     pitch0 = f0[sf.featureset.index('avg_pitch')]
     duration0 = f0[sf.featureset.index('avg_duration')]
@@ -100,8 +105,10 @@ def preprocess(features, subset=None):
       preprocessed.append(f0)
   # The last constituent:
   f = list(features[-1])
-  f[sf.featureset.index('avg_pitch')] = pitch_interval
-  f[sf.featureset.index('avg_duration')] = duration_ratio
+  if len(features) == 1:
+    print "Warning, nonsegmented work found"
+  f[sf.featureset.index('avg_pitch')] = 0
+  f[sf.featureset.index('avg_duration')] = 0
   if subset:
     preprocessed.append([f[i] for i in subset])
   else:
@@ -139,7 +146,7 @@ def trainHMM(hmm, features_set, expression_set, f_discretization=10, e_discretiz
     obs = [tuple(discretize(f, f_discretization, 'linear')) for f in features]
     states = []
     # Select the subset of features that will be used and discretise
-    for f, p in zip(features, expression):
+    for f, p in zip(features_set[work], expression):
       exp = discretize_expression(p, e_discretization, sensitivity=sensitivity) 
       #states.append(tuple(list(exp) + list(featureset)))
       states.append(exp)
@@ -156,6 +163,7 @@ def render(score, segmentation, hmm, f_discretization=10, e_discretization=30, s
   # Find the best expressive explanation for these features
   print "Finding best fitting expression"
   (p, states) = hmm.viterbi(observations)
+  print p, states
   expression = []
   for state in states:
[…]

### 2011-07-20 Code/structure.py

@@ -282,6 +282,38 @@ def reasonableSegmentation(notes):
   # First get level one segments from a first order onset tree then subdivide the segments using a second order pitch tree on level 1
   return subsegmentation(segmentation(first_order_tree(onset, notes), 1), second_order_tree, pitch, 1)
 
+def adaptiveSubsegmentation(segments, treefunction, feature, tolerance=0, threshold=8):
+  newsegmentation = []
+  for segment in segments:
+    if len(segment) <= threshold:
+      newsegmentation.append(segment)
+      continue
+    newsegmentation += segmentation(treefunction(feature, segment, tolerance), 1)
+  return newsegmentation
+
+def adaptiveSegmentation(tree):
+  if isinstance(tree, Leaf):
+    return [[tree.contents]]
+  terminal = False
+  nodes = len(leaf_nodes(tree))
+  singletons = 0
+  for node in tree:
+    if isinstance(node, Leaf):
+      singletons += 1
+      if singletons > int(nodes) / 16:
+        terminal = True
+        break
+  if terminal:
+    return [[leaf.contents for leaf in leaf_nodes(tree)]]
+  result = []
+  for node in tree:
+    result += adaptiveSegmentation(node)
+  return result
+
+def newSegmentation(notes):
+  return adaptiveSegmentation(list_to_tree(first_order_tree(onset, notes)))
+
+
 if __name__ == '__main__':
   import sys
   if len(sys.argv) > 1:
@@ -296,7 +328,8 @@ if __name__ == '__main__':
   w = db.select()
   score = Score(db.getScore1(w))
   melodyscore = score.melody()
-  melody = tools.parseScore(melodyscore)
+  #melodyscore.show()
+  melody = tools.parseScore(melodyscore, range(1, 9))
   trees = [second_order_tree(onset, melody, 0.5), second_order_tree(pitch, melody, 0.0, ), first_order_tree(onset, melody, 0.0), first_order_tree(pitch, melody)]
 
   for tree in trees:
@@ -319,9 +352,11 @@ if __name__ == '__main__':
   tree = int(raw_input(''))
   print "Choose level"
   level = int(raw_input(""))
-  groups = segmentation(trees[tree], level)
-  groups = subsegmentation(groups, second_order_tree, pitch, 1)
-  groups = reasonableSegmentation(melody)
+#  groups = segmentation(trees[tree], level)
+#  groups = subsegmentation(groups, second_order_tree, pitch, 1)
[…]

### 2011-07-20 Code/tools.py

@@ -92,6 +92,30 @@ def chooseFeatures():
 def datasets():
   return os.listdir('data')
 
+
+def extendedCorpusInfo(name):
+  (f, e, m) = loadFeatures(name)
+  version = m['version']
+  featureset = m['featureset']
+  unique = {}
+  print '============================================'
+  for work in f:
+    unique[(work[0], work[1])] = unique.get((work[0], work[1]), []) + [work[2]]
+  for work in unique:
+    print '{0}: {1}'.format(work, unique[work])
+  print '============================================'
+  for work in f:
+    features = f[work]
+    notes = 0
+    if 'const_length' in featureset:
+      index = featureset.index('const_length')
+      notes += sum([x[index] for x in features])
+    else:
+      # Some featureset strings miss the const_lenght index
+      # This code crashes if the featureset truly doesn't have this feature
+      notes += sum([x[12] for x in features])
+    print '{0}: notes: {1} const: {2}'.format(work, notes, len(features))
+
 def corpusInfo(name):
   (f, e, m) = loadFeatures(name)
   avg_notes = 0
@@ -121,7 +145,7 @@ def newParseScore(score):
   midifile.close()
   return NoteList('/tmp/tmp.mid')
   
-def parseScore(score):
+def parseScore(score, measures=None):
   output = NoteList() 
   defaultvelocity = 63
   for part in score.parts:
@@ -132,6 +156,9 @@ def parseScore(score):
       performancetime = measure.offset
       if not isinstance(measure, m21.stream.Measure):
         continue
+      if measures:
+        if not measure.number in measures:
+          continue
       for voice in measure.voices:
         for note in voice:
           notes = []
@@ -153,7 +180,7 @@ def parseScore(score):
 
 def recursive_print(l):
   if hasattr(l,'__iter__'):
-    return '({0})'.format(','.join([recursive_print(x) for x in l]))
+    return '[ {0}]'.format(' '.join([recursive_print(x) for x in l]))
   else:
     if hasattr(l,'name'):
       return l.name()

### 2011-07-20 Code/train.py

@@ -33,7 +33,7 @@ def train(trainset):
     deviations = db.getDeviation1(query)
     alignment = Alignment(score, deviations)
     melody = alignment.melody()
-    segments = structure.reasonableSegmentation(tools.parseScore(melody))
+    segments = structure.newSegmentation(tools.parseScore(melody))
     const += len(segments)
     lengths = sum([len(s) for s in segments])
     m = max([len(s) for s in segments])

### 2011-07-20 Paper/Paper.tex
Updated paper, generated performances

@@
 
+%\usepackage{fullpage}
+
 % Trees
 \usepackage{qtree}
+
+\usepackage{subfigure}
 
@@
 Since support for automatic expression is mostly absent or very poor in music editing software, some genres of music have evolved to sound acceptable even without expression.
-If automatic expression better in this software, computer music production could greatly widen its field of application.
+If good working automatic expression was available in software, computer music production could greatly widen its field of application.
+
+A performance rendering system is a computer program that automatically creates an expressive, humanlike performance of a musical score.
+Early attempts like Director Musices REF have tried to create use sets of rules to create performances.
+These systems
+Expressively performing a musical score is known as performance rendering. A performance rendering system aims to create automate A yearly competition is held among performance rendering systems
 
@@
 For this purpose the parse trees of delta onset (inter-onset interval, or simply IOI) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions.
-In figure \ref{pitchonsettrees} a pitch and onset parse tree of the first few bars of etude 23 of Chopin are displayed.
+These two trees are displayed in figure \ref{fig:pitchonsettrees}.
 
 \begin{figure}
+\centering
+\subfigure[Delta onset tree]{
+\includegraphics[scale=0.4]{img/onset_sym}
+\label{fig:donset}
+}
+\subfigure[Second order delta pitch tree]{
+\includegraphics[scale=0.3]{img/second_order_pitch_sym}
+\label{fig:dpitch}
+}
+\caption{Parse trees of the first eight bars of piano sonata KV331 III. (Turkisch March)by Mozart }
+\label{fig:pitchonsettrees}
+\end{figure}
 
-\end{figure}
-NOTE(we now use only onset on the mozart corpus, explain adaptive segmentation)
-
-In the tree in figure \ref{pitchonsettrees} we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes.
-To translate the tree into a segmentation we do the following for for every node in the tree starting with the root node:
+In the trees in figure \ref{fig:pitchonsettrees} we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes.
+To translate the tree into a segmentation we use algorithm \ref{alg:segmentation} starting with the root node.
 \begin{algorithm}
+\caption{Segmentation}
 \label{alg:segmentation}
@@
 We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart.
+Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if you use the delta tree from figure  \ref{fig:donset} you will not get this segmentation.
+This segmentation was based on a onset delta tree over the entire piece instead of just the first eight bars}
 
@@
 \caption{First four segments of piano sonata KV331 III by Mozart (grace notes are not shown)}
+\label{fig:mozartsegments}
 \end{figure}
-
-NOTE (SECOND ORDER PITCH INTERVAL AND "SMOOTHED" IOI TREES FIGURES plus DISCUSSION)
 
@@
 \subsection{Discretization}
-To calculate probabilities, we must discretize the feature and parameter vectors.
+
+We chose to make the model discrete. We discretize expression and scorefeatures in a different way.
 
@@
 
-\begin{itemize}
-\item Extract melody
-\item Extract scorefeatures from melody
-\item lookup notes in deviation, ignore missing notes
-\item Extract expressive and non-expressive melody and extract expression features
-\item Perform notes other than melody notes the same as the nearest melody note
-\end{itemize}
+%\begin{itemize}
+%\item Extract melody
+%\item Extract scorefeatures from melody
+%\item lookup notes in deviation, ignore missing notes
+%\item Extract expressive and non-expressive melody and extract expression features
+%\item Perform notes other than melody notes the same as the nearest melody note
+%\end{itemize}
 
@@
 The low number of bins does give the performances bit of a cartoonistic quality, but that does not make them any less interesting.
+
+The segmentation was based only delta onset trees since this worked best for Mozart.
+
+Since transition probabilities are not smoothed only transitions that actually occur in the corpus have a nonzero probability
 
@@
 \item Generalize approach to incorporate hierarchical structure, let structure extend to note level, so note level expression and structurelevel expression become integrated
+\item Backoff smoothing
+\item Staccato statistics
 \end{itemize}

### 2011-07-20 Code/performancefeatures.py

@@ -127,8 +127,8 @@ def vanDerWeijExpression(alignment, segments):
       (start_tempo, tempo_direction) = (tempos[0], 0)
     else:
       note_onsets = [n.on for n in segment]
-      (start_loudness, loudness_direction) = linear_fit(note_onsets, loudnesses)
-      (start_tempo, tempo_direction) = linear_fit(range(len(tempos)), tempos)
+      (start_loudness, loudness_direction) = (0, 0) #linear_fit(note_onsets, loudnesses)
+      (start_tempo, tempo_direction) = (0, 0) #linear_fit(range(len(tempos)), tempos)
 
     expression.append((average_ioi_ratio, average_relative_loudness, average_articulation, average_tempo, start_tempo, start_loudness, tempo_direction, loudness_direction))
 

### 2011-07-20 Code/performancerenderer.py

@@ -222,7 +222,7 @@ def loadperformance():
   seq = Sequencer()
   seq.play(performance)
 
-def visualize(segmentation, expression, visualize=None, store=None, title=None):
+def visualize(segmentation, expression, visualize=None, store=None, title=None, legend=False):
   notes = []
   onsets = []
   values = []
@@ -236,12 +236,13 @@ def visualize(segmentation, expression, visualize=None, store=None, title=None):
       onsets.append(converter.ticks_to_milliseconds(note.on)/1000.0)
       values.append([expr[i] for i in visualize])
   import matplotlib.pyplot as plt
-  fig = plt.figure(1)
+  fig = plt.figure(figsize=(12, 4))
   for i in visualize:
     plt.plot(onsets, [v[i] for v in values], label=param[i])
   plt.ylabel('Deviation')
   plt.xlabel('Score time (seconds)')
-  plt.legend(bbox_to_anchor=(0., 1), loc=2, borderaxespad=0.)
+  if legend:
+    plt.legend(bbox_to_anchor=(0., 1), loc=2, borderaxespad=0.)
 
   if title:
     plt.title(title)

### 2011-07-21 Paper/Paper.tex
Draft almost finished

@@
 
-In Western art music tradition it is customary that composers write down their compositions in scores.
-The task of a performer is to some extend to accurately reproduce the score, however, a perfect reproduction of a score generally sounds robotic and unpleasant.
+In the Western art music tradition it is customary that composers write down their compositions as scores.
+The task of a performer is to some extent to accurately reproduce the score; however, a perfect reproduction of a score generally sounds robotic and unpleasant.
 What makes a performance appealing is that the performer deviates from the score, altering timing, articulation and loudness, creating an expressive performance.
 
-Given a hypothetical perfect synthesizer, performing music with computers is a trivial task.
+Given a hypothetical perfect synthesizer, performing scores with computers is a trivial task.
 However \textit{expressively} performing music is not and much research has focussed on this issue.
-Computers are widely used in music production.
+Such research is interesting from a scientific point of view as well as a practical point of view.
+Artificial intelligence, cognitive science and musicology may all benefit from a better understanding of expression in music.
+
+%Such systems Computers are widely used in music production.
 Since support for automatic expression is mostly absent or very poor in music editing software, some genres of music have evolved to sound acceptable even without expression.
@@
 
+A rather practical approach to expressive music performance is to recreate it.
 A performance rendering system is a computer program that automatically creates an expressive, humanlike performance of a musical score.
-Early attempts like Director Musices REF have tried to create use sets of rules to create performances.
-These systems
-Expressively performing a musical score is known as performance rendering. A performance rendering system aims to create automate A yearly competition is held among performance rendering systems
+Performance rendering systems have always faced a problem of evaluation.
+It is debatable how much we should value correlation of generated performances with human performances of the same piece.
+A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both can be seen valid interpretations.
+This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience.
 
-\begin{itemize}
-\item Introduce performance rendering
-\item Introduce YQX (What is it and why is it good)
+Over the years, many performance rendering systems have been proposed.
+Early attempts, like Director Musices \cite{friberg2000generating}, have applied sets of rules to the problem.
+As the availablity of corpora grew, machine learning systems have been proposed as well.
+Gerhard Widmer is working with a large dataset containing almost all works by Chopin, performed by one pianist.
+A performance rendering system using this dataset called YQX \cite{yqx} was proposed by Sebastian Flossman et al.
+in 2008.
+This system won in al three categories of Rencon2008.
+At Rencon2011 it was ranked best in a blind evaluation by experts and ranked second in the evaluation by an audience.
+It should be noted that despite its age, Director Musices ranked third in the blind evaluation, suggesting that although performance rendering systems have advanced, the field is still in its infancy.
 
-\end{itemize}
+Many of the performance rendering systems that use machine learning, including YQX, predict expression at note level.
+Some do use structure.
+Pop-E\cite{hashida2006pop}, a rule based performance rendering system, uses a structure analysis based on Lerdahls and Jackendoff's grouping preference rules, but the structural analysis must be supplied to it by a human expert.
+In this thesis we will motivate the need for structure level expression and propose a system that purely uses structure level expression.
+The YQX system is very successful at the moment, therefore we will use this system as a reference for introducing our own system.
+In the following section we will provide a short analysis of some shortcomings of the YQX system and argue what we believe is the reason for these shortcomings.
+
+Section \ref{sec:approach} will describe the system as a whole.
+What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}.
+Section \ref{sec:method} will describe every component of the system in detail.
 
@@
 However, very few approaches make extensive use of structure.
-Some machine learning systems like the \textit{Music Interpretation System} make use of some aspects of GTTM and Widmer uses musical \textit{closure} [Narmour, introduce this properly] as a feature in his succesful \textit{YQX} system.
+Some machine learning systems like the \textit{Music Interpretation System} make use of some aspects of GTTM and Widmer uses musical \textit{closure} [Narmour, introduce this properly] as a feature in his successful \textit{YQX} system.
 This means that only the level of structure defined by closure is used by this system.
@@
 
+\section{Musical Structure}
+\label{sec:structure}
 \subsection{Motivation}
 
-NOTE(Widmer didn't write the YQX paper, find out how to address the authors)
-
-The YQX system, as Widmer admitted, tended to sometimes produce nervous sounding changes in expression.
+The YQX system, as the authors admit, tended to sometimes produce nervous sounding changes in expression.
 He presents two extensions that seek to generate smoother performances.
@@
 The system simply misses one component of expression, namely structure level expression.
-NOTE(that Widmer had to artificially introduce using rules. ) ?
 
 The system presented here will use structure exclusively to generate performances and will completely ignore note level expression.
-Section \ref{approach} will describe the system as a whole.
-What we mean by structure level expression is clarified in section \ref{structure}.
-The approach for extracting structure is described in section \ref{deltaframework}.
-
-The expressiveness of the performances it generates is therefore limited.
-In section XX we will describe how this system could be integrated with a note level performance rendering system to hopefully produce some variant of YQX that doesn't need explicit rules to produce expressive performances.
+We hope that this system is able capture the sudden changes of expression that a note level system cannot easily learn.
+Sudden changes in expression occur frequently in expressive performances, a few notes may be played very loud, or very slow followed by soft notes played very fast.
+When done at the right moment, such sudden changes can greatly improve the humanlike feel of the performance, however when done at the wrong moment, they can ruin a performance.
+We think that at structure level it will be easier to capture the context in which sudden changes occur.
 
@@
 
-\subsection{Musical structure}
+\subsection{Structure Level Expression}
 When listening to music, the listener's musical intuition assigns a certain hierarchical structure to the music: Notes make up phrases, phrases make up themes and themes make up a piece.
@@
 We could represent musical structure as a parse tree.
-This paradigm corresponds to the intuition that a melody is not simply a sequence of notes but that notes form phrases and we can look at phrases at different levels.
+This paradigm corresponds to the intuition that a melody is not simply a sequence of notes but that notes form phrases.
+A phrase is always a constituent but a constituent is not always a phrase, it can be a distinguished part of a phrase or a distinguished set of phrases as well.
 
-It must be noted that the resulting parse tree can be highly ambiguous and even experienced listeners may not agree on the correct parsetree of a piece.
+We must note that musical scores can be highly ambiguous and even experienced listeners may not agree on the correct parsetree of a piece.
 Quite often there may simply be more than one parse tree that makes musical sense.
@@
 
-Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, expression is still only predicted per note.
+Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, expression is only predicted per note.
 The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''.
-The only way to overcome this problem was to introduce methodologies that limited the expressiveness of performances.
-We see consider this trade-off to be inherent to note-level expression based systems.
+The only way to overcome this problem was to introduce methods that limited the expressiveness of performances.
+We consider this trade-off to be inherent to note-level expression based systems.
 To solve it, some notion of structure level expression is required.
 
-NOTE(THE MOTIVATION SECTION SAYS MORE OR LESS THE SAME, MAKE IT COMPATIBLE)
-
 \section{Approach}
+\label{sec:approach}
 
@@
 Instead we will try to predict only constituent level expression.
-The assumption is that this kind of expression really exists in performances and that is different and independend from note level structure.
+The assumption is that this kind of expression really exists in performances and that is different and independent from note level structure.
 We think that a constituent level system also corresponds better to how actual human performers play music.
@@
 The performances are of high quality and played by famous pianists.
-See section \ref{corpus} for more details on the corpus.
+See section \ref{sec:corpus} for more details on the corpus.
 
-A structural analysis is used to derrive hierarchical structure of every score in the corpus, however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents.
+A structural analysis is used to derive a hierarchical structure for every score in the corpus, however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents.
 After segmentation, four score features, two of which are direct generalizations of YQX's score features, are extracted for each constituent.
@@
 
-The simplification of ignoring note level expression is of course unjustified and severly limits the expressiveness of the performances that can be generated.
-However, if succesful, the resulting performances will clearly demonstrate the phenomenon of structure level expression.
-A succesful performance rendering system should incorporate both structure and note level expression, section \ref{integration} will explore two possibilities for an integration of SBPR and note level performance rendering.
-
 The succes of a SBPR system depends largely on two factors.
@@
 The following section address these issues.
-
-NOTE
-\begin{itemize}
-\item Mention GTTM
-\item Give short overview of the system?
-\end{itemize}
 
@@
 \section{Method}
+\label{sec:method}
 
-This section will describe individual components of the system sketched in section \ref{approach} in more detail.
-The structural analysis of is based on the delta framework, which will be described in section \ref{deltaframework}.
-Section \ref{segmentation} will discuss how the delta framework is applied to get a segmentation.
-Sections \ref{features} and \ref{targets} will describe how the scorefeatures and expression parameters are calculated.
-Section \ref{model} will describe how a hidden Markov model is trained on our data.
+This section will describe individual components of the system sketched in section \ref{sec:approach} in more detail.
+The structural analysis is based on the delta framework, which will be described in section \ref{sec:deltaframework}.
+Section \ref{sec:segmentation} will discuss how the delta framework is applied to get a segmentation.
+Sections \ref{sec:scorefeatures} and \ref{sec:targets} will describe how the scorefeatures and expression parameters are calculated.
+Section \ref{sec:model} will describe how a hidden Markov model is trained on our data.
 
 \subsection{The Delta Framework}
+\label{sec:deltaframework}
 
-In his Phd thesis \ref{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework.
-He relates his work to the work of Lerdahl and Jackendoff \ref{lerdahl1996generative} but claims to have found a more general approach.
-Below, I will shortly describe the deltaframework as proposed by Van der Berg.
+In his Phd thesis \cite{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework.
+He relates his work to the work of Lerdahl and Jackendoff \cite{lerdahl1996generative} but claims to have found a more general approach.
+Below, I will shortly describe the delta framework as proposed by Van der Berg.
 
@@
 A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure.
-The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed and $A$ is the resulting analysis, is defined as \footnote{The version here is a slightly reformulated, although functionally the same, version of Van den Berg's deltarule.
+The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed and $A$ is the resulting analysis, is defined as \footnote{The version here is a slightly reformulated, although functionally eqivalent, version of Van den Berg's deltarule.
 See \cite{markwin} for his original version.}:
@@
 This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note.
-Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right, for our purposes it is important that at least one of these right interpretations is chosen.
-
-NOTE(The delta rule needs to be slightly modified to apply it to second order deltafunction, is this trivial?) ?
+Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.
 
@@
 \subsection{Segmentation}
-\label{segmentation}
+\label{sec:segmentation}
 
@@
 \subsection{Constituent Features}
-\label{scorefeatures}
+\label{sec:scorefeatures}
 
@@
 The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis.
-NOTE(Too literally?)
 
-We can generalize pitch interval and duration ratio per note to constituent features: \textit{mean pitch interval} and \textit{mean duration ratio}.
-Definitions can be found in table \ref{features}.
+We can generalize pitch interval and duration ratio per note to \textit{context features}: \textbf{mean pitch interval} and \textbf{mean duration ratio}. Definitions can be found below. 
+\begin{description}
+\item[\textbf{Mean pitch interval}] 
+The difference between the mean pitch of the current constituent and the mean pitch of the next constituent, zero if the current is the last constituent
+\item[\textbf{Mean duration ratio}]
+The logarithmic ratio between the mean note duration of the current constituent and the mean note duration of the next constituent
+\end{description}
+
 Since I-R arch is related to note-level expression it does not generalize well to a constituent level feature.
@@
 However at constituent we would also like to say something about the constituent itself.
-For this purpose \textit{mean delta pitch} and \textit{mean delta duration} are used.
-These features say something about the amount change in pitch and the amount of change in note duration.
-
-The complete set of score features consists of two \textit{context features} and two \textit{constituent features}:
+For this purpose the \textit{constituent features} \textbf{mean delta pitch} and \textbf{mean delta duration} are used.
+These features say something about the amount change in pitch and the amount of change in note duration:
 \begin{description}
-\item[\textbf{Mean pitch interval}] 
-The difference between the mean pitch of the current constituent and the mean pitch of the next constituent, zero if the current is the last constituent
-\item[\textbf{Mean duration ratio}]
-The logarithmic ratio between the mean note duration of the current constituent and the mean note duration of the next constituent
 \item[\textbf{Mean delta pitch}] The mean of all absolute pitch intervals within one constituent. 
@@
 
-NOTE(Clearly, expressive markings in the score play a role in how the constituent should be played.
-If the segmentation is specific enough  there will hopefully be at most expressive marking within each constituent.
-)?
+Note the difference between these features and the context features.
+The context features use the mean pitch of one constituent and the the mean duration within constituent and compare these to the mean pitch and mean duration of the next constituent.
+The constituent features use the average of all pitch intervals within one constituent and the average of all differences in duration within one constituent, giving a measure of the spread of pitch and differentation of rhythm within the constituent.
+
+The complete set of score features consists of the two context features and the two constituent features:
 
 \subsection{Expression Parameters}
-\label{targets}
+\label{sec:targets}
 
@@
 \subsection{Model}
+\label{sec:model}
 
@@
 Where $c(x)$ is the number of occurences of $x$ in the corpus.
-The expression transition probability is defined by
+The expression transition probability is defined by the unigram count of $e_i$ divided by the bigram count of $e_{i-1}, e_i$.
 \begin{equation}
@@
 \section{Implementation}
+\label{sec:implementation}
 
-\begin{itemize}
-\item Extract melody
-\item Extract scorefeatures from melody
-\item lookup notes in deviation, ignore missing notes
-\item Extract expressive and non-expressive melody and extract expression features
-\item Perform notes other than melody notes the same as the nearest melody note
-\end{itemize}
+This section will discuss practical issues that we faced creating a complete performance rendering system based on the method described in the previous section.
+
 \subsection{Corpus and Representation}
-We were lucky to find and receive permission to use the CrestMusePEDB \ref{crestmuse} corpus, which is a corpus that contains expressive performances of Western classical music by famous pianists.
+\label{sec:corpus}
+We were lucky to find and receive permission to use the CrestMusePEDB \cite{hashida2008new}, which is a dataset that contains expressive performances of Western classical music by famous pianists.
 The music includes works by Bach, Beethoven and Chopin.
@@
 \subsection{Discretization}
+\label{sec:discretization}
 
@@
 \subsection{Smoothing}
+\label{sec:smoothing}
 
@@
 
-\section{Evaluation and Results}
+\section{Results}
+\label{sec:results}
 
-Using the method that we now have extensively described we have tried to generate a few performances of piano works by Mozart.
-The system was trained on 42 performances of 13 works by Mozart, see appendix \ref{appendix:corpus} for a more details.
-The choice for Mozart was motivated by the fact that our segmentation works best when phrase transitions are marked by long notes or silences, which is often the case in the work of Mozart.
-Another reason for choosing Mozart in favor of for example Chopin is that we think structure level expression is more outspoken in performances of Mozart's music.
+We compiled two corpora from our the CrestMusePEDB.
+One contained 42 performances of 13 piano works by Mozart, the other contained 49 performances of 19 works by Chopin.
+For more details on each corpus see appendix \ref{appendix:corpus}.
+We suspect that performances of Mozart's music most outspokenly use structural level expression.
+Expression within a constituent tends to be more unsteady in performances of Chopin's music.
+
+To test the system we trained it on one of the two corpora and let it perform a number of works.
+Here we show the results of four renderings of Chopin's Mazurka No.
+19, Op.
+30-2 and Mozart's Piano Sonata 310 III.
+To see what the effect was of each corpus on different styles of music we performed each work on both corpora.
+For reference, we also show a performance by one of the pianists in the corpus.
+Given the small dataset and the free nature of expression we did not expect the performances to correlate much with performances by human pianists.
+
+For all four performances the system was trained with a discretization factor of five. In practice, this means that the constituent features (
+\ref{sec:features}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well.
+The expression features where discretized into ten bins, with a sensitivity of five.
+The bins had to be this large to recognise most of the score features in the score that was to be performed.
+
+Segmentation was done using only delta onset trees. This proved to be most reliable for Mozart.
+
+When rendering a Mozart performance on the Mozart corpus and a Chopin performance on the Chopin corpus any of the performances of the piece that had to be performed were left out of the corpus during training.
+
+Plots showing dynamics, tempo and articulation of the resulting performances are shown in figure \ref{fig:mozartperformances} and \ref{fig:chopinperformances}.
+The values on the y-axis, labeled deviation, are the exponents of the expression parameters, which themselfs are logarithmic ratios.
+In other words, the values indicated by the axes are ratios.
+A tempo of $2.0$ means twice as fast as the average tempo, an articulation of $0.5$ means the notes are played twice as short as notated in the score, a dynamic value of $0.5$ means the notes are played half as loud as the average loudness of the performances.
+Likewise, to translate the output of the system into a performance, the mean tempo and loudness have to be set manually.
+
+The performances are plotted per note.
+The block like character of the axes is caused by the segmentation.
+For the sake of comparison, the performances by human pianists have been averaged over each constituent and discretized, so these plots show how the performance is represented in the corpus.
+
+Of course staring at plots of music performances is not nearly as interesting as actually listening to them.
+For this purpose, a few demo performances including the ones shown here can be found online at [DEMO URL].
 
@@
 
-We discretized the score features using a discretization parameter of five.
-This means the constituent features (see section \ref{sec:features}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well.
-The expression features where discretized into ten bins, with a sensitivity of five.
-These settings made sure we recognised most of the score features of a new score.
-The low number of bins does give the performances bit of a cartoonistic quality, but that does not make them any less interesting.
+\begin{figure}
+\centering
+\subfigure[Performance by the system trained on the Mozart corpus]{
+\includegraphics[scale=0.4]{img/Mozart-snt310-3_MozartNewSegmentation2_legend}
+\label{fig:donset}
+}
+\subfigure[Performance by the system trained on the Chopin corpus]{
+\includegraphics[scale=0.4]{img/Mozart-snt310-3_ChopinNewSegmentation}
+\label{fig:donset}
+}
+\subfigure[Performance by Maria Jo\~ao Pires]{
+\includegraphics[scale=0.4]{img/Mozart-snt310-3_pires}
+\label{fig:donset}
+}
+\caption{Three performances of Mozart's Piano Sonata 310 III.}
+\label{fig:mozartperformances}
+\end{figure}
 
-The segmentation was based only delta onset trees since this worked best for Mozart.
+\begin{figure}
+\centering
+\subfigure[Performance by the system trained on the Mozart corpus]{
+\includegraphics[scale=0.4]{img/Chopin-mzk019_MozartNewSegmentation2}
+\label{fig:donset}
+}
+\subfigure[Performance by the system trained on the Chopin corpus]{
+\includegraphics[scale=0.4]{img/Chopin-mzk019_ChopinNewSegmentation}
+\label{fig:donset}
+}
+\subfigure[Performance by Samson Francois]{
+\includegraphics[scale=0.4]{img/Chopin-mzk019_franc}
+\label{fig:donset}
+}
+\caption{Three performances of Chopin's Mazurka No. 19, Op. 30-2}
+\label{fig:chopinperformances}
+\end{figure}
 
-Since transition probabilities are not smoothed only transitions that actually occur in the corpus have a nonzero probability
+\section{Discussion}
+\label{sec:discussion}
 
-\subsection{Subjective Listening}
-\subsection{Correlation}
-\section{Integration with Note Level Performance Rendering}
+Despite the large bin size, many combinations of features and expression parameters were never encountered during training.
+The same applies to transitions of expression, only small portion of all possible transitions of expression are actually encountered in the corpus.
+Some of the feature likelyhoods are smoothed but the transition probabilities are not smoothed at all.
+Since our model does not allow zero probability transitions, our model is heavily overfitting.
+The resulting performances only contain changes of expression that were actually encountered in the corpus.
+
+In most applications of machine learning, this degree of overfitting would not be acceptable.
+However, in this particular application of machine learning we seem to be able to get away with it.
+Playing a phrase transitions exactly as encountered in the corpus seems to be acceptable as long as we play the right transitions at the right moment.
+The overfitting and the large bins size give the performances a bit of a stereotypical and cartoonistic feel, but that is mostly caused by the small corpus.
+Whether it makes musical sense to perform Chopin on a system trained on Mozart is left for the reader to decide.
+
+Although the output of the system is polyphonous, the system at its core only uses a single voice for its performances.
+We use the highest notes in the top staff for this and hope that these are the melody notes.
+Apart from the fact that this is not always the case, the other notes generally should note be played with the same expression.
+Bass notes for example are often played slightly softer and more legato than melody notes.
+
+The system does not make use of expressive markings in the score.
+A mature performance rendering system should incorporate these in some way.
+Even a human performer is not expected to read the mind of the composer and uses the expressive markings as a guide for performance.
+Disregarding articulation marks and dynamic markings has a significant impact on the results: the system has an overall slight preference for playing staccato and softly.
+This is due to the fact that staccato notes or notes that should be played soft are treated as normal notes and bias the statistics.
+
+Ignoring note level expression is quite an extreme measure to take when  it has been an adequate way to demonstrate the phenomenon of structure level expression.
+Yet, we cannot deny the need for note level expression in a performance rendering system that intends to completely simulate human expression.
+Section \ref{sec:futurework} suggests a possible integration of structure and note level performance rendering systems.
+
+%overfitting
+
+%Bach
+
+%Since transition probabilities are not smoothed only transitions that actually occur in the corpus have a nonzero probability
+
+%Unfortunately we do not have acces to the large dataset that YQX uses.
+The dataset we use is smaller.
+Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger.
+We have discretize into rather large bins for this reason, resulting in cartoonistic performances.
+
+%The system does not make use of expressive markings in the score.
+A mature performance rendering system should incorporate these in some way.
+Een a human perform is not expected to read the mind of the composer.
+
+%We think we can afford to do this because:...
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+In this thesis we have introduced a way of recognizing and using structure level expression.
+We argued why we think it is a crucial aspect of performance rendering systems and we critisized  note level performance rendering for not being able to correctly learn when to make daring and large changes in expression as well as lacking the ability to convey structure level expression.
+
+Our goal was to create a system that was the complement of a note level performance rendering system.
+A system that dared to make sudden big changes in expression and clearly accentuates musical structure.
+Our results show that our system indeed is not shy this kind of big expressive changes.
+The performances of the system contain sometimes smooth, sometimes sudden changes of expression.
+It is up to the listener to judge whether these changes are actually an improvement over the smoother performances of note level performance rendering systems.
+
+The system that we proposed reduces the idea of structure to a segmentation of the score into constituents based on a structural analysis.
+For the structural analysis we used the delta framework\ref{markwin}, proposed by Markwin van den Berg.
+We noticed that a segmentation based on onset delta trees works well for some music but does not recognize all constituent transitions.
+To overcome this we suggested to use a second order pitch delta tree, but we ended up discarding this idea because it still produced awkward segmentations.
+Our system would therefore not work very well on music that has a consistend rhythmic structure like some of Bach's music.
+In the end, an onset tree segmentation worked good enough to produce musically relevant segmentations given the the right genre of music.
+
+To characterize constituents and their relations to other constituents we used constituent likely-hoods and bigram transition probabilities of expression.
+The resulting hidden Markov model could be used in combination with Viterbi decoding to generate performances.
+This approach is very similar to stochastic part of speech tagging using hidden Markov models.
+
+\section{Future Work}
+\label{sec:futurework}
+
+The abilities of performance rendering systems still are not even close to the abilities of real human pianists.
+In relation to this system we can clearly identify a number of things that would bring the system a little bit closer to the abilities of a human pianist.
+
+The segmentation is not always perfect.
+The delta framework offers us more analytic power than we have used in this system but we were not able to sensibly combine its analyses into one segmentation apporach.
+We are unsure how we should combine different delta trees into some sort of consensus.
+
+Such a consensus tree would be especially useful in an extension of our system to a true structure based performance rendering system, namely one that utilizes multilevel structure.
+
+Another open problem is how we should deal with multiple voices.
+Some proposals have been made in this field.
+Tae Hun Kim et al.
+\cite{kim2010performance} split a piece into melody, bass and harmony and use separate probabilistic models for each of them.
+When using different expression for different voices, another problem presents itself, namely how to synchronize the voices so that the performance still sounds like it is played by one artist or a group of artists with an adequate notion of timing.
+Pop-e \cite{hashida2006pop} uses an automatic analysis to determine the attentive part: the voice of the music perceived as melody at any point in the piece.
+It uses this information to synchronize expression in different voices.
+
+As we mentioned in section \ref{sec:discussion}, our system ignores expressive markings.
+The result is that depending on the corpus, the performance renderer has unwanted biases.
+A possible approach to incorporating expressive markings in the system is to determine what, given an expressive mark in the, is the average way to play a given note and normalize this out of the corpus.
+Say staccato notes are played, on average, half the length of the notated length.
+We specify articulation deviation relative to this half duration instead of the full duration.
+
+We hope that in the future, as the CrestMusePEDB project develops, data sparsity will be less of a problem.
+For now however, we are left with having to do discretization and smoothing.
+The smoothing technique we used, Good-Turing smoothing, is arguably not the best way to smooth the corpus.
+Good-Turing smoothing assigns equal probabilities to all unseen score features.
+It is likely that unseen scorefeatures that are very similar to scorefeatures that we have seen should also be played in a similar way as the scorefeatures they resemble.
+A different smoothing technique may be able to use this information.
+We could use a smoothing technique similar to Katz-Backoff REF: upon encountering an unseen feature vector, discretize again the corpus into increasingly larger bins until we do find the featurevector in our corpus.
+
+The system would certainly benefit from a notion of repetition and similarity.
+Repetition is a very good indicator of constituent breaks.
+Repetition and similarity could also be used to improve expressiveness of performances.
+It is probably telling when a phrase is repeated three times and then slightly altered the fourth time.
+Although finding similarity and musically significant repetition is a subject of its own the delta framework could help to define repetition arbitrarily of transposition or rhythm.
+A list of pitch deltas can for example be used to detect repetition independend of transposition and a list of duration deltas can be used to detect repetition of rhythm independend of the notes used.
 
@@
 
-An intergration of constituent level expression and note level expression can provide a solution to this problem.
-We can define expressive tempo relative and dynamics relative to the expression parameters of the constituent
+An integration of constituent level expression and note level expression can provide a solution to this problem.
+We can define expressive tempo relative and dynamics relative to the expression parameters of the constituent.
+A possible approach is to first define constituent level expression and use a note level expression system on each constituent seperately.
 
-\section{Conclusion}
-In this thesis we have critisized performance rendering systems that only use note level expression.
+Another approach is to use hierarchical structure.
+Expression of every constituent can be defined relative to the parent constituent.
+If we let the hierarchy extend to the note level (the delta framework does this) we get a smooth integration of note level performance rendering and structure level performance rendering.
+In such a system notes are terminal notes in a structure tree, note level expression is specified relative to the parent constituent's expression which is again relative to the its own parent constituent.
+Ultimately the root node's expression contains the base expressive parameters.
 
-We have introduced structure level expression and argued that a good performance rendering systems needs some notion of structure level expression.
-
-We have shown that we can create sensible segmentations of music based on an application of the delta framework. \cite{markwin}
-
-We have suggested a possible intergration of SBPR with note level performance rendering
-\section{Discussion}
-
-Unfortunately we do not have acces to the large dataset that YQX uses.
-The dataset we use is smaller.
-Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger.
-We have discretize into rather large bins for this reason, resulting in cartoonistic performances.
-
-We think we can afford to do this because:...
-
-\begin{itemize}
-\item Top notes is not always the melody, musical attention
-\item Bass and harmony shouldn't be played with the same expression as melodynotes
-\item Dataset size
-\item Targets for expression level expression
-\item Expressive markings
-\item Staccato/articulation statistics
-\item Backoff smoothing (doesn't assign equal probabilities to unseen events)
-\end{itemize}
-
-\section{Future Work}
-\begin{itemize}
-\item Combine deltatrees
-\item Use loudness and tempo direction instead of means (requires more data)
-\item Generalize approach to incorporate hierarchical structure, let structure extend to note level, so note level expression and structurelevel expression become integrated
-\item Backoff smoothing
-\item Staccato statistics
-\end{itemize}
-\subsection{Repetition}
-The system would certainly benefit from a notion of repetition and similarity.
-Repetition is a very good indicator of constituent breaks.
-Repetition and similarity could also be used to improve expressiveness of performances.
-It is probably telling when a phrase is repeated three times and then slightly altered the fourth time.
-Although finding similarity and musically significant repetition is a subject of its own the delta framework could help to define repetition arbitrarily of transposition or rhythm.
-A list of pitch deltas can for example be used to detect repetition independend of transposition and a list of duration deltas can be used to detect repetition of rhythm independend of the notes used.
+Finally, as we mentioned in section \ref{sec:approach} the ability of the system to characterize constituents and their relations to other constituents in a musically relevant way is an important factor influencing the results.
+We do not have the illusion that a set of four scorefeatures and bigram transition probabilities are a valid model of human musical intuition, let alone producing highly expressive performances.
+A better notion of repetition, a notion of key, harmony and general musical knowledge are the least that seems to be required for that purpose.
 

### 2011-07-21 Paper/Paper.tex
Draft finished

@@
 However \textit{expressively} performing music is not and much research has focussed on this issue.
-Such research is interesting from a scientific point of view as well as a practical point of view.
-Artificial intelligence, cognitive science and musicology may all benefit from a better understanding of expression in music.
+A wide selection of scientific fields may have interest in such research: Artificial intelligence, cognitive science and musicology all benefit from a better understanding of expression in music.
 
@@
 It is debatable how much we should value correlation of generated performances with human performances of the same piece.
-A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both can be seen valid interpretations.
+A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both are considered high quality interpretations.
 This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience.
@@
 
+What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}.
 Section \ref{sec:approach} will describe the system as a whole.
-What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}.
 Section \ref{sec:method} will describe every component of the system in detail.
+Practical issues are discussed in section \ref{sec:implementation}.
+Results are presented in section \ref{sec:results} and discussed in section \ref{sec:discussion}.
+Suggestions for improving the system and future research can be found in section \ref{sec:futurework}.
 
@@
 
-The YQX system, as the authors admit, tended to sometimes produce nervous sounding changes in expression.
-He presents two extensions that seek to generate smoother performances.
-The problem with these extensions is, as Widmer admits himself, that the increased smoothness comes at the expense of expressitivity.
-To counter this, he adds three explicit rules that he uses to postprocess the performances.
-We think the reason that Widmer stumbled upon this tradeoff between expressiveness and nervousness because the nervousness is inherent to the way performances are generated at note-level.
-The system simply misses one component of expression, namely structure level expression.
+The YQX system, as Flossman et al.
+admit, tended to sometimes produce nervous sounding changes in expression.
+They present two extensions tailored towards generating smoother performances.
+The problem with these extensions is, as Flossman et al.
+also admit himself, is that the increased smoothness comes at the expense of expressitivity.
+To compensate for this, three explicit rules are added to postprocess the performances.
+We think the reason that Flossman et al.
+stumbled upon this tradeoff between expressiveness and nervousness is that the nervousness is inherent to note level performance rendering.
+Note level systems simply are not suited to capture one component of expression, namely structure level expression.
 
@@
 The system presented here ignores note level expression.
-Instead we will try to predict only constituent level expression.
+Instead we will try to predict only constituent (or structure) level expression.
 The assumption is that this kind of expression really exists in performances and that is different and independent from note level structure.
 We think that a constituent level system also corresponds better to how actual human performers play music.
-A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.
+%A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.
 
@@
 
-The system will be similar to YQX in a number of ways, but instead of predicting expression per note, it will predict expression per constituent.
+The system will be similar to YQX in a number of ways, but with the crucial difference that expression will not be predicted per note, but per constituent.
 Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.
@@
 
-The delta framework is based on the intuition that differences between notes indicate splits between constituents.
+The delta framework is based on the intuition that differences between features of notes indicate splits between constituents.
 The higher the difference, the higher the level of split in the parse tree (where the root note is at the highest level).
@@
 These are: \textbf{onset}, \textbf{pitch}, \textbf{loudness} and \textbf{release} (called offset by Van den Berg).
-From these two other features can be derived: \textbf{duration} and \textbf{virtual duration}.
+From these, two other features can be derived: \textbf{duration} and \textbf{virtual duration}.
 Duration is defined as \textbf{release($n_i$)} - \textbf{onset($n_i$)} while virtual duration is defined as \textbf{onset($n_{i+1}$)} - \textbf{onset($n_i$)}.
@@
 A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure.
-The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed and $A$ is the resulting analysis, is defined as \footnote{The version here is a slightly reformulated, although functionally eqivalent, version of Van den Berg's deltarule.
+This ruled, called the delta rule is shown in algorithm \ref{alg:deltarule}, where DeltaRule($M_{ij}$) is a recursive call to the algorithm itself, $M_{ij}$ is the ordered list of notes to be analyzed and $A$ is the resulting analysis.\footnote{The version here is a slightly reformulated, although functionally equivalent, version of Van den Berg's deltarule.
 See \cite{markwin} for his original version.}:
@@
 \caption{The delta rule}
-\label{deltarule}
+\label{alg:deltarule}
 \end{algorithm}
 
-The delta rule converts a sequence of notes into a nested list structure that can easily be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.
+The delta rule converts a sequence of notes into a nested list structure that can be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.
 
-It also possible to define higher order delta function (deltas of deltas).
+It is also possible to define higher order delta function (deltas of deltas).
 Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches.
@@
 For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$.
-Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[n_1, n_2], n_3]$.
+Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[[n_1, n_2], n_3]$.
 This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note.
@@
 
-Having defined first and second order deltafunctions and the deltarule, we can parse a piece of music into a delta tree or parse tree.
+Having defined first and second order deltafunctions and the deltarule, we can use first and second order deltafunctions to parse a piece of music into a parse tree.
 Since we have a set of six basic note features, we have a set of six interpretations(parse trees) available for a piece of music.
@@
 A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music.
-However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, derrived from deltatrees.
+However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, which can be more reliably derrived from deltatrees.
 
-The goal is to find a `safe' method to segment music, that is, a method that does not generate segmentations that clash too much with human musical intuition.
-For this purpose the parse trees of delta onset (inter-onset interval, or simply IOI) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions.
-These two trees are displayed in figure \ref{fig:pitchonsettrees}.
-
+The goal is to find a `safe' method to segment music, that is, a method that generates segmentations that do not clash with human musical intuition.
+For this purpose the parse trees of delta onset (inter-onset interval, or simply \textsc{IOI}) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions.
 \begin{figure}
@@
 
-In the trees in figure \ref{fig:pitchonsettrees} we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes.
+A delta onset tree and a second order delta pitch tree of the first eight bars of Mozart's Turkish March can be found in figure \ref{fig:pitchonsettrees}.
+In his figure we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes.
 To translate the tree into a segmentation we use algorithm \ref{alg:segmentation} starting with the root node.
@@
 \label{alg:segmentation}
+For some node $n$, do:
 \begin{enumerate}
-\item If the node is a note, at it as a singleton constituent to the segmentation
-\item Expand the node.
-\item If the node contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
-\item If the node contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
+\item If $n$ is a note, add it as a singleton constituent to the segmentation
+\item Expand $n$.
+\item If $n$ contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
+\item If $n$ contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
 \end{enumerate}
@@
 We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart.
-Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if you use the delta tree from figure  \ref{fig:donset} you will not get this segmentation.
+Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if the delta tree from figure  \ref{fig:donset} is used the segmentation will be different.
 This segmentation was based on a onset delta tree over the entire piece instead of just the first eight bars}
@@
 
-%Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes.
+%Using this modification of the delta rule we can bring the delta onset tree down to a shallow tree in which it is safe to choose groupings defined by the first level nodes.
 
-For some music like Mozart, using only a delta onset tree works very well.
-However, some music, like Bach's inventionen uses almost no differentiation in IOI.
-Hence, a segmentation based solely on inter onset interval will not serve our purposes for this music.
+For some music like Mozart's, using only a delta onset tree works very well.
+However, some music, like Bach's Inventionen uses almost no differentiation in IOI.
+Hence, a segmentation based solely on inter onset interval will not work well for these pieces.
 
 To overcome this problem we will use pitch.
-While Bach's inventionen use almost no differentiation in IOI they can be segmented nonetheless.
+While Bach's Inventionen use almost no differentiation in IOI they can be segmented nonetheless.
 We will not make any attempt to combine pitch interval trees and IOI trees.
@@
 
-The sort of changes in pitch that indicate constituent transitions or not characterized very well by a first order pitch interval tree.
+The sort of changes in pitch that indicate constituent transitions are not characterized very well by a first order pitch interval tree.
 For example a melody that has four notes that are chromatically ascending followed by four notes that have octave intervals between them should be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals.
 A first order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments.
-For this reason we use a second order pitch interval tree that does not suffer from this.
+A second order pitch interval puts the split in this case exactly where we would want it.
 
@@
 
-We can now convert a piece of music into a serie of constituents.
-These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed.
+We can now convert a piece of music into a series of constituents.
+These constituents will be used to predict expression so we must be able to charaterize them in a way that correlates with the way they are performed.
 Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.
@@
 The duration ratio is the logarithmic ratio of the duration of the current note and the duration of the next note.
-The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis.
+The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis \cite{narmour1990analysis}.
 
@@
 At note level there is not much to say about the current note besides the pitch and duration.
-However at constituent we would also like to say something about the constituent itself.
+However at constituent level we would also like to say something about the constituent itself.
 For this purpose the \textit{constituent features} \textbf{mean delta pitch} and \textbf{mean delta duration} are used.
@@
 
-The complete set of score features consists of the two context features and the two constituent features:
+The complete set of score features consists of the two context features and the two constituent features.
 
@@
 Awkward because if a score note is not followed by a rest, the notated silence after it is zero, rendering articulation undefined.
-Inconsisted because all the other features use logarithmic ratios and we see no reason to not use a logarithmic ratio for articulation as well.
+Inconsisted because all the other features use logarithmic ratios and we see no reason not to use a logarithmic ratio for articulation as well.
 
-IOI ratio and loudness is defined relative to mean performance tempo and loudness.
+IOI ratio and loudness are defined relative to mean performance tempo and loudness.
 However, to capture micro expression in the form of small changes in onset relative to the beat, it seems more logical to define this feature relative to the local tempo, instead of relative to the global tempo.
@@
 
-We can now reduce a piece of music, represented as a sequence of notes $N_{ij}$, of which we have a score and a performance, to a sequence of score feature vectors $F$ and expression parameter vectors $E$.
+We can now reduce a piece of music, represented as a sequence of notes $N_{ij}$, of which we have a score and a performance, to a sequence of \textit{score feature vectors} $F$ and \textit{expression parameter vectors} $E$.
 First, we segment the score into constituents:
@@
 
-If we parse a number of pieces is this manner we can use it to do some statistics.
-We can for example calculate the features likely-hood which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$.
+We can construct a corpus by parsing a number of works in this manner.
+Given such a corpus, the rendering of a performance can be formulated as maximizing the probability: $P(E|F)$.
+We can estimate an approximation of this probability by calculating feature likelyhoods and transition probabilities.
+We approximate the features likely-hood, which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$, with the following probability:
 
@@
 Where $c(x)$ is the number of occurences of $x$ in the corpus.
-The expression transition probability is defined by the unigram count of $e_i$ divided by the bigram count of $e_{i-1}, e_i$.
+The expression transition probability is approximated by the unigram count of $e_i$ divided by the bigram count of $e_{i-1}, e_i$.
 \begin{equation}
@@
 
-For all four performances the system was trained with a discretization factor of five. In practice, this means that the constituent features (
-\ref{sec:features}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well.
+For all four performances the system was trained with a discretization factor of five.
+In practice, this means that the constituent features (see section \ref{sec:scorefeatures}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well.
 The expression features where discretized into ten bins, with a sensitivity of five.
@@
 The system that we proposed reduces the idea of structure to a segmentation of the score into constituents based on a structural analysis.
-For the structural analysis we used the delta framework\ref{markwin}, proposed by Markwin van den Berg.
+For the structural analysis we used the delta framework \cite{markwin}, proposed by Markwin van den Berg.
 We noticed that a segmentation based on onset delta trees works well for some music but does not recognize all constituent transitions.

### 2011-08-10 Paper/Paper.tex
Report finished.

@@
 
-% Multiline tables
+% Multi line tables
 \usepackage{array}
@@
 
-% [DO NOT CHANGE] The addess of the educational programme.
+% [DO NOT CHANGE] The address of the educational programme.
 University of Amsterdam\\
@@
 % Be sure to include (1) institute (is appropriate), (2) faculty (if
-% appropriate), (3) organisation name, (4) organisation address (2 lines).
+% appropriate), (3) organization name, (4) organization address (2 lines).
 Institute for Language and Logic\\
@@
 A dataset containing performances aligned to scores is then used to learn how structure relates to expression.
-A hidden markov model is used to model this relation.
+A hidden Markov model is used to model this relation.
 Finally, the system can be applied to a a new score to produce an expressive performance.
@@
 
-{\bf Keywords:} performance rendering, musical structure, constituent structure, polyphonic piano music
+{\bf Keywords:} performance rendering, musical structure, constituent structure
 \end{abstract}
@@
 Given a hypothetical perfect synthesizer, performing scores with computers is a trivial task.
-However \textit{expressively} performing music is not and much research has focussed on this issue.
+However, \textit{expressively} performing music is not, and much research has focussed on this issue.
 A wide selection of scientific fields may have interest in such research: Artificial intelligence, cognitive science and musicology all benefit from a better understanding of expression in music.
@@
 A rather practical approach to expressive music performance is to recreate it.
-A performance rendering system is a computer program that automatically creates an expressive, humanlike performance of a musical score.
+A performance rendering system is a computer program that automatically creates an expressive, human-like performance of a musical score.
 Performance rendering systems have always faced a problem of evaluation.
 It is debatable how much we should value correlation of generated performances with human performances of the same piece.
-A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both are considered high quality interpretations.
-This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience.
+A Chopin interpretation by Arthur Rubinstein can be radically different from a Chopin interpretation by Vladimir Horowitz, yet both are considered high quality interpretations.
+This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience \cite{hiraga2002rencon}.
 
@@
 Early attempts, like Director Musices \cite{friberg2000generating}, have applied sets of rules to the problem.
-As the availablity of corpora grew, machine learning systems have been proposed as well.
+As the availability of corpora grew, machine learning systems have been proposed as well.
 Gerhard Widmer is working with a large dataset containing almost all works by Chopin, performed by one pianist.
@@
 Some do use structure.
-Pop-E\cite{hashida2006pop}, a rule based performance rendering system, uses a structure analysis based on Lerdahls and Jackendoff's grouping preference rules, but the structural analysis must be supplied to it by a human expert.
+Pop-E \cite{hashida2006pop}, a rule based performance rendering system, uses a structure analysis based on Lerdahl and Jackendoff's grouping preference rules \cite{lerdahl1996generative}, but the structural analysis must be supplied to it by a human expert.
 In this thesis we will motivate the need for structure level expression and propose a system that purely uses structure level expression.
@@
 Practical issues are discussed in section \ref{sec:implementation}.
-Results are presented in section \ref{sec:results} and discussed in section \ref{sec:discussion}.
+Results are presented in section \ref{sec:results} and evaluated in section \ref{sec:discussion}.
 Suggestions for improving the system and future research can be found in section \ref{sec:futurework}.
 
-%Early performance rendering systems used rule-based approaches to recognise small local structures that are used as indications to add certain kind of expression.
+%Early performance rendering systems used rule-based approaches to recognize small local structures that are used as indications to add certain kind of expression.
 A prominent ongoing rule based system, \textit{Director Musices} uses 30 rules with weighting parameters that take music features as input and produce expressive actions as output.
@@
 The problem with these extensions is, as Flossman et al.
-also admit himself, is that the increased smoothness comes at the expense of expressitivity.
-To compensate for this, three explicit rules are added to postprocess the performances.
+also admit themselves, that the increased smoothness comes at the expense of expressivity.
+To compensate for this, three explicit rules are added to post-process the performances.
 We think the reason that Flossman et al.
@@
 Sudden changes in expression occur frequently in expressive performances, a few notes may be played very loud, or very slow followed by soft notes played very fast.
-When done at the right moment, such sudden changes can greatly improve the humanlike feel of the performance, however when done at the wrong moment, they can ruin a performance.
+When done at the right moment, such sudden changes can greatly improve the human-like feel of the performance, however when done at the wrong moment, they can ruin a performance.
 We think that at structure level it will be easier to capture the context in which sudden changes occur.
@@
 In a performance, this structure may be accentuated in different ways.
-Accentuation happens at different levels, at note level performers may slow down at the end a phrase or introduce small pauses in the performance at phrase transitions.
+Accentuation happens at different levels, at note level performers may slow down at the end of a phrase or introduce small pauses in the performance at phrase transitions.
 At constituent level one phrase may be played very loud, fast or staccato, while the next may be played slow, soft and legato.
@@
 In this analogy we see a piece of music as a sentence, which consists of constituents that individually can be made of constituents as well.
-We can recursively subdivide constituents in more constituents until we reach a terminal symbol.
+We can recursively subdivide constituents into more constituents until we reach a terminal symbol.
 In NLP this may be a word, in music, this may be a note.
@@
 
-We must note that musical scores can be highly ambiguous and even experienced listeners may not agree on the correct parsetree of a piece.
+We must note that musical scores can be highly ambiguous and even experienced listeners may not agree on the correct parse tree of a piece.
 Quite often there may simply be more than one parse tree that makes musical sense.
@@
 
-Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, expression is only predicted per note.
+Although the YQX does have some notion of structure\footnote{One of YQX's note features is distance to nearest point of closure, see section \ref{sec:scorefeatures}.}, expression is only predicted per note.
 The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''.
@@
 Instead we will try to predict only constituent (or structure) level expression.
-The assumption is that this kind of expression really exists in performances and that is different and independent from note level structure.
+The assumption is that this kind of expression really exists in performances and that it is different and independent from note level structure.
 We think that a constituent level system also corresponds better to how actual human performers play music.
-%A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.
+%A performance rendering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.
 
@@
 The system will be similar to YQX in a number of ways, but with the crucial difference that expression will not be predicted per note, but per constituent.
-Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.
+Every constituent will be played with consistent expression, the articulation, dynamics and tempo change only at constituent breaks.
 
@@
 
-A structural analysis is used to derive a hierarchical structure for every score in the corpus, however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents.
+A structural analysis is used to derive a hierarchical structure for every score in the corpus; however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents.
 After segmentation, four score features, two of which are direct generalizations of YQX's score features, are extracted for each constituent.
@@
 Since we have a segmentation and every score note is associated with a performance note\footnote{In reality, not every score note is associated with a performance note since the pianist may have missed some notes.
-These notes will be ignored} we can also define expression per constituent.
+These notes will be ignored.} we can also define expression per constituent.
 Three parameters, analogous to YQX's targets, will be used to describe expression per constituent.
@@
 
-The resulting data is used to train a first order hidden Markov model.
+The resulting data is used to train a first-order hidden Markov model.
 The system uses this model to generate performances given a score.
 To do this, the score is segmented into constituents, score features are extracted for each constituent.
-Finally viterbi decoding is used to find the sequence of expression parameters that best explains the observed score features.
+Finally Viterbi decoding is used to find the sequence of expression parameters that best explains the observed score features.
 
-The succes of a SBPR system depends largely on two factors.
+The success of a SBPR system depends largely on two factors.
 The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents in score features.
-The following section address these issues.
+The following section addresses these issues.
 
-%The following section will introduce an application of the delta framework \ref{markwin} that intends to segment a score into musically meaningfull constituents.
-Section \ref{scorefeatures} will deal with the characterization of individual constituents, section \ref{expressionfeatures} will explain how expressive parameters of a constituent are determined and translated into expressioin.
+%The following section will introduce an application of the delta framework \ref{markwin} that intends to segment a score into musically meaningful constituents.
+Section \ref{scorefeatures} will deal with the characterization of individual constituents, section \ref{expressionfeatures} will explain how expressive parameters of a constituent are determined and translated into expression.
 Section \ref{performancerendering} will explain how performances are
@@
 The structural analysis is based on the delta framework, which will be described in section \ref{sec:deltaframework}.
-Section \ref{sec:segmentation} will discuss how the delta framework is applied to get a segmentation.
-Sections \ref{sec:scorefeatures} and \ref{sec:targets} will describe how the scorefeatures and expression parameters are calculated.
+Section \ref{sec:segmentation} will discuss how the delta framework is used to get a segmentation.
+Sections \ref{sec:scorefeatures} and \ref{sec:targets} will describe how the score features and expression parameters are calculated.
 Section \ref{sec:model} will describe how a hidden Markov model is trained on our data.
@@
 
-In his Phd thesis \cite{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework.
-He relates his work to the work of Lerdahl and Jackendoff \cite{lerdahl1996generative} but claims to have found a more general approach.
-Below, I will shortly describe the delta framework as proposed by Van der Berg.
+In his Ph.
+D.
+thesis \cite{markwin}, Markwin van den Berg introduces a formal way of parsing music into parse trees: the delta framework.
+He relates his work to the work of Lerdahl and Jackendoff's Generative Theory of Tonal Music \cite{lerdahl1996generative} but claims to have found a more general approach.
+Below, I will shortly describe the delta framework as proposed by Van den Berg.
 
 The delta framework is based on the intuition that differences between features of notes indicate splits between constituents.
-The higher the difference, the higher the level of split in the parse tree (where the root note is at the highest level).
-Van der Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parsetree following this intuition.
+The larger the difference, the higher the level of the split in the parse tree (where the root note is at the highest level).
+Van den Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parse tree following this intuition.
 
@@
 
-\[M_{ij} = [n_i, n_{i+1}, \cdots, n_j]\]
+\[M_{i,j} = [n_i, n_{i+1}, \cdots, n_j]\]
 
@@
 
-The basic note features can be used to define delta functions, for example $\Delta\textbf{Pitch} = \textbf{Pitch}(n_i) - \textbf{Pitch}(n_{i-1})$.
-In general, a delta function $\delta(i)$ is defined as the difference of two notes in some feature $\phi$: $\delta(i) = \phi(n_i) - \phi(n_{i-1})$.
+The basic note features can be used to define delta functions, for example $\delta_{\textbf{Pitch}} = \textbf{Pitch}(n_i) - \textbf{Pitch}(n_{i-1})$.
+In general, a delta function $\delta(i)$ is defined as the difference of two notes in some feature $\phi$: $\delta_{\phi}(i) = \phi(n_i) - \phi(n_{i-1})$.
 We can apply a delta function to every pair of succeeding notes in a sequence to get a list of deltas:
-\[\Delta M_{ij} = [\delta(n_{i+1}), \delta(n_{i+2}), \cdots, \delta(n_{j})]\]
+\[\Delta_{\phi} M_{i,j} = [\delta_{\phi}(n_{i+1}), \delta_{\phi}(n_{i+2}), \cdots, \delta_{\phi}(n_{j})]\]
 
-A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure.
-This ruled, called the delta rule is shown in algorithm \ref{alg:deltarule}, where DeltaRule($M_{ij}$) is a recursive call to the algorithm itself, $M_{ij}$ is the ordered list of notes to be analyzed and $A$ is the resulting analysis.\footnote{The version here is a slightly reformulated, although functionally equivalent, version of Van den Berg's deltarule.
-See \cite{markwin} for his original version.}:
+A recursive rule, called the delta rule, can be used to translate a list of deltas into a grouping structure.
+This rule, called the delta rule is shown in algorithm \ref{alg:deltarule}, where DeltaRule($\phi, M_{i,j}$) is a recursive call to the algorithm itself, $\phi$ is the feature that we use for the delta function, $M_{i,j}$ is the ordered list of notes to be analyzed and $A$ is the resulting analysis.\footnote{The version here is a slightly reformulated, although functionally equivalent, version of Van den Berg's delta rule.
+See \cite{markwin} for his original version.} The algorithm is initialized with the entire piece of music, $M_{0, n}$ and the feature, $\phi$, to use for the delta function: DeltaRule($\phi, M_{0,n}$).
 \begin{algorithm}
 \begin{algorithmic}
-\STATE{$D_{i+1,j} \leftarrow \Delta M_{ij}$}
+\STATE{$D_{i+1,j} \leftarrow \Delta_{\phi} M_{i,j}$}
 \STATE{$A \leftarrow []$}
-\STATE $m \leftarrow \textbf{ max}(D)$
-\FOR{$\delta \textbf{ in } D$}
-\IF{$\delta = m$}
-\STATE $p = \textbf{ index of }\delta \textbf{ in } D$
-\STATE $q = \textbf{ index of the next occurence of m in }D \textbf{ or } j$
-\STATE $\textbf{append DeltaRule(}M_{pq}\textbf{) to } A$
+\STATE $m \leftarrow \mbox{ max}(D)$
+\STATE $p \leftarrow i$
+\FOR{$d \mbox{ in } D$}
+\IF{$d = m$}
+\STATE $q \leftarrow \mbox{ index of }d \mbox{ in } D$
+\STATE $\mbox{\textbf{append} DeltaRule(}\phi, M_{p,q}\mbox{) to } A$
+\STATE $p \leftarrow q$
 \ENDIF
 \ENDFOR
+\STATE $\mbox{\textbf{append} DeltaRule(}\phi, M_{p,j}\mbox{) to } A$
 \RETURN{$A$}
@@
 
-The delta rule converts a sequence of notes into a nested list structure that can be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.
+The delta rule converts a sequence of notes into a nested list structure that can be interpreted as a tree.
+The delta rule parses a piece of music into a \textit{parse tree}.
+The resulting parse tree depends on what feature $\phi$ we used for the delta function.
+Every feature yields a different parse tree.
 
 It is also possible to define higher order delta function (deltas of deltas).
-Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches.
-Van den Berg notes that this generates ambigous grouping structures as for example a second order deltarule needs at least three notes to be specified.
+Such higher order functions can detect patterns that would remain undetected by first-order delta function.
+Second order pitch delta functions, for example, can be used find constituent splits between a group of steadily ascending notes and a group of notes that have wildly varying pitches.
+
+%Van den Berg notes that this generates ambiguous grouping structures as for example a second order delta rule needs at least three notes to be specified.
 For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$.
-Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[[n_1, n_2], n_3]$.
-This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note.
-Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.
+Because we want to use a second order delta function for segmentation only, we choose to use only one interpretation: $[[n_1, n_2], n_3]$.
+This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order delta function) we would want a constituent split to happen between the second and third note.
+Note again that segmentation, like structure, may often be ambiguous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.
 
-Having defined first and second order deltafunctions and the deltarule, we can use first and second order deltafunctions to parse a piece of music into a parse tree.
-Since we have a set of six basic note features, we have a set of six interpretations(parse trees) available for a piece of music.
+%Having defined first and second order delta functions and the delta rule, we can use first and second order delta functions to parse a piece of music into a parse tree.
+
+Since we have six basic note features, we have six first-order delta functions and six interpretations (parse trees) for a piece of music.
+Although each parse tree provides us musically relevant information, we would like to have only one tree.
 Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent.
@@
 Unfortunately, the yield rule addresses the problem of how to interpret multiple parse trees quite poorly.
-We would like to have some way of combining parsetrees into one `consensus tree'.
+We would like to have some way of combining parse trees into one `consensus tree'.
 This is not what the yield rule does.
@@
 A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music.
-However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, which can be more reliably derrived from deltatrees.
+However, because of the problems with interpreting different parse trees and because delta trees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, which can be more reliably derived from delta trees.
 
@@
 \centering
+
 \subfigure[Delta onset tree]{
 \includegraphics[scale=0.4]{img/onset_sym}
-\label{fig:donset}
+\label{fig:subfig1}
 }
+
 \subfigure[Second order delta pitch tree]{
 \includegraphics[scale=0.3]{img/second_order_pitch_sym}
-\label{fig:dpitch}
+\label{fig:subfig2}
 }
@@
 A delta onset tree and a second order delta pitch tree of the first eight bars of Mozart's Turkish March can be found in figure \ref{fig:pitchonsettrees}.
-In his figure we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes.
-To translate the tree into a segmentation we use algorithm \ref{alg:segmentation} starting with the root node.
+In this figure we can see that some nodes contain only other nodes, some nodes contain only notes and some nodes are mixed and contain notes and other nodes.
+We will use this differentiation between node types in the translation of a tree into a segmentation, which is shown in algorithm \ref{alg:segmentation}.
+The algorithm is initialized with the root node.
 \begin{algorithm}
@@
 \begin{enumerate}
-\item If $n$ is a note, add it as a singleton constituent to the segmentation
-\item Expand $n$.
-\item If $n$ contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
-\item If $n$ contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
+\item \textbf{If} $n$ is a note.\\
+\textbf{Then} add $n$ as a singleton segment to the segmentation.\\
+ \textbf{Else} expand $n$.
+\item \textbf{If} $n$ is mixed and directly contains more than $x$ notes.\\
+\textbf{Then} add all notes recursively contained by $n$ as a segment to the segmentation.\\
+\textbf{Else} start from step 1 for every child of $n$.
+% If $n$ contains at least $x$ notes, add the notes recursively contained by $n$ to the segmentation.
 \end{enumerate}
@@
 
-The value of $x$ represents the tolerance of singleton constituents.
-We need this parameter because sometimes very long notes cause top level splits in the onset tree.
-We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart.
-Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if the delta tree from figure  \ref{fig:donset} is used the segmentation will be different.
-This segmentation was based on a onset delta tree over the entire piece instead of just the first eight bars}
+The value of $x$ represents the tolerance of singleton segments.
+We need this parameter because, in the delta onset tree, very long notes cause one note to split of a constituent at a high level, in this case we want to accept that this one note will be a singleton segment.
+We have found that setting $x$ to a variable number, namely: the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart.
+Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if the delta tree from figure  \ref{fig:subfig2} is used the segmentation will be different.
+This segmentation was based on an onset delta tree over the entire piece instead of just the first eight bars.}
 
@@
 
-For some music like Mozart's, using only a delta onset tree works very well.
-However, some music, like Bach's Inventionen uses almost no differentiation in IOI.
-Hence, a segmentation based solely on inter onset interval will not work well for these pieces.
+For some works, like most works by Mozart, using only a delta onset tree works very well.
+However, for other, like Bach's Inventionen, a delta onset tree is not sufficient since the music contains almost no differentiation in IOI (most notes are equidistant in onset).
 
 To overcome this problem we will use pitch.
-While Bach's Inventionen use almost no differentiation in IOI they can be segmented nonetheless.
+Even though Bach's Inventionen use almost no differentiation in IOI they can be segmented nonetheless.
 We will not make any attempt to combine pitch interval trees and IOI trees.
@@
 
-The sort of changes in pitch that indicate constituent transitions are not characterized very well by a first order pitch interval tree.
-For example a melody that has four notes that are chromatically ascending followed by four notes that have octave intervals between them should be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals.
-A first order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments.
+The sort of changes in pitch that indicate constituent transitions are not characterized very well by a first-order pitch interval tree.
+Let us take the example from the previous section, where we have a melody of eight notes, four of which are chromatically ascending followed by four notes that have octave intervals between.
+We want this to be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals.
+A first-order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments.
 A second order pitch interval puts the split in this case exactly where we would want it.
@@
 We can now convert a piece of music into a series of constituents.
-These constituents will be used to predict expression so we must be able to charaterize them in a way that correlates with the way they are performed.
-Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.
+These constituents will be used to predict expression, so we must be able to characterize them in a way that correlates with the way they are performed.
+Analogous to YQX we are looking for the \textit{context} of the constituent, but in contrast to YQX we also need some description of the constituent itself.
 
@@
 \item[\textbf{Mean pitch interval}] 
-The difference between the mean pitch of the current constituent and the mean pitch of the next constituent, zero if the current is the last constituent
+The difference between the average pitch of the current constituent and the average pitch of the next constituent, zero if the current is the last constituent
 \item[\textbf{Mean duration ratio}]
-The logarithmic ratio between the mean note duration of the current constituent and the mean note duration of the next constituent
+The logarithmic ratio between the average note duration of the current constituent and the average note duration of the next constituent
 \end{description}
 
-Since I-R arch is related to note-level expression it does not generalize well to a constituent level feature.
+Since I-R arch is related to note-level expression, it does not generalize well to a constituent level feature.
 
-The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in mean pitch and mean duration.
+The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in mean pitch and mean duration to the next constituent.
 At note level there is not much to say about the current note besides the pitch and duration.
@@
 \begin{description}
-\item[\textbf{Mean delta pitch}] The mean of all absolute pitch intervals within one constituent. 
-\item[\textbf{Mean delta duration}] The mean of all absolute differences in duration of succeeding notes within one constituent
+\item[\textbf{Mean delta pitch}] The average of all absolute pitch intervals within one constituent. 
+\item[\textbf{Mean delta duration}] The average of all absolute differences in duration of succeeding notes within one constituent
 \end{description}
@@
 Note the difference between these features and the context features.
-The context features use the mean pitch of one constituent and the the mean duration within constituent and compare these to the mean pitch and mean duration of the next constituent.
-The constituent features use the average of all pitch intervals within one constituent and the average of all differences in duration within one constituent, giving a measure of the spread of pitch and differentation of rhythm within the constituent.
+The context features use the mean pitch of one constituent and the mean duration within constituent and compare these to the mean pitch and mean duration of the next constituent.
+The constituent features use the average of all pitch intervals within one constituent and the average of all differences in duration within one constituent, giving a measure of the spread of pitch and differentiation of rhythm within the constituent.
 
 The complete set of score features consists of the two context features and the two constituent features.
+
+The system does not make use of expressive markings in the score.
+A mature performance rendering system should incorporate these in some way.
+Even a human performer is not expected to read the mind of the composer and uses the expressive markings to guide his performance.
+Disregarding articulation marks and dynamic markings has a significant impact on the results: the system has an overall slight preference for playing staccato and softly.
+This is due to the fact that staccato notes or notes that should be played soft are treated as normal notes and bias the statistics.
 
@@
 Some concepts that we think fall under structure level expression are \textit{crescendo} or \textit{decrescendo}, \textit{ritardando} and \textit{piano} or \textit{forte} etc.
-Concepts like crescendo, decrescendo and ritardando arguably fall under note level expression and for simplicity we will not consider them structure level expression.
+From these, concepts like crescendo, decrescendo and ritardando arguably fall under note level expression and for simplicity we will not consider them structure level expression.
 In fact, we will only look at the mean tempo, the mean articulation and the mean loudness of a constituent.
@@
 
-We are going to define our own expression parameters in a similar way but let us first look at some issues with the definitions YQX uses.
-The definition of articulations seems to be awkward and inconsistend.
+We are going to define our own expression parameters in a similar way, but let us first look at some issues with the definitions YQX uses.
+The definition of articulations seems to be a bit awkward and inconsistent.
 Awkward because if a score note is not followed by a rest, the notated silence after it is zero, rendering articulation undefined.
-Inconsisted because all the other features use logarithmic ratios and we see no reason not to use a logarithmic ratio for articulation as well.
+Inconsistent because all the other expression parameters use logarithmic ratios and we see no reason not to use a logarithmic ratio for articulation as well.
 
@@
 Structure level expression may help to define the concepts of local tempo and local loudness as we can simply take the mean tempo and loudness within one constituent and take this to be the local tempo and loudness.
+See section .
 
@@
 Admittedly, we still use mean loudness and tempo to calculate tempo and loudness ratio.
-This is a symptom of using a segmentation instead of hierarchical structure.
-When using hierarchical structure, these parameters could be defined relative to the parent constituent.
+This is a limitation inherent to using a segmentation instead of hierarchical structure.
+When using hierarchical structure, the parameters could be defined relative to the parent constituent.
 
@@
 
-We can now reduce a piece of music, represented as a sequence of notes $N_{ij}$, of which we have a score and a performance, to a sequence of \textit{score feature vectors} $F$ and \textit{expression parameter vectors} $E$.
+We can now reduce a piece of music, represented as a sequence of notes $M_{i,j}$, of which we have a score and a performance, to a sequence of \textit{score feature vectors} $F$ and \textit{expression parameter vectors} $E$.
 First, we segment the score into constituents:
 
-\[\texttt{segment}(N_{ij}) = \{c_1, c_2, \cdots, c_n\}\]
+\[\mbox{segment}(M_{ij}) = \{c_1, c_2, \cdots, c_n\}\]
 We can then extract feature and parameter values from the score and the performance:
@@
 \end{align*}
-where $f_i$ is a feature vector, $p_i$ is the the mean pitch interval, $d_i$ is the mean duration ratio, $\Delta p_i$ is the mean delta pitch and $\Delta d_i$ is the mean delta duration.
+where $f_i$ is a feature vector, $p_i$ is the mean pitch interval, $d_i$ is the mean duration ratio, $\Delta p_i$ is the mean delta pitch and $\Delta d_i$ is the mean delta duration.
 And the expression parameters: 
@@
 Given such a corpus, the rendering of a performance can be formulated as maximizing the probability: $P(E|F)$.
-We can estimate an approximation of this probability by calculating feature likelyhoods and transition probabilities.
-We approximate the features likely-hood, which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$, with the following probability:
+We can estimate an approximation of this probability by calculating feature likelihoods and transition probabilities.
+We approximate the features likelihood, which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$, with the following probability:
 
@@
 
-Where $c(x)$ is the number of occurences of $x$ in the corpus.
+Where $c(x)$ is the number of occurrences of $x$ in the corpus.
 The expression transition probability is approximated by the unigram count of $e_i$ divided by the bigram count of $e_{i-1}, e_i$.
@@
 \end{equation}
-where we make the simplification that expression in one constituent is only dependend on the expression in the previous constituent which is of course not true.
+where we make the simplification that expression in one constituent is only dependent on the expression in the previous constituent which is of course not true.
 
-The problem of creating a performance of a score is now reduced to finding a suitable sequence of expression vectors given a sequence of feature vectors.
-We can calculate the probability of one expression vector of a performance as the product the probabilities defined in equations \ref{emission} and \ref{transition}.
+The problem of creating a performance of a score is now reduced to finding a suitable sequence of expression vectors, given a sequence of feature vectors.
+We can calculate the probability of one expression vector of a performance as the product of the probabilities defined in equations \ref{emission} and \ref{transition}.
 The probability of the entire performance is approximated by the product of the individual expression vector probabilities.
 
-\begin{equation}
+\[
 P(E|F) = \displaystyle\prod_{i=1}^{n}P(f_i|e_i)P(e_i|e_{i-1})
-\end{equation}
+\]
 
@@
 
-\[E^* = \texttt{argmax}_E P(E|F)\]
+\[E^* = \mbox{argmax}_E P(E|F)\]
 
-which can be found using Viterbi decoding.
+which can be found using Viterbi decoding \cite{forney1973viterbi}.
 
@@
 \label{sec:corpus}
-We were lucky to find and receive permission to use the CrestMusePEDB \cite{hashida2008new}, which is a dataset that contains expressive performances of Western classical music by famous pianists.
-The music includes works by Bach, Beethoven and Chopin.
+We were lucky to find and receive permission to use the CrestMusePEDB \cite{hashida2008new}, which is a database containing expressive performances of Western classical music by famous pianists, including Glenn Gould, Vladimir Horowitz and Maurizio Pollini.
+The music in this database includes works by Bach, Beethoven and Chopin.
 
@@
 
-To prepare a score and performance from the corpus for use in our system we use the score to extract melody notes.
+To prepare a score and performance from the database for use in our system we use the score to extract melody notes.
 We do this simply by taking notes in the highest voice in the top staff.
@@
 
-The deltafunctions that are used in the segmentation process use pitch intervals and duration ratios.
+The delta functions that are used in the segmentation process use pitch intervals and duration ratios.
 The pitch intervals use MIDI note numbers which range from 21 to 108 and go up one semi-tone with each step.
 Durations are in milliseconds, calculated from the score and a standard tempo, set at 120 beats per minute.
+
+We constructed two corpora from the database, one consisting of all performances of Chopin and one consisting of all performances of Mozart in the CresMusePEDB.
 
@@
 
-We chose to make the model discrete. We discretize expression and scorefeatures in a different way.
+We chose to make the model discrete. We discretize expression and score features in a different way.
 
@@
 The expression parameters are all logarithmic ratios so they theoretically vary from $-\infty$ to $\infty$.
-Therefore we first normalize the parameters by dividing through the mizimum absolute value found in the corpus.
+Therefore we first normalize the parameters by dividing it by the minimum absolute value found in the corpus.
 Discretization of a normalized expression parameter $p$ into $d$ bins is now done as follows:
 
-\begin{equation}
-D(p) = \texttt{floor}\left(\frac{d}{1+e^{-sp}}\right)
-\end{equation}
+\[
+D(p) = \mbox{floor}\left(\frac{d}{1+e^{-sp}}\right)
+\]
 
@@
 
-\begin{equation}
+\[
 D^{-1}(p) = s^{-1} d^{-1}(-\log(p^{-1}) - 1)
-\end{equation}
+\]
 
@@
 
-\begin{equation}
-D(f) = \texttt{floor}(f*d)
-\end{equation}
+\[
+D(f) = \mbox{floor}(f*d)
+\]
 
@@
 We do not want the conditional probabilities from equation \ref{emission} to become zero so we have to smooth these probabilities.
-We use a smoothing technique known as simple Good-Turing smoothing.
+We use a smoothing technique known as simple Good-Turing smoothing \cite{gale1995good}.
 The idea is that we use the probabilities of things we have seen once for the things we have never seen.
 Recall how we calculate the conditional probability of a feature vector:
+
 \[
@@
 
-Let us call the $c(f_i, e_i)$ the coincedence count.
+Let us call the $c(f_i, e_i)$ the coincidence count.
 Let $N_c$ be the number of things with frequency $c$ in the corpus.
-For example if there are five coincedences $(f_x, e_x)$ nd no other pair of $f$ and $e$ occurs five times, then $N_{5} = 20$.
-Good-Turing re\"estimates counts according to this formula:
+For example if there are five coincidences $(f_x, e_x)$ nd no other pair of $f$ and $e$ occurs five times, then $N_{5} = 20$.
+Good-Turing re-estimates counts according to this formula:
 
@@
 In our case, the probability we seek is $P(f_i|e_i)$, the sample size $N$ is therefore the number of times we have seen $e_i$: $c(e_i)$.
-$N_c$ is the number of coincedences with $e_i$ that we have seen $c$ times.
+$N_c$ is the number of coincidences with $e_i$ that we have seen $c$ times.
 The count $c$ is $c(f_i, e_i)$ and $c^*$ is derived using equation \ref{count}.
@@
 \[
-P(f_{\texttt{unseen}}|e_i) = \frac{1}{U} \frac{N_1}{c(e_i)} 
+P(f_{\mbox{unseen}}|e_i) = \frac{1}{U} \frac{N_1}{c(e_i)} 
 \]
-Where U is the number of unseen coincedences, which is determined by the number of different observations in the corpus plus the new observations from the score minus the number of coincedences with $e_i$.
+Where U is the number of unseen coincidences, which is determined by the number of different observations in the corpus plus the new observations from the score minus the number of coincidences with $e_i$.
 
-In order to make this work we cannot use $N_c$ directly since it will not be defined for every $c$. We use a least square apprixmation using the following function.
+In order to make this work we cannot use $N_c$ directly since it will not be defined for every $c$. We use a least square approximation to the following function.
 
@@
 This completes the definition of simple Good-Turing smoothing.
-However, even now there may still be some expression parameter vectors that occur with unique feature vectors, so for that expression only $N_1$ is defined.
-We cannot fit a function to one sample, so if we have only one $Nc$ sample, we simply turn off Good-Turing smoothing and accept the fact that some probabilities will be zero.
+However, even now there may still be some expression parameter vectors that occur with only unique feature vectors, so that for that expression only $N_1$ is defined.
+We cannot fit a function to one sample, so if we have only one $N_c$ sample, we simply turn off Good-Turing smoothing and accept the fact that some probabilities will be zero.
+
+In practice we have to use quite a large bin size.
+Despite this large bin size, many combinations of features and expression parameters were never encountered during training.
+The same applies to transitions of expression: only small portion of all possible transitions of expression are actually encountered in the corpus.
+Some of the feature likelihoods can be smoothed like we described above, but the transition probabilities are not smoothed at all.
+Since our model does not allow zero probability transitions (this would make the likelihood of the entire performance zero), our model is heavily overfitting.
+The resulting performances only contain changes of expression that were actually encountered in the corpus.
+
+In most applications of machine learning, this degree of overfitting would not be acceptable.
+However, in this particular application we seem to be able to get away with it.
+Playing phrase transitions exactly as encountered in the corpus seems to be acceptable as long as we play the right transitions at the right moment.
+The overfitting and the large bins size gives the performances a bit of a stereotypical and cartoonistic feel.
+A larger corpus allows would reduce the overfitting and the requirement of large bins, but this kind of corpus is not available to us at the moment.
 
@@
 After rendering an expressive performance we can simply put them back in and estimate their expressive parameters.
-We do this by this by by giving each  bass and harmony note the expressive parameters of the last played melody note.
+We do this by giving each bass and harmony note the expressive parameters of the last played melody note.
+
+Unfortunately the notes we treat as melody notes are not always the notes that humans perceive as melody notes.
+Sometimes the melody is voices other than the top voice.
+In other cases, like Bach's fugues, there are multiple melodies in multiple voices at the same time.
+In addition to that, bass and harmony notes should generally not be played with th same expression as melody notes.
+Bass notes, for example, tend to be played slightly softer and more legato than melody notes.
 
@@
 %\item Extract melody
-%\item Extract scorefeatures from melody
+%\item Extract score features from melody
 %\item lookup notes in deviation, ignore missing notes
@@
 
-We compiled two corpora from our the CrestMusePEDB.
+We compiled two corpora from the CrestMusePEDB.
 One contained 42 performances of 13 piano works by Mozart, the other contained 49 performances of 19 works by Chopin.
-For more details on each corpus see appendix \ref{appendix:corpus}.
 We suspect that performances of Mozart's music most outspokenly use structural level expression.
 Expression within a constituent tends to be more unsteady in performances of Chopin's music.
+In musical terms, more expression like riterdando, crescendo and rubato is used.
 
@@
 To see what the effect was of each corpus on different styles of music we performed each work on both corpora.
-For reference, we also show a performance by one of the pianists in the corpus.
-Given the small dataset and the free nature of expression we did not expect the performances to correlate much with performances by human pianists.
+For reference, we also show a performance by a real human pianist that was used during training.
+Given the small dataset and the free nature of expression we did not expect the performances to correlate much with performances by human pianists and as can be seen in the results they indeed do not.
 
@@
 In practice, this means that the constituent features (see section \ref{sec:scorefeatures}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well.
-The expression features where discretized into ten bins, with a sensitivity of five.
-The bins had to be this large to recognise most of the score features in the score that was to be performed.
+The expression features were discretized into ten bins, with a sensitivity of five.
+The bins had to be this large to recognize most of the score features in the score that was to be performed.
 
-Segmentation was done using only delta onset trees. This proved to be most reliable for Mozart.
+Segmentation was done using only delta onset trees. This proved to be most reliable for Mozart and Chopin.
 
-When rendering a Mozart performance on the Mozart corpus and a Chopin performance on the Chopin corpus any of the performances of the piece that had to be performed were left out of the corpus during training.
+When rendering a Mozart performance on the Mozart corpus and a Chopin performance on the Chopin corpus all performances of the piece that had to be performed were left out of the corpus during training.
 
-Plots showing dynamics, tempo and articulation of the resulting performances are shown in figure \ref{fig:mozartperformances} and \ref{fig:chopinperformances}.
-The values on the y-axis, labeled deviation, are the exponents of the expression parameters, which themselfs are logarithmic ratios.
+Plots showing dynamics, tempo and articulation of the resulting performances are shown in figure \ref{fig:mozartperformances} and \ref{fig:chopinperformances}.\footnote{Of course, staring at plots of music performances is not nearly as informative as actually listening to them.
+For this purpose, a few performances including the ones shown here can be found online at \texttt{http://soundcloud.com/expressiveperformance/sets/final-results/}.} The values on the y-axis, labeled deviation, are the exponents of the expression parameters, which themselves are logarithmic ratios.
 In other words, the values indicated by the axes are ratios.
@@
 
-Of course staring at plots of music performances is not nearly as interesting as actually listening to them.
-For this purpose, a few demo performances including the ones shown here can be found online at [DEMO URL].
-
 %The system was tested on works from Mozart as we do not expect our system to generalize well to other composers.  We used
@@
 \centering
+
 \subfigure[Performance by the system trained on the Mozart corpus]{
@@
 }
+
 \subfigure[Performance by the system trained on the Chopin corpus]{
@@
 }
+
 \subfigure[Performance by Maria Jo\~ao Pires]{
@@
 
-\section{Discussion}
+\section{Discussion and Evaluation}
 \label{sec:discussion}
 
-Despite the large bin size, many combinations of features and expression parameters were never encountered during training.
-The same applies to transitions of expression, only small portion of all possible transitions of expression are actually encountered in the corpus.
-Some of the feature likelyhoods are smoothed but the transition probabilities are not smoothed at all.
-Since our model does not allow zero probability transitions, our model is heavily overfitting.
-The resulting performances only contain changes of expression that were actually encountered in the corpus.
+%. Whether it makes musical sense to perform Chopin on a system trained on Mozart is left for the reader to decide. 
+In the introduction, we mentioned the problem of evaluation.
+The most effective evaluation that is available at the moment is participation in the performance rendering contest, Rencon \cite{hiraga2002rencon}.
+Unfortunately, participating in this contest was not feasible for this project.
+Therefore we are left with our own subjective observations and qualifications.
 
-In most applications of machine learning, this degree of overfitting would not be acceptable.
-However, in this particular application of machine learning we seem to be able to get away with it.
-Playing a phrase transitions exactly as encountered in the corpus seems to be acceptable as long as we play the right transitions at the right moment.
-The overfitting and the large bins size give the performances a bit of a stereotypical and cartoonistic feel, but that is mostly caused by the small corpus.
-Whether it makes musical sense to perform Chopin on a system trained on Mozart is left for the reader to decide.
+% Phrase transitions
+% Qualitative judgements
+% Comparison
+% Segmentation
 
-Although the output of the system is polyphonous, the system at its core only uses a single voice for its performances.
-We use the highest notes in the top staff for this and hope that these are the melody notes.
-Apart from the fact that this is not always the case, the other notes generally should note be played with the same expression.
-Bass notes for example are often played slightly softer and more legato than melody notes.
+When listening to the performances generated by our system, the effect of per-segment expression can clearly be heard.
+Making changes in dynamics at phrase transitions effectively increases the human feel of a performance.
+These dynamics changes work well when phrase transitions are correctly detected.
+Sometimes however, when a phrase transition is incorrectly detected, changes in dynamics can be very disrupting.
+This can be heard in the system's rendering of Mozart's Piano Sonata KV331 I.
+when trained on the Chopin corpus.
+At the end of the second bar, a transition is detected one note too early, causing a strong accent to be placed at an uncomfortable point in the piece.
+These incorrectly detected transitions are not systematic and in general the transitions detected by the system are musically plausible, if not, interesting.
 
-The system does not make use of expressive markings in the score.
-A mature performance rendering system should incorporate these in some way.
-Even a human performer is not expected to read the mind of the composer and uses the expressive markings as a guide for performance.
-Disregarding articulation marks and dynamic markings has a significant impact on the results: the system has an overall slight preference for playing staccato and softly.
-This is due to the fact that staccato notes or notes that should be played soft are treated as normal notes and bias the statistics.
+Changes in tempo are riskier than changes in dynamics.
+Small changes in dynamics, when made at phrase transitions at least, do not disrupt a performance as much as small changes in tempo at the wrong moment.
+Often, the tempo deviations the system makes sound good or at least acceptable, but the system does tend to sometimes produce awkward sounding changes in tempo, like a sudden slow down that can be heard a number of times in our results online.
 
-Ignoring note level expression is quite an extreme measure to take when  it has been an adequate way to demonstrate the phenomenon of structure level expression.
+Despite the fact that our system only determines expression per segment, small segments and gradual changes in dynamics or tempo can give a crescendo or accelerando feel to certain parts of the music.
+It was interesting to see that this could be achieved using only structure level expression.
+
+The system clearly accentuates structure, but only on the level of the segmentation.
+Although locally the changes in dynamics, articulation and tempo often seem sensible, at a higher level there does not seem to be an idea behind it all and the overall flow of the performance seems a bit random.
+When theme is repeated a few times, the system may perform it completely different each time.
+While some variation when playing a repetition of a phrase makes a performance interesting, playing it completely different each time conveys the feeling that the performer does not really know what he is doing.
+This sort of randomness comes as no surprise since our model uses only transition probabilities and local expression probabilities.
+To really capture global flow of performance, a grasp of hierarchical structure and repetition seems crucial.
+
+Ignoring note level expression has been an adequate way to demonstrate the phenomenon of structure level expression.
 Yet, we cannot deny the need for note level expression in a performance rendering system that intends to completely simulate human expression.
 Section \ref{sec:futurework} suggests a possible integration of structure and note level performance rendering systems.
-
-%overfitting
-
-%Bach
-
-%Since transition probabilities are not smoothed only transitions that actually occur in the corpus have a nonzero probability
-
-%Unfortunately we do not have acces to the large dataset that YQX uses.
-The dataset we use is smaller.
-Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger.
-We have discretize into rather large bins for this reason, resulting in cartoonistic performances.
-
-%The system does not make use of expressive markings in the score.
-A mature performance rendering system should incorporate these in some way.
-Een a human perform is not expected to read the mind of the composer.
-
-%We think we can afford to do this because:...
 
@@
 
-In this thesis we have introduced a way of recognizing and using structure level expression.
-We argued why we think it is a crucial aspect of performance rendering systems and we critisized  note level performance rendering for not being able to correctly learn when to make daring and large changes in expression as well as lacking the ability to convey structure level expression.
+In this thesis, we have introduced a way of recognizing and using structure level expression.
+We argued why we think it is a crucial aspect of performance rendering systems and we criticized  note level performance rendering for not being able to correctly learn to make daring and large changes in expression as well as lacking the ability to convey structure level expression.
 
-Our goal was to create a system that was the complement of a note level performance rendering system.
-A system that dared to make sudden big changes in expression and clearly accentuates musical structure.
-Our results show that our system indeed is not shy this kind of big expressive changes.
-The performances of the system contain sometimes smooth, sometimes sudden changes of expression.
-It is up to the listener to judge whether these changes are actually an improvement over the smoother performances of note level performance rendering systems.
+Our goal was to create a system that, in contrast to a note level performance rendering system, dares to make sudden big changes in expression and clearly accentuates musical structure.
+Our results show that our system indeed is not shy about this kind of big expressive changes.
+The performances of the system contain sometimes smooth, sometimes sudden changes of expression that really do seem to be an improvement over note level performance rendering systems.% It is up to the listener to judge whether these changes are actually an improvement over the smoother performances of note level performance rendering systems.
 
-The system that we proposed reduces the idea of structure to a segmentation of the score into constituents based on a structural analysis.
-For the structural analysis we used the delta framework \cite{markwin}, proposed by Markwin van den Berg.
+The system that we proposed reduces the idea of structure to a segmentation of a score into constituents based on a structural analysis.
+For the structural analysis, we used the delta framework \cite{markwin}, proposed by Markwin van den Berg.
 We noticed that a segmentation based on onset delta trees works well for some music but does not recognize all constituent transitions.
-To overcome this we suggested to use a second order pitch delta tree, but we ended up discarding this idea because it still produced awkward segmentations.
-Our system would therefore not work very well on music that has a consistend rhythmic structure like some of Bach's music.
-In the end, an onset tree segmentation worked good enough to produce musically relevant segmentations given the the right genre of music.
+To overcome this we suggested to use a second order pitch delta tree, but we ended up discarding this idea because it produced too many incorrect splits in phrases.
+Our system would therefore not work very well on music that has a consistent rhythmic structure like some of Bach's music.
+In the end, an onset tree segmentation worked well enough to produce musically relevant segmentations given the right genre of music.
 
-To characterize constituents and their relations to other constituents we used constituent likely-hoods and bigram transition probabilities of expression.
+To characterize constituents and their relations to other constituents we used constituent likelihoods and bigram transition probabilities of expression.
 The resulting hidden Markov model could be used in combination with Viterbi decoding to generate performances.
@@
 
-The abilities of performance rendering systems still are not even close to the abilities of real human pianists.
-In relation to this system we can clearly identify a number of things that would bring the system a little bit closer to the abilities of a human pianist.
+%Clearly, performance rendering systems do not come close to the expressiveness of real human performances. 
+We have presented a new approach to performance rendering, one that is purely structure based.
+Our first results look promising but also clearly demonstrate some limitations of our system.
+We think that these limitations are easily identifiable and addressed and with some extensions, structure based performance rendering can bring us far.
+We will shortly list some of the limitations that we discussed in our evaluation in section \ref{sec:discussion} and provide some suggestions on how to address them.
 
-The segmentation is not always perfect.
-The delta framework offers us more analytic power than we have used in this system but we were not able to sensibly combine its analyses into one segmentation apporach.
-We are unsure how we should combine different delta trees into some sort of consensus.
-
-Such a consensus tree would be especially useful in an extension of our system to a true structure based performance rendering system, namely one that utilizes multilevel structure.
+The segmentation of a score is not always perfect and leaves room for improvement.
+Ultimately, we do not want discard the idea of a segmentation completely and move to a multilevel structure analysis, this will help to eliminate the randomness that we discussed in our evaluation.
+Whether the delta framework can be used for such an analysis remains a topic for future research.
+The delta framework offers us more analytic power than we have used in this system.
+But in order to use it to generate multilevel structure, a way to combine different delta trees into one consensus tree is needed.
 
@@
 \cite{kim2010performance} split a piece into melody, bass and harmony and use separate probabilistic models for each of them.
-When using different expression for different voices, another problem presents itself, namely how to synchronize the voices so that the performance still sounds like it is played by one artist or a group of artists with an adequate notion of timing.
-Pop-e \cite{hashida2006pop} uses an automatic analysis to determine the attentive part: the voice of the music perceived as melody at any point in the piece.
+When using different expression for different voices, another problem presents itself, namely how to synchronize the voices so that the performance still sounds like it is played by one performer or a group of performers with an adequate mastery of timing.
+Pop-e \cite{hashida2006pop} uses an automatic analysis to determine the attentive part in music: the voice of the music perceived as melody at any point in the piece.
 It uses this information to synchronize expression in different voices.
 
-As we mentioned in section \ref{sec:discussion}, our system ignores expressive markings.
+As we mentioned in our evaluation, our system ignores expressive markings.
 The result is that depending on the corpus, the performance renderer has unwanted biases.
-A possible approach to incorporating expressive markings in the system is to determine what, given an expressive mark in the, is the average way to play a given note and normalize this out of the corpus.
-Say staccato notes are played, on average, half the length of the notated length.
-We specify articulation deviation relative to this half duration instead of the full duration.
+A possible approach to incorporating expressive markings in the system is to determine what, given an expressive mark in the score, is the average way to play a given note and normalize this out of the corpus.
+Say staccato notes are played, on average, half length notated in the score, we specify articulation deviation relative to this half duration instead of the full duration.
+That way the system now recognizes the note as a normal staccato note instead of an extremely short legato note.
 
@@
 Good-Turing smoothing assigns equal probabilities to all unseen score features.
-It is likely that unseen scorefeatures that are very similar to scorefeatures that we have seen should also be played in a similar way as the scorefeatures they resemble.
+It is likely that unseen score features that are very similar to score features that we have seen should also be played in a similar way as the score features they resemble.
 A different smoothing technique may be able to use this information.
-We could use a smoothing technique similar to Katz-Backoff REF: upon encountering an unseen feature vector, discretize again the corpus into increasingly larger bins until we do find the featurevector in our corpus.
+We could use a smoothing technique similar to Katz-Backoff \cite{katz1987estimation}: Upon encountering an unseen feature vector, discretize again the corpus into increasingly larger bins until we do find the feature vector in our corpus.
+
+% where the frequency of an N-gram with zero occurrences is approximated by backing off to the count of the (N-1)-grams, (N-2)-grams, etc. until the count is not zero anymore.
 
@@
 It is probably telling when a phrase is repeated three times and then slightly altered the fourth time.
-Although finding similarity and musically significant repetition is a subject of its own the delta framework could help to define repetition arbitrarily of transposition or rhythm.
-A list of pitch deltas can for example be used to detect repetition independend of transposition and a list of duration deltas can be used to detect repetition of rhythm independend of the notes used.
+Although finding similarity and musically significant repetition is a subject of its own, the delta framework could help to define repetition arbitrarily of transposition or rhythm.
+A list of pitch deltas can for example be used to detect repetition independent of transposition and a list of duration deltas can be used to detect repetition of rhythm independent of the notes used.
 
@@
 Timing alterations of notes are always defined relative to the base tempo.
-This does not correspond to the intuition that the \textit{the tempo itself} is altered during the performance and that rhythmic changes should be seen relative to the local tempo.
+This does not correspond to the intuition that the \textit{the tempo itself} is altered during the performance, resulting in a local tempo, and that rhythmic changes should be seen relative to this local tempo.
 The same applies to the way YQX looks at loudness.
@@
 We can define expressive tempo relative and dynamics relative to the expression parameters of the constituent.
-A possible approach is to first define constituent level expression and use a note level expression system on each constituent seperately.
+A possible approach is to first define constituent level expression and use a note level expression system on each constituent separately.
 
-Another approach is to use hierarchical structure.
-Expression of every constituent can be defined relative to the parent constituent.
-If we let the hierarchy extend to the note level (the delta framework does this) we get a smooth integration of note level performance rendering and structure level performance rendering.
-In such a system notes are terminal notes in a structure tree, note level expression is specified relative to the parent constituent's expression which is again relative to the its own parent constituent.
-Ultimately the root node's expression contains the base expressive parameters.
-
-Finally, as we mentioned in section \ref{sec:approach} the ability of the system to characterize constituents and their relations to other constituents in a musically relevant way is an important factor influencing the results.
-We do not have the illusion that a set of four scorefeatures and bigram transition probabilities are a valid model of human musical intuition, let alone producing highly expressive performances.
+As we mentioned in section \ref{sec:approach} the ability of the system to characterize constituents and their relations to other constituents in a musically relevant way is an important factor influencing the results.
+We do not have the illusion that a set of four score features and bigram transition probabilities are a valid model of human musical intuition, let alone producing highly expressive performances.
 A better notion of repetition, a notion of key, harmony and general musical knowledge are the least that seems to be required for that purpose.
 
-\section{Acknowledgements}
+Finally we would like to make a small comment about the possible value of performance rendering in an artistic sense.
+Computer generated performance, and in a broader view, computer generated art, is often criticized for merely copying a very specific human behavior and not really contributing something new.
+We think a very small hint at the way computer generated performances may one day be able to provide a really new perspective on music performance, is given by our system: when listening to the performances of the systems we noticed sometimes the system made surprising phrase transitions or suggested a phrase transition at a surprising point in the music.
+Surprising, because no pianist would probably think about making these expressive decisions, yet they did not sound bad, making the performance truly unique.
+This closing note gives us hope that the field of computer generated performances not only copies human performances but humans may one day learn from computer generated performances.
 
+%\section{Acknowledgements}
+
+%I would like to thank Remco Scha for providing guidance and support. I am very grateful to the CrestMusePEDB project for kindly providing me access to their database that they have p
+\pagebreak
  \bibliographystyle{plain} % plain, nature, jbact
@@
 
-\appendix
-\section{Corpus}
-\label{appendix:corpus}
-\begin{tabular}{ll}
-\textbf{Work} & \textbf{Performer}\\
-\hline
-Sonata KV331 \MakeUppercase{\romannumeral 1}. & Hiroko Nakamura\\
-& Glenn Gould\\
-& Christoph Eschenbach\\
-& Ingrid Haebler\\
-& Lili Kraus\\
-& Maria Jo\~ao Pires\\
-& Alicia De Larrocha\\
-& kn?\\
-& yi?\\
-& tn?\\
-& Norio Shimizu\\
-& mo\\
-& mk?\\
-& ea?\\
-& nm?\\
-& kt?\\
-& tm?\\
-\hline
-Sonata KV331 \MakeUppercase{\romannumeral 2}. & Hiroko Nakamura\\
-& Maria Jo\~ao Pires\\
-& Alicia De Larrocha\\
-& ea?\\
-\hline
-Sonata KV331 \MakeUppercase{\romannumeral 3}. & Hiroko Nakamura\\
-& Maria Jo\~ao Pires\\
-& Christoph Eschenbach\\
-\hline
-Sonata KV545 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
-& Glenn Gould\\
-\hline
-Sonata KV545 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
-& Glenn Gould\\
-\hline
-Sonata KV545 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
-& Glenn Gould\\
-\hline
-Sonata KV279 \MakeUppercase{\romannumeral 1}. & Glenn Gould\\
-& Maria Jo\~ao Pires\\
-\hline
-Sonata KV279 \MakeUppercase{\romannumeral 2}. & Glenn Gould\\
-& Maria Jo\~ao Pires\\
-\hline
-Sonata KV279 \MakeUppercase{\romannumeral 3}. & Glenn Gould\\
-& Maria Jo\~ao Pires\\
-\hline
-Sonata KV310 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
-& Glenn Gould\\
-& Hiroko Nakamura\\
-Sonata KV310 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
-Sonata KV310 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
-Sonata KV570 \MakeUppercase{\romannumeral 3}. & nm?\\
-\end{tabular}
+%\appendix
+%\section{Corpus}
+%\label{appendix:corpus}
+%\begin{tabular}{ll}
+%\textbf{Work} & \textbf{Performer}\\
+%\hline
+%Sonata KV331 \MakeUppercase{\romannumeral 1}. & Hiroko Nakamura\\
+%& Glenn Gould\\
+%& Christoph Eschenbach\\
+%& Ingrid Haebler\\
+%& Lili Kraus\\
+%& Maria Jo\~ao Pires\\
+%& Alicia De Larrocha\\
+%& kn?\\
+%& yi?\\
+%& tn?\\
+%& Norio Shimizu\\
+%& mo\\
+%& mk?\\
+%& ea?\\
+%& nm?\\
+%& kt?\\
+%& tm?\\
+%\hline
+%Sonata KV331 \MakeUppercase{\romannumeral 2}. & Hiroko Nakamura\\
+%& Maria Jo\~ao Pires\\
+%& Alicia De Larrocha\\
+%& ea?\\
+%\hline
+%Sonata KV331 \MakeUppercase{\romannumeral 3}. & Hiroko Nakamura\\
+%& Maria Jo\~ao Pires\\
+%& Christoph Eschenbach\\
+%\hline
+%Sonata KV545 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
+%& Glenn Gould\\
+%\hline
+%Sonata KV545 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
+%& Glenn Gould\\
+%\hline
+%Sonata KV545 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
+%& Glenn Gould\\
+%\hline
+%Sonata KV279 \MakeUppercase{\romannumeral 1}. & Glenn Gould\\
+%& Maria Jo\~ao Pires\\
+%\hline
+%Sonata KV279 \MakeUppercase{\romannumeral 2}. & Glenn Gould\\
+%& Maria Jo\~ao Pires\\
+%\hline
+%Sonata KV279 \MakeUppercase{\romannumeral 3}. & Glenn Gould\\
+%& Maria Jo\~ao Pires\\
+%\hline
+%Sonata KV310 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
+%& Glenn Gould\\
+%& Hiroko Nakamura\\
+%Sonata KV310 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
+%Sonata KV310 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
+%Sonata KV570 \MakeUppercase{\romannumeral 3}. & nm?\\
+%\end{tabular}
 

### 2011-08-10 Code/alignment.py

@@ -51,10 +51,10 @@ class Alignment:
   def expressiveMelody(self):
     return self.performance(self.melody())
 
-  def melodyPerformance(self):
-    return self.performance(self.score, self.melody())
+  def melodyPerformance(self, articulations=None):
+    return self.performance(self.score, self.melody(), articulations)
 
-  def performance(self, score=None, melody=None):
+  def performance(self, score=None, melody=None, articulations=None):
     print "Creating performance"
     if not score:
       score = self.score
@@ -94,6 +94,7 @@ class Alignment:
               # If a melody is defined we need to know if the current note is a melody note
               # If it is not, we need to know if it is played at the same time as a melody note
               # If this isn't the case, use the tempo curve for expression and some measure of dynamics(nearest melody note?)
+              articulation = None
               if melody:
                 melodynote = None
                 # 
@@ -132,6 +133,8 @@ class Alignment:
                     print '{0} {1}'.format(measure.number, melodynote.offset)
                     deviation = None
                   deviation = self.alignment[melodynote.id, str(melodynote.pitch)]
+                  if articulations:
+                    articulation = articulations[melodynote.id, str(melodynote.pitch)]
                   if not deviation: continue
                   # Attack and release should be zero in this case
                   deviation = (0, 0, deviation[2], deviation[3])
@@ -139,6 +142,8 @@ class Alignment:
                   if not (melodynote.id, str(melodynote.pitch)) in self.alignment:
                     print '{0} {1}'.format(measure.number, melodynote.offset)
                   deviation = self.alignment[melodynote.id, str(melodynote.pitch)]
+                  if articulations:
+                    articulation = articulations[melodynote.id, str(melodynote.pitch)]
                 lastmelodynote = melodynote
               else:
                 if not (note.id, str(n.pitch)) in self.alignment:
@@ -159,12 +164,19 @@ class Alignment:
               onvel = deviation[2] * self.deviations.basedynamics
               offvel = 0
               on_ms = performancetime + self.deviations.getExpressiveTime(measure.number, note.offset)
-              off_ms = on_ms + self.deviations.getExpressiveDuration(measure.number, note.offset, n.duration.quarterLength)
+              if articulation:
+                duration = self.deviations.getExpressiveDuration(measure.number, note.offset, n.duration.quarterLength)
+                articulated_duration = articulation * float(duration)
+                print '{0} * {1} = {2}'.format(duration, articulation, articulated_duration)
[…]

### 2011-08-10 Code/perform.py

@@ -28,6 +28,7 @@ def vanDerWeijPerformSimple(score, melody, segments, expression_vectors, average
   deviations.basedynamics = average_loudness
   tempo_deviations = {}
   alignment = {}
+  articulations = {}
 
  
   for segment, expression in zip(segments, expression_vectors):
@@ -74,12 +75,13 @@ def vanDerWeijPerformSimple(score, melody, segments, expression_vectors, average
       articulated_duration = articulation * float(duration)
       release = (articulated_duration - duration) / converter.microseconds_to_ticks(deviations.getBeatDuration(measure, int(scorenote.offset)))
       alignment[scorenote.id, str(scorenote.pitch)] = [0, release, average_relative_loudness, 0]
+      articulations[scorenote.id, str(scorenote.pitch)] = articulation
   
   a = Alignment(score, deviations, noAlign=True)
   a.alignment = alignment
   a.m = melody
 #  toFile(a.m, 'score.txt') 
-  return a.melodyPerformance()
+  return a.melodyPerformance(articulations=articulations)
 
 
 def perform(score, melodyscore, onset, expression, dynamics=None, tempo=None, converter=NoteList()):

### 2011-08-10 Code/performancerenderer.py

@@ -351,6 +351,9 @@ def test(f_discretization=10, e_discretization=30, indep=False, selection=None, 
   elif segmentation == 'new':
     print 'Using new segmentation'
     onset = structure.newSegmentation(melody)
+  elif segmentation == 'notelevel':
+    print 'Using notelevel segmentation'
+    onset = structure.noteLevel(melody)
 
   #onset = structure.groupings(structure.list_to_tree(structure.first_order_tree(structure.onset, melody, 0.1)), 1)
   #namelist = []

### 2011-08-10 Code/structure.py

@@ -313,6 +313,9 @@ def adaptiveSegmentation(tree):
 def newSegmentation(notes):
   return adaptiveSegmentation(list_to_tree(first_order_tree(onset, notes)))
 
+def noteLevel(notes):
+  return [[n] for n in notes]
+
 
 if __name__ == '__main__':
   import sys

### 2011-08-10 Code/train.py

@@ -33,7 +33,8 @@ def train(trainset):
     deviations = db.getDeviation1(query)
     alignment = Alignment(score, deviations)
     melody = alignment.melody()
-    segments = structure.newSegmentation(tools.parseScore(melody))
+    #segments = structure.newSegmentation(tools.parseScore(melody))
+    segments = structure.noteLevel(tools.parseScore(melody))
     const += len(segments)
     lengths = sum([len(s) for s in segments])
     m = max([len(s) for s in segments])
